Changes in wing morphology rather than wingbeat kinematics enabled evolutionary miniaturization of hoverflies

Version 1

Abstract

Abstract

Due to physical scaling laws, size greatly affects animal locomotor ability and performance. Whether morphological and kinematic traits always jointly respond to size variation is however poorly known. Here, we examine the relative importance of morphological and kinematic changes in mitigating the consequence of size on aerodynamic force production for weight support in flying insects, focusing on hovering flight of hoverflies (Syrphidae). We compared the flight biomechanics, aerodynamics, and morphology of eight hoverfly species varying from 5 to 100 mg. Our study reveals no effect of body size on wingbeat kinematics among species, suggesting that morphological rather than kinematic changes may compensate for the reduction in weight support associated with an isometric reduction in size. Computational fluid dynamics simulations confirmed that variations in wing morphology, and not kinematics, allow species of different sizes to generate weight support. We specifically show that relatively larger wings and aerodynamically more effective wing shape have evolved in smaller hoverflies, mitigating the reduction in aerodynamic weight support with decreasing size. Altogether, these results suggest that hovering flight of hoverflies underpins highly specialised wingbeat kinematics, which have been conserved throughout evolution; instead, wing morphological adaptations have enabled the evolutionary miniaturisation of hoverflies.

Introduction

Evolutionary changes in size are common over the course of animal diversification (Clauset and Erwin, 2008; Hanken and Wake, 1993), strongly impacting many aspects of morphology, physiology, and behaviour. Investigating how changes in body size affect other traits (i.e., scaling) is thus essential to understand the causes of traits variation in animals. Traditionally rooted in the field of morphology and physiology (Gould, 1966; Hill, 1950; Pélabon et al., 2014), allometric studies have broadened to explore how more complex traits scale with animal size. These include locomotor kinematics (e.g. flight, stride or swimming kinematics, Bejan and Marden, 2006; Riskin et al., 2010), and behaviour (e.g. mating or escape tactics, Dial et al., 2008). Although behavioural or biomechanical traits are often more noisy and challenging to measure, their inclusion in allometric studies is crucial to better understand trait evolution (Cloyed et al., 2021). Indeed, most phenotypes involve both morphological and behavioural traits, each responding in potentially different ways to size variation (Clemente and Dick, 2023). Variation in scaling between body size and other traits is generally most apparent between species, and can be partly explained by neutral divergence during phylogenetic history (Blomberg et al., 2003). Understanding the cause of trait variation among species thus requires controlling for the influence of phylogeny on allometric relationships.

In this study, we address the consequences of size variation between species (i.e. evolutionary allometry), and specifically question the relative importance of wing morphology versus wingbeat kinematic changes in mitigating the effect of size on flight ability.

Flight typically combines a suite of morphological and biomechanical traits, providing a relevant context to jointly examine how morphology and locomotor kinematics change with size. The physical laws governing flight allow drawing predictions on the scaling of morphology and kinematics in flying animals.

These physical laws set both upper and lower bounds on the size of flying animals. For large flying animals, the aerodynamic lift for weight support scales linearly with the surface area of the wing and quadratically with the forward speed of the flying animal(Norberg, 2007). Physical scaling laws dictate that, for isometric scaling, wing surface area scales with (body mass2/3) (Schmidt-Nielsen, 1975). Thus, due to constraints for maintaining weight support, larger flying animals exhibit disproportionately larger wings and tend to fly faster (Norberg, 2012). This positive allometry in wing surface area and flight speed with body mass is needed to generate enough lift to stay in the air.

Smaller flying animals such as insects generally fly slower, and thus air movements over the wing are governed by wing flapping, especially in hovering flight. The aerodynamic lift forces produced by flapping wings therefore scale linearly with the second-moment-of-area (S2) of the wing, and quadratically with the angular velocity of the beating wing (C. P. Ellington, 1984). For isometry, the second moment of wing area scales with (body mass)4/3. As this growth factor is larger than one, wing area decreases at higher rate compared to body weight, and therefore the aerodynamic force reduces more rapidly than body weight. Disproportionately larger wings and/or a higher wingbeat frequency is thus expected to evolve in smaller flying insects as a compensatory mechanism. The negative allometric scaling in wingbeat frequency with body mass is supported by the consistent negative correlation between wingbeat frequency and size observed in most flying animals (Norberg, 2007; Rayner, 1988; Riskin et al., 2010; Tercel et al., 2018). A general trend toward disproportionately larger wings in smaller species is however less striking among species (but see Danforth, 1989; C. Ellington, 1984a; Outomuro et al., 2013)), suggesting that morphological changes in response to size variation are not always straightforward.

Although empirical data often align with expectations drawn from physical laws, the scaling of morphology and kinematic with size can sometimes be challenging to predict. Selective pressures associated with ecological specialisation may influence the evolution of morphology and kinematic, and conflict with the constraints imposed by physical scaling laws. Each component may then adjust for the constraints imposed by weight support in varying degree. For example, changes in morphology have been shown to predominantly mitigate the consequences of body size: among hummingbird species of various sizes, changes in wing area – but less in wingbeat kinematics – compensate for weight support (Skandalis et al., 2017). Similarly, wing area but not wingbeat frequency was found to covary with body size among species of bees (Duell et al., 2022; Grula et al., 2021). Such discrepancies with the general scaling expectation may point at a particular selective regime maintaining specialised flight kinematics and ability.

To what extent the evolution of wingbeat kinematics and wing morphology is influenced by physical scaling demands or by ecological specialisation is poorly understood. Here we focused on an ecologically specialised insect group, the hoverflies, to test if the consequences of size variation are mostly mitigated by changes in wingbeat kinematics or in wing morphology.

Hoverflies (Diptera: Syrphidae) are among the most species-rich groups of nectar-foraging insects and exhibit striking variation in size, ranging from 0.5 mg to over 200 mg (Katzourakis et al., 2001). Despite these large size differences, all adult hoverflies show exceptional hovering abilities. This highly specialised flight behaviour may have been promoted because it enhances foraging efficiency, as they hover in front of flowers for nectar feeding. It could also have been favoured because hovering plays an important role in hoverfly mating behaviour: male hoverflies aggregate at specific locations such as under a tree and hover steadily, waiting for passing females to pursue (Gilbert, 1984; Heinrich and Pantle, 1975). Hovering abilities may thus be correlated with mating success (Downes, 1969; Gilbert, 1984). The strong selection acting on the flight of hoverflies may thus interfere with general scaling expectations between body mass, wing morphology and wingbeat kinematics, especially during evolutionary miniaturisation. Tiny hoverflies require relatively large wings or high wingbeat frequencies to maintain weight support during hovering.

Here, we address the outcome of this evolutionary process by examining the scaling relationships between wing morphology and body mass in eight hoverfly species ranging from 5 to 100 mg, and conjointly assessing how wing morphology and wingbeat kinematics scale with body mass. We then combined computational fluid dynamic simulations with quasi-steady aerodynamic modelling to estimate aerodynamic forces produced by the different species in hovering flight. Based on this, we determined how the relative changes in wingbeat kinematics and wing morphology among the eight species contribute to producing weight support during hovering flight. Because flight behaviour is more flexible than morphology (Blomberg et al., 2003), we predict that changes in wingbeat kinematics will prevail over changes in wing morphology. Alternatively, if selection for specialised hovering flight is limiting kinematic flexibility, we expect weight support to be achieved through changes in wing morphology rather than in wingbeat kinematics.

Material and methods

Collecting and identifying hoverflies

We collected hoverflies from the wild, in September 2022 on the campus of Wageningen University & Research in the Netherlands (51°59’01.0”N 5°39’32.7”E; ca. 10 m). A total of eight hoverfly species were captured, including 44 individuals (n=5.5±2.9 individuals per species, mean±standard deviation) (Figure 1). Flies were captured with hand-held nets and brought in the lab to record their flight within an hour after capture.

The studied hoverfly species shown with their phylogenetic relationships. Phylogenetic positions of the studied species in a larger hoverfly phylogeny (91 species) are shown in the bottom left corner. Sample size as number of individuals for each species is in parentheses. Mean values (± standard deviation) for body mass are presented, along with a photography of a representative individual and a wing from each species. Phylogeny was obtained from (Wong et al., 2023). Photographs were taken by the second author.

Species were first distinguished visually, but their identity was subsequently ascertained using DNA barcoding. This was done by sending a piece of leg of each individual to the Canadian Centre for DNA Barcoding (CCDB), where barcoding was performed following standard automated protocols of the BOLD Identification System (http://www.boldsystems.org/)(Ratnasingham and Hebert, 2007). Sequenced DNA was then compared with a reference library to establish a species match. Sexes were determined visually using criteria from (Gilbert, 2015).

Flight experiments

The flight experiments were carried out in a custom-built octagonal flight arena made of transparent Plexiglas (50×50×48 cm, height×width×length; Figure 2A)(Cribellier et al., 2022). Three synchronised high-speed cameras (Photron FASTCAM SA-X2, set at 5000 frames s−1, spatial resolution of 1084×1084 pixels, shutter speed of 1/10000 s) equipped with a Nikon Sigma 135 mm lens were positioned around the arena to provide a top view, and inclined left and right side view of the flying insects. The camera positioning resulted in a zone of focus of approximately 12 cm3 in the centre of the arena in which body and wing movement were in view and in focus of all three cameras. To increase contrast and therefore facilitate subsequent tracking, the cameras were back-lit using three infrared light panels, placed behind the bottom and lower side walls of the octagon arena, opposite each camera (Figure 2A). Prior to filming, cameras were calibrated with the direct linear transformation (DLT) technique(Hartley and Sturm, 1995) by digitizing a wand of 2.8 cm long moved throughout the zone of focus of the cameras. To track the wand movement, we trained an artificial neural network using the pose estimation Python library DeepLabCut(Nath et al., 2019). Computation of the DLT coefficients was then done using the MATLAB program easyWand (Theriault et al., 2014).

Quantifying the in-flight wingbeat kinematics and wing morphology of hoverflies. (A) Hoverflies were released in an octagon-shaped flight arena. We recorded stereoscopic high-speed videos of the flying hoverflies using three synchronised high-speed video cameras; from the videos we reconstructed the three-dimensional body and wingbeat kinematics. Infrared light panels positioned as the bottom of the arena enabled high contrast between the flying insect and the background. (B) Conventional wing angles were measured at each time step (t=0.0004 s) in the body reference frame. ϕ, wing stroke angle within the stroke plane; η, wing deviation angle out of the stroke plane; θ, wing rotation angle along the spanwise axis; U, air velocity vector relative to the wing; α, angle-of-attack of the wing. (C) Wing morphological parameters including their definitions: wingspan R, wing surface area S, radial position along the span r, local wing chord c at distance r, and the second-moment-of-area S2.

During the flight experiment, all the individuals from a single species were released together in the arena to increase the probability that an individual flew through the intersecting fields of view of the three cameras. For each species, the filming experiment was carried out until a minimum of five flight sequences were obtained, whenever possible. Although this procedure facilitated the recording, this prevented identifying which specific individual or sex was recorded flying, leaving open the possibility of recording the same individual multiple times. Note that this approach aimed at estimating the average wingbeat kinematics per species as we could not directly assign flight measurements at the individual level.

Wingbeats selection

First, we determined the flight speed throughout each recorded flight sequence. We did this by tracking the hoverfly body position throughout the flight sequence using a trained DeepLabCut neural network (Nath et al., 2019). From this we calculated flight speed using a central differentiation scheme. Based on these speed data, we selected the flight sequences with the lowest average flight speed, and then we selected within each of these sequences the wingbeat performed during the slowest part of the flight trajectory. For this selection process, we only included the flight trajectory sections in which the hoverfly was flying straight, to prevent possible variation in wingbeat kinematics stemming from flight manoeuvres. We digitised a minimum of three wingbeats per species, each taken from a different flight sequence. For six of the species we tracked three wingbeats, and for two species (Melanostoma mellinum and Sphaerophoria scripta) we tracked six wingbeats.

Quantifying wingbeat kinematics

We quantified the wing movement using the manual stereoscopic video tracker Kine in MATLAB (Fontaine et al., 2009; Fry et al., 2003) originally designed to track the wingbeat kinematics of fruit flies. The tracker was tailored to hoverflies by building wing models matching the wings of the studied hoverfly species. Wing models were obtained by digitizing the wing outline on high-resolution photographs and consisted of 41 2D coordinates representing the wing shape, with the known hinge and tip positions. To reduce complexity, wings were considered as rigid flat plates, although real hoverflies wings endure significant deformation during the wing stroke (Walker et al., 2010).

We tracked the wingbeat kinematics on down-sampled videos (from 5000 to 2500 frames s−1), as this preserved high enough temporal resolution to capture the wing movement (n=27.9±5.3 frames per wingbeat). The tracking provided us with the position and orientation of the body and wings in each consecutive video frame. The position and orientation of the body were defined in the world reference frame, as a 3D position vector (X=[x,y,z]) and the consecutive Euler angles body yaw, pitch and roll. The angular orientation of the wings was expressed in the hoverfly body coordinate system(Muijres et al., 2014) (Figure 2B), as Euler angles relative to the stroke plane: the stroke angle (ϕ), deviation angle (η) and rotation angle (θ) of each wing (Figure 2B). The stroke plane of each wing was defined as the plane at a 45° angle to the long axis of the body passing through the hinge position of that wing. The three angles (stroke, deviation and rotation angle) were measured separately on the left and right wing and then averaged between the two wings. The time history of the averaged angles was then fitted with a 4th order Fourier series for all three angles (Muijres et al., 2014), and their temporal derivatives were computed analytically (i.e. stroke, deviation and rotation rate). We estimated the absolute angular velocity of the beating wing where and are the stroke rate and deviation rate, respectively. The angle of attack (α) was computed as the angle between the wing plane and the velocity vector of the wing (Figure 2B).

Based on the temporal dynamic of the wing angles, we derived the following parameters: wingbeat frequency (f=1/Δt, where Δt is the duration of the wingbeat), stroke, deviation, and rotation amplitude (defined as Aϕ=|ϕmax −ϕmin|, Aη=|ηmax −ηmin|, and Aθ =|θmax −θmin|, respectively), and both mean and peak wing angular velocity ( and ωpeak, respectively). We quantified body kinematics during each wingbeat using the mean flight speed, climb angle, and the pitch angle of both body and stroke-plane. Flight speed U was estimated as the temporal derivative of body positions, and climb angle as γclimb=atan(Uver/Uhor), where Uver and Uhor are the vertical and horizontal component of the flight velocity. Body pitch angle βbody was calculated as the angle between the body axis and the horizontal. The equivalent stroke-plane pitch angle was defined as the pitch angle of the stroke plane relative to the horizontal, calculated as βstroke-plane=βbody–45°.

To verify whether the studied wingbeats were representative of the overall recorded flight behaviour, we calculated the mean wingbeat frequency over the full flight sequence as , where nwingbeats and Tsequence are the number of wingbeats in a sequence and the sequence duration, respectively. This mean wingbeat frequency was then compared with that of the studied wingbeats. To estimate whether the analysed wingbeats were representative for hovering flight we calculated the advance ratio of all wingbeats as , where is the wingbeat-average angular velocity of the beating wing and R is the span of a single wing. Hovering is generally defined as flight with advance ratios J<0.1, i.e. when the forward flight speed is less than 10% of the wingbeat-induced speed of the wingtip (C. Ellington, 1984a).

Finally, although hoverfly flight speed was minimised in the wingbeat selection procedure, remaining variation in body kinematics may still affect the wingbeat kinematics and therefore confound with changes in wingbeat kinematics resulting from scaling effect. We therefore assessed if body kinematics affected wingbeat kinematics by performing multiple regression analyses where each wingbeat kinematic metric was treated as the dependent variable and flight speed and climb angle as model effects.

Quantifying morphology

After being filmed, individuals were sacrificed by freezing them at -20°C, and weighed with a resolution of 0.1 mg using an analytical balance (XSR204 Mettler Toledo). Wings were then detached from the body and photographed using a digital microscope (Zeiss Stemi SV11). From these photographs, we measured the wing area S, wingspan R and weight-normalised wingspan R*, mean wing chord , second-moment-of-area S2, and the normalised second-moment-of-area using WingImageProcessor in MATLAB (available at https://biomech.web.unc.edu/wing-image-analysis/). Weight-normalised wingspan is defined as R*=R/m1/3, and is a parameter describing isometric variations in wingspan. The normalised second-moment-of-area is defined as , and is a non-dimensional parameter defining wing shape changes independent of changes in wing size. Both parameters allowed us to compare variations in wingspan and wing shape between species of different sizes, respectively.

Second, we used a landmark-based geometric morphometric method to quantify more precisely the variation in wing shape between species (Bookstein, 1997). Wing shape outline was determined using 300 semi-landmarks equidistantly spaced along the wing outline. Semi-landmarks are commonly used to describe curvy shapes such as wing outlines, which lack typical identifiable landmarks (Gunz and Mitteroecker, 2013). Semi-landmarks are allowed to slide along the curve to establish geometric correspondence (Gunz and Mitteroecker, 2013). After this alignment, the landmarks are treated as regular ones in subsequent analyses. We placed one fixed landmark at the base of the wing, fixing the overall landmark configuration with respect to this homologous position available for all specimens. All landmarks were digitised using TpsDig2 (Rohlf, 2015). Wing outlines were superimposed by performing a generalised procrustes analysis (Rohlf and Slice, 1990) using the geomorph R package (Adams and Otárola-Castillo, 2013). This procedure isolates the shape information from the extraneous variations, namely: the size, position, and orientation. To visualize wing shape variation, we then performed a principal component analysis (PCA) on the superimposed coordinates. Shape changes carried on the PCA axes were visualised using the plotRefToTarget function in geomorph.

Testing the influence of phylogeny on flight kinematics and morphological traits

To assess whether shared evolutionary history among the studied species influenced variation in morphology and wingbeat kinematics, we tested if closely related species showed more similar values in the measured parameters than distantly related ones. This was done by computing phylogenetic signals with the Blomberg’s K-statistics (Blomberg et al., 2003) on each flight and morphological parameters using the phytools R package (Revell, 2012). Tests were performed on the mean values per species using the most recent phylogeny of hoverflies obtained from (Wong et al., 2023), pruned to match our sample of eight species. Because none of the parameters showed significant phylogenetic signal (i.e. variation in traits was unaffected by shared ancestry in our species sample), all consecutive analyses were conducted without phylogenetic correction. Whenever possible, we thus performed analyses at the individual level instead of using the mean per species.

To assess the influence of phylogeny on the evolution of body size in hoverflies more generally, we also tested the phylogenetic signal on a larger number of species using data gathered from literature. This was done using mean values of thorax width (a proxy of body size in insects) of 29 hoverfly species. Here, the morphological data was obtained from (Gilbert, 1985), and the phylogeny was obtained from (Wong et al., 2023).

To gain insight into whether hoverfly species most likely underwent size reduction or enlargement over the course of their diversification, we visualised evolutionary changes in body size through the phylogeny using the contmap function from the R package phytools (Revell, 2012). The contmap function maps a continuous trait (here body size) onto the phylogeny by estimating the ancestral states at the internal nodes using maximum likelihood and interpolating the states along each edge using equation 2 of (Felsenstein, 1985) (see (Revell, 2013)).

Modelling the aerodynamics of hoverfly flight

To estimate how changes in wing morphology and wingbeat kinematics affect the aerodynamic forces required for flight, we modelled the aerodynamic force production by flapping wings using a simplified quasi-steady approach as (C. Ellington, 1984b) where F is the aerodynamic thrust force produced during a wingbeat. This aerodynamic force thus varies quadratically with the angular velocity of the beating wing (ω2), and linearly with air density (ρ), the second-moment-of-area of the wing (S2), the wing angle-of-attack (α) and the angle-of-attack-specific thrust coefficient (CFα).

The second-moment-of-area S2 is the only morphological parameter in this equation, thus capturing how wing morphology affects aerodynamic force production, from a first principle. But this second-moment-of-area parameter combines both wing size and shape characteristics (i.e., wing area and its distribution over the spanwise axis (C. Ellington, 1984c)). To disentangle the respective effect of wing shape and size on aerodynamic force production, the second moment-of-area can be decomposed as , where R is the wingspan, is the mean wing chord, and is the normalised second-moment-of-area of the wing. Here, wingspan and average wing chord are size dependent parameters; the non-dimensional second-moment-of-area describes wing shape as the distribution of area along the spanwise wing axis, independently of wing size.

Both angular velocity (ω) and the angle-of-attack (α) of the beating wing are wingbeat kinematics parameters that directly affect aerodynamic force production. A flying insect can vary the angular velocity of its beating wings by changing both its wingbeat frequency and amplitude, as angular velocity scales with the product of wingstroke amplitude and wingbeat frequency (ω∼Aϕ f).

The decomposed model for estimating the effect of wing morphology and kinematics of aerodynamic force production in flapping insect flight is thus

Physical scaling of kinematics and morphology parameters with size

In hovering flight, this aerodynamic thrust force should be directed vertically upwards and equal to the weight of the animal. Thus, to maintain weight support among different sizes of hoverflies, the aerodynamic force should scale linearly with body mass. This cannot be achieved through isometric scaling, i.e. the expected scaling if geometrical similarity is preserved.

For isometric scaling, the morphological parameters of interest scale with body mass as R3 ∼ m1/3, and , and thus both S2 and F scale with body mass as S2∼F∼m4/3. This value is larger than one, and so with an isometric reduction in size the aerodynamic force for weight support reduces more quickly than body mass. Thus, compared to large species of hoverflies, small species need relatively larger wings or adjust their wingbeat kinematics to maintain weight support.

Using our aerodynamic force model (Eqns 1-2) we can estimate both (1) how wing morphology should scale with size if wingbeat kinematics are kept constant, and (2) how the wingbeat kinematics should scale if wing morphology scales isometrically. (1) To maintain weight support with non-varying wingbeat kinematics, the second-moment of-area should scale linearly with body mass, resulting in negative allometry. This could be achieved by negative allometric scaling in the wingspan, chord and normalised second-moment-of-area (R<m1/3, and ), or a combination of these. (2) To maintain weight support when wing morphology scales isometrically, wing angular velocity should scale with body mass as ω∼m−1/6. Again, a hovering hoverfly could achieve this using a comparative adjustment in wingstroke amplitude or wingbeat frequency, or a combination of both. Furthermore, the animal could adjust the aerodynamic force vector using an equivalent adjustment in the wing angle-of-attack α.

To examine how wing morphology and wingbeat kinematics scales with body mass, and whether this scaling is isometrically or allometrically, we regressed body mass (m) against the morphology and kinematics metrics in Eqns 1-2 (, R3, ω2, f, Aϕ and α) using reduced-major-axis (RMA) regressions. RMA is commonly used in allometric studies as it assumes a similar level of error in both variables(Aiello, 1992; Smith, 2009). The symmetric error structure in both variables assumed in RMA makes this method more suited to summarize the relationship between two variables without assuming a causal direction(Warton et al., 2012). In the regression analysis, we used the averaged value at mid-stroke of ω and α, where aerodynamic force production is close to maximum (Tian et al., 2013). Measurements were all log10-transformed prior to regression analysis.

For each morphological metric, we tested if the observed slope significantly deviated from the isometric expectation (expected scaling if geometrical similarity is preserved: R ∼ m1/3; ) using the sma function in the smatr R package (Warton et al., 2012). To better visualize variations in wing shape and size relative to body mass among species, we scaled all wings with the length-scale for isometry (liso=m1/3), and calculated the corresponding weight-normalised wingspan as R* = R/m1/3. We performed additional RMA regressions to test if other kinematic adjustments were associated with changes in body mass, in addition to the ones considered in the aerodynamics model (Eqn 2). This includes regressions of body mass versus the amplitudes of the measured wing angles (Aϕ, Aη, Aθ), and the corresponding peak angular rates .

Computational fluid dynamics simulations

We performed Computational Fluid Dynamics (CFD) simulations to quantify aerodynamic force produced by the different flying hoverfly species. These simulations allowed us to explicitly quantify the effect of size, wingbeat frequency and morphology on aerodynamic force production. Furthermore, we used the CFD results to test the validity of our simplified quasi-steady aerodynamic model (Eqns 1-2).

Each simulation was performed with a single rigid, flat wing with species-specific contours, and the averaged wingbeat kinematics across all species. The computational domain was 8R×8R×8R, and we simulated three consecutive wingbeat cycles. The wing has a thickness of 0.0417R. We assumed hovering flight mode, so no mean inflow was imposed (U∞ = 0 m s-1).

We performed the simulations using our in-house open-source code WABBIT (Wavelet Adaptive Block-Based solver for Insects in Turbulence)(Engels et al., 2022, 2021). The computational approach is based on explicit finite differences that solve the incompressible Navier–Stokes equation, in an artificial compressibility formulation. Wavelets are used to dynamically adapt the discretisation grid to the actual flow at every given time t. This allows us to focus the computational effort where it is required to ensure a given precision, while the grid is significantly coarsened wherever possible. Starting from a coarse base grid, we allow up to seven refinement levels, each one refining the grid by a factor of two. In this work we use the Cohen-Daubechies-Feauveau wavelet CDF42. The grid is composed of blocks of identical size with Bs=22 points in all three spatial directions, yielding an effective maximum resolution of 352 points / R. An additional simulation with eight refinement levels confirmed validity of our results. As demonstrated in numerous validation tests, obtained numerical data can be regarded as quasi-exact in the sense that they are equivalent to an experiment with identical geometry and kinematics. More details on the code can be found in(Engels et al., 2021); the required parameter files to reproduces the simulations presented here can be found in the supplementary. To accelerate the computation, each simulation used up to 600 CPU cores with 2.7 TB of memory in total.

From a fluid dynamics point of view, changing wing length R or wingbeat frequency f solely changes the Reynolds number Re (Batchelor, 1967). For each wing contour, we simulated two different values for Re, one corresponding to a fixed averaged frequency and wing length for all species, and a second one where the wingspan was species-specific. This was initially done to separately assess the combined effect of Reynolds number and wing shape from that of wing shape only. The non-dimensional aerodynamic force was however very similar in both cases, hence the influence of Reynolds number is negligible. In general, the range of Reynolds numbers for the studied hoverfly species is rather narrow (Re ∈ [324,1310]). We finally assessed the contribution of interspecific variation in wingbeat frequency to force production by rescaling all aerodynamic forces with the square of the species-specific and average wingbeat frequency ratio .

Assessing the relative contribution of wing size, shape and kinematics changes to aerodynamic force production

By combining the CFD simulation results with our physical scaling law analysis, we assessed the relative contribution of changes in wing size, shape and kinematics between the studied species on the aerodynamic force production required for weight support.

Because wingbeat kinematics was found to vary little across species and to be uncorrelated with body mass, we based this analysis on the average wingbeat kinematics of all species combined. We did assess the contribution of interspecific variation in wingbeat frequency on aerodynamic force production for weight support, as this varied most between species. For this analysis, we used Eqn 3 to reconstruct the aerodynamic forces produced by each wing type (species), based on both the species-specific wingbeat frequency as well as for the average wingbeat frequency across all species .

We assessed the relative contribution of changes in wing morphology to weight support among species using the following model. According to our quasi-steady aerodynamic model, vertical aerodynamic force for weight support scales with the second-moment-of-area of the wing, which is decomposed into the wing size and shape parameters wing span, mean chord and normalised second-moment-of-area as .

Because isometric scaling results in a reduction in vertical force to weight ratio with decreasing size, smaller insects would require wings with relatively high second-moment-of-area. We estimated how hoverflies achieve this by estimating the scaling between the wing morphology parameters and the vertical aerodynamic force produced by the beating wings, estimated using the CFD simulations. The resulting scaling factors show how variations in wingspan, chord length and normalised second-moment-of-area contribute to producing weight support across sizes. By comparing the scaling factors with those for isometric scaling, we determined how allometric adaptations of wing size and shape alter these aspects among hoverflies of different body sizes.

Results

In this study, we investigated the morphology, and hovering flight biomechanics and aerodynamics of eight hoverfly species, varying in body mass from 5 to 100 mg. The sampled species were relatively well spread throughout the phylogeny of hoverflies (Figure 1). Variation in body mass across species was however unevenly distributed: only one species, Eristalis tenax, weighed more than 50 mg, while the two closely-related species Syrphus nigrilinearus and Eupeodes nielseni had highly similar body masses (Figure 1). Number of males and females from the sampled species was very unbalanced (Table S1), limiting our capacity to investigate for difference between sexes within species. Sexual dimorphism should nonetheless be negligible compared to differences between species.

Influence of phylogeny on the evolution of wingbeat kinematics, wing morphology and body size

There was no significant effect of phylogeny on the flight and morphological parameters among the studied species (n=8) (Table S1). When testing the phylogenetic signal on body size on a larger number of species (n=29), we however detected a marked effect of phylogeny (P=0.005; Blomberg’s K=1.5). The mapping of ancestral state estimates onto the phylogeny suggested that body size has mostly decreased over the course of the diversification of hoverfly species (Figure S1). Although not visible among the eight studied species, variation in body size in hoverflies appears structured by phylogeny overall. Larger sizes are found in the more basal species (e.g. genus Volucella) whereas smaller sizes is found in more recently diverged species (e.g. genus Platycheirus).

No correlation between wingbeats kinematics and body mass

We quantified a total of 33 wingbeats across eight hoverfly species, including at least three wingbeats per species

(Figure 3). The wingbeat frequencies of the studied wingbeats were highly correlated with the average wingbeat frequency calculated on the entire flight sequence (; n=33 flight sequences; r=0.98, P<0.001). Here, the number of wingbeats per flight sequence was 26.2±28.4. This suggests that the digitised wingbeats reliably reflect those of the full flight sequences. Furthermore, the mean advance ratio for the digitised wingbeats was below the generally accepted threshold of 0.1, confirming that the hoverflies were operating in a hovering flight mode.

Relationship between wingbeat kinematics and body mass among hoverfly species. (A-E) Temporal dynamics of the wingbeat kinematics throughout the cycle of all digitised wingbeats. Separate wingbeats are colour coded by body mass (see legend on top), and the black lines show the average wingbeat kinematics for all wingbeats combined. (A-C) Temporal dynamics of the three conventional wingbeat kinematics angle (stroke, deviation, and rotation angle, respectively). (D-E) Angular speed and angle-of-attack of the wings throughout the wingbeat cycle, derived from the stroke and rotation angle, respectively. (F-I) Derived Wingbeat kinematics parameters for each studied species versus the body mass of that species. The kinematics parameters are mean angular wing speed, stroke amplitude, angle-of-attack at mid stroke, and wingbeat frequency. Each data point shows mean ± standard error. None of the wingbeat kinematic parameters were significantly associated with body mass.

The eight hoverfly species exhibited similar wingbeat kinematics (Figure 3A-E) whereby they all moved their wings forward and backwards sinusoidally within the stroke plane (Figure 3A), with a wingstroke amplitude of Aϕ=100°±18° (Figure 3G). Movements out of the stroke-plane, as expressed by the deviation angle, are minimal (Figure 3B). Therefore, the oscillatory angular speed of the wingbeats (Figure 3D) are primarily caused by the wing stroke movement, and peak at approximately 5°×104 s-1 (Figure 3F). During the majority of each wingbeat (forward and backward wingstroke), the wing operates at approximately constant wing rotation angle and angle-of-attack (Figs 3C and 3E, respectively). At the end of the forward and backward wingstroke, the wings rapidly supinate and pronate, respectively. This causes the wing to flip upside down at stroke reversal, allowing it to operate again at the constant wing angle-of-attack of (Figure 3H). The hoverflies flew with a mean body pitch angle of βbody=30°±16°. The corresponding pitch angle of the wing stroke-plane was βstroke-plane=−15°±16°, showing these hoverflies hover with an inclined stroke-plane.

The temporal dynamics of wingbeat kinematics was not visually associated with variation in body mass (Figure 3A-E). Accordingly, none of the derived wingbeat kinematics parameters varied significantly with body mass (Table 1, Figure 3F-I). These include the parameters that directly affect aerodynamic force production (Eqn 1-2, angular velocity ω, angle-of-attack α, stroke amplitude Aϕ and wingbeat frequency f), and the additionally estimated wingbeat kinematic parameters (rotation amplitude Aθ, deviation amplitude Aη, peak rotation velocity and peak deviation velocity ) (Table 1).

Furthermore, variations in wingbeat kinematics were little affected by the hoverfly flight kinematics. Only flight speed had a positive significant effect on the angular velocity of the beating wings (Table S2). Thus, faster flying hoverflies beat their wings at higher angular velocity. When controlling for this effect of flight on wing angular velocity, the regression between angular wing speed and body mass remained non-significant (P=0.41).

These combined results suggest that variation in wingbeat kinematics between species cannot be explained by the demands of maintaining weight support throughout (isometric) size changes. Adjustments for weight support among hoverfly species thus more likely occur through changes in wing morphology.

Wing morphology covaries with body mass

Wing morphology strongly covaried with body mass: the second moment of wing area (S2), wingspan (R), wing chord , and the non-dimensional second moment of wing area all significantly correlated with body mass (Figure 4). Covariation between the wing morphological parameters and body mass all deviated from the isometric expectation (Table 2). Second moment of wing area, wingspan and mean chord length showed all a slight but significant negative allometry (P=0.002, P=0.015 and P=0.014, respectively), and were thus disproportionately larger in smaller species (Figure 4A-C, Table 2). The deviated more strikingly from the isometric trend (P<0.001, Figure 4D, Table 2), indicating a marked change in wing shape associated with body mass variation. Note that among the smallest species the negative correlation between the and body mass reduces (Figs 4D). In fact, for the species weighing less than 20 mg normalised second-moment-of-area remains mostly constant (correlation between and body mass among species weighing less than 20 mg: r=0.34; P=0.157). This suggests an upper limit to the increase in normalised second-moment-of-area with reducing size.

Wing morphology parameters versus body mass for all studied species, including isometric scaling and the observed reduced major axis regression. (A-D) body weight (abscissa) versus on the ordinate the second-moment-of-area, wingspan, mean chord and normalised second-moment-of-area, respectively. Each data point shows results for a different individual, and is color-coded according to species (see top of A). The expected slopes for isometric scaling are indicated by dashed grey lines, and the best fitting reduced major axis regressions are indicated by continuous black lines (see top of B). All fitted regression slopes are significantly lower than expected under isometry, suggesting negative allometric scaling of wing size and shape with respect to body mass.

Our geometric morphometrics analysis revealed two main axes of wing shape variation among the studied species associated with the principal components PC1 and PC2 that explain 65.24% and 19.07% of the variation, respectively) (Figure S2). Shape variation carried on PC1 was strongly correlated with the as well as body mass (r=−0.78; P<0.001) (Figure 5), depicting the changes in wing shape associated with body mass variation among hoverflies species. This change in reflects a wing surface area mostly located distally in smaller species, whereas most wing surface area is located near the wing base in larger hoverflies species (Figure 5).

Changes in wing shape and size associated with body mass variation and according the geometric morphometrics analysis. (A) normalised second-moment-of-area (S2) vs. the primary Principal Component (PC1) of the geometric morphometrics analysis on the wing outlines (left), and weight-normalised wing outlines (with coordinates X*=X/m1/3) color-coded with body mass (see legend) (right). (B) body mass (m) versus the primary Principal Component (PC1) (left), and weight-normalised wing outlines and wingspan (R*) versus body mass (right). Small transparent data points and large opaque data points show values per individual and per species, respectively. All data points and wing outlines in B are color-coded according to species (see right of B). (A) Variation in the normalised second-moment-of-area (S2) were captured well by PC1 (see also Fig. S2). (B) In larger species, wing surface area was located more proximally (lower values on PC 1) than in smaller species, in which wing area tend to be located more distally (higher value on PC 1). The left and right gritted wing shapes on the PC1 axis show the hypothetical wings for PC1=-0.08 and PC1=0.06, respectively. PC1 explains 65.24% of the variations in wing shape (see also figure. S2). (A-B) Combined with the changes in wing shape (left), weight-normalised wing size and wingspan (R*) tend to be larger in smaller species (right).

The combined correlative and morphometrics analyses thus shows that as their size decreases, hoverflies exhibit wings with both larger normalised second-moment-of-area, and relatively larger wings as depicted by the larger weight-normalised wingspan (Figure 5).

Aerodynamic modelling confirms the main role of wing morphology for weight support

As among the eight hoverfly species, wingbeat kinematics did not vary significantly with body mass (P=0.46), we performed all CFD simulations with the average wingbeat kinematics of all species combined (Figure 6A). We did vary wingbeat frequency (Figure 6B), and the wing shape and size between species.

Aerodynamic forces produced by hovering hoverflies, as estimated using Computational Fluid Dynamic (CFD) simulations. (A-B) For our simulations, we used the species-specific wing shapes and sizes (figure 5), and average wingbeat kinematics (A) and the mean wingbeat frequency (B) across all eight studied hoverfly species. (C) The resulting temporal dynamic of vertical forces throughout the wingbeat cycle, coloured by species (see legend above B). (D) The wingbeat-average vertical force versus body mass, for all simulated hoverfly species operating at both the average wingbeat frequency for all species and their species-specific frequency (square and round data points, respectively). Trendlines for weight support (F=mg), and results of the average wingbeat frequency and species-specific frequency simulations are shown as grey dashed, black dashed and black solid lines, respectively. With the species-specific wingbeat frequency (shown in panel B) hoverflies are closer to producing weight support than for the average wingbeat frequency simulations.

The resulting aerodynamic forces differed between species mostly in magnitude, but the temporal dynamics of force production was strikingly similar (Figure 6C and S3). For all species, the temporal dynamics of the vertical aerodynamic force showed a large force peak preceded by a small peak, during each wing stroke (forward and backward stroke). The back stroke produces a larger vertical force and a more distinct higher harmonic. These dynamics are similar between species, despite their different wing shapes. This suggests that wing shape does not strongly affect the characteristics of aerodynamic force production.

The wingbeat-average vertical aerodynamic forces showed a strong scaling with body mass (Figure 6D). Hence, similar wingbeat kinematics but different wing morphology enable achieving weight support in the studied hoverflies species. For each studied species, we estimated aerodynamic forces with wings flapping at the mean wingbeat frequency of all species, and using their species-specific wingbeat frequency (Figure 6B,D). All species were closer to producing weight support with their species-specific wingbeat frequency than for the equivalent average wingbeat frequency case (Figure 6D). Variation in wingbeat frequency thus appeared necessary for individual hoverflies to achieve weight support, although between species wingbeat frequencies do not correlate with body mass.

The correlation between vertical forces estimated using CFD and the different wing morphology parameters (Figure 7A-D; S2, R, and) were strikingly similar to the comparative correlation between body weight and these morphology parameters (Figure 4). This shows that our simplified aerodynamic model (Eqns 1-2) captures the major effects of wing morphology on aerodynamic force production.

Relationships between wing morphology and vertical force production during simulated hovering flight. (A-D) Wingbeat-average vertical force (abscissa) vs. on the ordinate the dimensional and normalised second-moment-of-area, wingspan, and mean wing chord, respectively. Data points are results for different species, as color-coded according to the legend above panel E). The expected slopes for isometric scaling are indicated by dashed grey lines, and the best fitting reduced major axis regressions are indicated by continuous black lines (see top of A). All fitted regression slopes are significantly lower than expected under isometry, suggesting negative allometric scaling. (E) The relative contributions of different aspects of wing morphology (R, , and ) to total aerodynamic force production (100% for ) in the case of isometric scaling and for the observed allometric scaling (in grey and black, respectively).

We continued to assess the relative contribution of the different wing morphology parameters (R, and ) to vertical force production (Figure 7E). This shows that for both the isometric scaling and the observed allometric scaling, variations in wingspan affect the changes in vertical force the most (75% and 74%, respectively). In contrast, the allometric changes in , and especially in, provided most of the allometric change in vertical force production compared to the isometric case. In fact, for our observed allometric case, the changes in mean chord contributed relatively less to vertical force production than for the isometric case (a reduction of 9%). This corresponds to the 11% increase in relative contribution of allometric changes in to the associated aerodynamic forces (Figure 7E). This shows that allometric changes in wing shape, rather than wing size, between species of different sizes have the largest relative effect on achieving weight support.

Discussion

Allometric changes in wing morphology enables miniaturised hoverflies to maintain weight support during hovering

In this study, we investigated whether variation in body mass among hoverfly species is associated with changes in wingbeat kinematics or in wing morphology. Both components directly influence the aerodynamic forces generated during flapping flight (C. Ellington, 1984b), and therefore could accommodate for weight support during evolutionary changes in size occurring throughout the diversification of hoverfly species. Comparing how wingbeat kinematics and wing morphology scale with body mass across hoverfly species, we showed that wing size and shape rather than wingbeat kinematics scale with body mass. This suggest that wing morphology might respond stronger than wingbeat kinematics to the constraints imposed by weight support in hoverflies.

Changes in wing morphology associated with body mass variation are commonly observed in flying animals (Le Roy et al., 2019; Norberg, 2007; Rayner, 1988; Tercel et al., 2018), indicating that wing and body morphology evolve together as an integrated phenotype. Increasing size is usually not only associated with larger wings, but also accompanied by changes in wing shape (Danforth, 1989; Outomuro et al., 2013). Teasing apart the changes in wing morphology resulting from constraints imposed by body mass from that of other selective pressures is however not straightforward. It is generally facilitated by comparing species with similar ecology, because this limits the effect of contrasted ecological pressures at play in different species (García and Sarmiento, 2012). Hoverfly species present a fairly similar ecology in that all adults are specialised flower-visitors (Klecka et al., 2018). Comparable selective regime are thus likely acting on the evolution of flight performance and associated flight kinematics and morphology in adult hoverflies.

Shared ancestry can also affect the evolution of wing morphology and its effect should thus be controlled to understand how body mass may constrain the evolution of other traits (Blomberg et al., 2003). Among the eight species studied here, variation in the measured morphological traits was not explained by phylogenetic distance. The observed changes in wing morphology accompanying body mass variation can thus be assumed free from the influence of phylogeny. This situation however does not apply to all hoverfly species. Indeed, the test of phylogenetic signal on an enlarged number of species revealed that phylogeny explain body mass variation in hoverflies more broadly (Figure S2). Changes in wing morphology resulting from constraints imposed by body mass are therefore expected to be confounded with the effect of phylogeny in a sample including more species. The association between body mass and phylogenetic history was further highlighted by the ancestral state estimation of body size performed on 29 species. This showed that most recently diverged hoverfly species tend to be smaller, and most likely derive from an ancestor of larger size. Throughout the diversification of hoverflies, evolutionary changes in size thus most plausibly showed a decreasing trend, resulting in evolutionary miniaturisation. This is in line with the general trend observed in insect more generally (Polilov, 2015). Hence, our findings are better discussed in the perspective of decreasing size.

As size decreases, wings become relatively less effective in producing the aerodynamic forces required for weight support (Schmidt-Nielsen, 1975). The constraints imposed by weight support during hovering flight should then promote the evolution of disproportionately large wings, or an increased wingbeat frequency or amplitude in smaller hoverflies species. We found that none of wingbeat kinematic parameters scaled with body mass but that wings indeed tended to be disproportionately larger in smaller species (i.e. negative allometry).

Here, all the wing morphology parameters associated with aerodynamic force production exhibited significant negative allometry, including wingspan R, mean chord , and both dimensional and dimensionless second-moment-of-areas (S2 and , respectively). Changes in multiple aspects of the wing morphology may hence contribute to mitigating the negative effect of size reduction on flight ability. A negative allometry between wing and body size has been found in other insect groups: for solitary bees this was found at both the interspecific and intraspecific level (Grula et al., 2021), and at the intraspecific level in rhinoceros beetles (Kawano, 1995). Disproportionately larger wing in larger species (positive allometry) was however found in dragonflies (May, 1981). Smaller insects may not systematically show disproportionately larger wings because of the diversity of mechanisms mitigating weight support involving either morphological or kinematic changes, but also because of contrasted flight modes across taxa (e.g., gliding flight in dragonflies).

The covariation between wing shape and body mass is most apparent among the hoverfly species with body mass above 20 mg, as for species with body masses below this value the normalised second-moment-of-area of the wing remains almost constant (Figure 4D; i.e. Platycheirus clypeatus, Melanostoma mellinum and Tropidia scita in our study). The flattening of the to body mass curve with decreasing size suggests an upper limit in the normalised second-moment-of-area among hoverfly species of approximately 0.58 (Figure 4D). Thus, for small hoverfly species the redistribution of wing area towards the wingtip might be limited, potentially due to structural constraint. The bending moment on the wing hinge and consequently the flight power muscles scale with the third-moment-of-area of the wing , which scales with the product of and R (Muijres et al., 2017). Thus, an increase in comes at a stringent flight muscle force requirement; this might explain the observed plateau in for small hoverfly species. More research is needed to quantify these apparent conflicting evolutionary driving forces on wing shape.

Here we further investigated the relative contribution of the allometric changes in the different aspect of wing morphology to force production by decomposing analytically the allometric component of R, , and (Figure 7). This highlighted that allometric changes in resulted in the largest additional force production compared to isometric scaling. This change in wing shape that most strikingly deviated from isometry was emphasised in our geometric morphometrics analysis. Indeed, in our Principal Component Analysis (PCA) performed on superimposed wing shape coordinates, the primary axis PC1 explained a striking 65% of the variations in wing shape, and was strongly correlated with normalised second-moment-of-area . This PC1 shape variation depicts a shift in the spanwise distribution of wing area, from more proximally located in larger species to more distally located in smaller species (Figure 5). Interestingly, similar changes in wing shape associated with decreasing body size have been found among species of Hymenoptera (Danforth, 1989). An illustration of the most pronounced end of this pattern is the highly spatulated wing shape found in tiny wasp species (e.g. in the Mymaridae or Mymarommatidae families). As highlighted by our combined quasi-steady model and CFD results, such wing shape confers increased lift force production during flapping flight (C. Ellington, 1984b), and hence also contribute to maintaining weight support with decreasing body size. The reduction in body size occurring through the diversification of hoverfly species may thus have favoured wing shapes in smaller species that produce aerodynamic forces more effectively.

Factors decoupling wingbeat kinematics from the influence of body mass

Our study revealed that wingbeat kinematics is relatively stable across hoverfly species, with none of the measured kinematics parameters significantly correlated with body mass. Performing computational fluid dynamic simulations using the average wingbeat kinematics for all species, but incorporating differences in wing morphology, demonstrated that weight support was indeed achieved across species. Important however is that variation in wingbeat frequency – albeit uncorrelated with body mass – has proven crucial for correctly estimating the aerodynamic force necessary to achieve weight support. Adjustments in wingbeat frequency thus appear to play a role in achieving weight support during hovering for individual hoverflies. The hovering hoverflies might thus primarily use wingbeat frequency to finely tune the vertical force required for weight support. Unlike birds and bats, flying insects cannot morph actively their wings to control aerodynamic force production (Harvey et al., 2022). As a result, flying insects rely purely on wingbeat kinematics changes to trim aerodynamic forces and torques during steady flight (Muijres et al., 2017), and to adjust these forces and torques during manoeuvring flight (Hedrick et al., 2009; Muijres et al., 2014). Our CFD simulation results suggest that hovering hoverflies use relatively simple wingbeat frequency modulations to compensate for the continuous daily fluctuations in body mass, as a result of food and water intake and excretion.

The lack of covariation between in-flight wing movement and body mass goes against the prediction that wingbeat kinematics is more likely to be adjusted with varying body mass due to the greater flexibility of flight as compared to morphology. This finding is also surprising given that size variation is generally accompanied with changes in wingbeat kinematics in flying animals (Norberg, 2007; Rayner, 1988; Riskin et al., 2010; Tercel et al., 2018). A previous study dissecting the flapping wing kinematics of miniaturised insects found an increased wing deviation in smaller species enabling higher wing velocity and hence aerodynamic force production(Lyu et al., 2019). Such a change in wing kinematics associated with decreasing size can be explained by the increased viscous drag constraint that tiny flying insects experience, requiring higher aerodynamic force generation. Here we did not find an increased wing deviation angle in smaller hoverfly species, most-likely because our study focus on a too narrow range of size.

While only a handful studies have examined the effect of body mass on detailed wingbeat kinematics in insects (such as wingbeat amplitudes and angle-of-attack Belyaev et al., 2014; Lyu et al., 2019; Meresman and Ribak, 2017), the scaling between body mass and wingbeat frequency has been frequently assessed (Bartholomew and Casey, 1978; Byrne et al., 1988; Casey et al., 1985; De Nadai et al., 2021; Dudley, 1990; Tercel et al., 2018). In theory, an increase in wingbeat frequency is expected in smaller insects because flapping wing-based aerodynamic force production decreases faster than body mass. As such, it is puzzling that the reduction in size among our studied hoverfly species is not associated with increased wingbeat frequency.

Interestingly, inconsistent results on the scaling between body mass and wingbeat frequency have been found among other groups of flying animals. Although predominantly negative (e.g. in butterflies (Dudley, 1990), in orchid bees (Darveau et al., 2005), in mosquitoes(De Nadai et al., 2021), and in a comparative study across insect orders (Tercel et al., 2018)), several studies found weak or no relationship between body mass and wingbeat frequency (e.g. in moths (Bartholomew and Casey, 1978), whiteflies (Byrne et al., 1988), bees (Duell et al., 2022; Grula et al., 2021)). Consistent with our findings, a study comparing the wingbeat kinematics during tethered flight among hoverfly species spanning a range of body mass similar to that investigated here found no significant effect of body mass on wingbeat kinematics (Belyaev et al., 2014).

A possible explanation to the heterogeneous trends found among insect groups is that the extent to which body mass constrains the evolution of wingbeat kinematics depends on species ecology. Adaptation to a particular ecological niche may drive the evolution of specialised flight abilities, promoting a specific wingbeat kinematics. Selective pressures exerted on these wingbeat kinematics may sometimes overcome the constraints imposed by weight support. In that case, relatively constant wingbeat kinematics would be maintained over a range of body mass, and changes in wing morphology alone are expected to adjust for weight support. Consistent with this hypothesis is the decoupling of wingbeat frequency from body mass variation, which has been highlighted in species where hovering flight ability is crucial for food acquisition. These include bees (Duell et al., 2022; Grula et al., 2021), sphingid moth (Bartholomew and Casey, 1978), hummingbirds (Skandalis et al., 2017), and hoverflies (Belyaev et al., 2014) as also in the present study. The absence of correlation between body mass and wingbeat kinematics observed in hoverflies may thus be due to their highly specialised hovering flight abilities.

The selective pressures promoting the evolution of unique hovering flight abilities in hoverflies are not precisely known but hovering steadily is likely advantageous for navigating between flowers when foraging for nectar. Good hovering flight performance may also have been promoted because it increases mating opportunities during the leck gathering of males (Downes, 1969; Gilbert, 1984; Heinrich and Pantle, 1975). Selection for increased foraging efficacy and/or mating opportunities may have maintained consistent wingbeat kinematics across hoverfly species despite variation in body mass.

Conclusion

Altogether, our results suggest that the hovering flight abilities of hoverflies may stem from highly specialised wingbeat kinematics that have been conserved throughout their diversification. In contrast, changes in wing morphology have evolved with the aerodynamic constraints associated with the apparent evolutionary miniaturisation of hoverflies. The existence of selective pressures maintaining stable wingbeat kinematics despite variation in body mass implies that high hovering flight performance can only be achieved over a limited range of wingbeat kinematic parameters, which remains to be ascertained.

Our study moreover highlights that unlike many other Diptera, hoverflies make use of a highly-stereotyped inclined stroke-plane wingbeat kinematics for hovering, where during the dorsal-ventral wingstroke the wing moves not only from back to front but also downwards (Mou et al., 2011). This wingbeat kinematics pattern may contribute to the specialised hovering abilities of hoverflies, as it is consistently observed among the eight species studied here, with their stroke-plane pitch angle during hovering being approximately orientated at −30°. Furthermore, despite significant changes in wing morphology and wingbeat frequency among the studied species, the temporal dynamics of aerodynamic force production is strikingly similar between them (Figures 6C and S3). This suggests that not only the wingbeat kinematics is conserved between species, but also the underlying aerodynamics including the relative use of different unsteady aerodynamic mechanisms (Sane, 2003). Thus, the apparent highly-specialised flight style of hoverflies is maintained among several species with a large range of body masses, and robust against variations in wing shape and wingbeat frequency.

List of symbols and abbreviations

eLife Assessment

This important study addresses the question of how wing morphology and kinematics changed as insect species miniaturized. The authors found no significant correlation between body size and wing kinematics across eight hoverfly species, and instead argue that evolutionary changes in wing size and shape enabled flight in smaller species. However, if the integrative approach to animal biomechanics is strong, the evidence supporting the general conclusion that changes in wing morphology, rather than kinematics, correlate with miniaturization is incomplete and would benefit from more detailed biomechanical analysis and improved methods for phylogenetic comparison.

Reviewer #1 (Public Review):

Summary:

In "Changes in wing morphology..." Roy et al investigate the potential allometric scaling in wing morphology and wing kinematics in 8 different hoverfly species. Their study nicely combines different new and classic techniques, investigating flight in an important, yet understudied alternative pollinator. I want to emphasize that I have been asked to review this from a hoverfly biology perspective, as I do not work on flight kinematics. I will thus not review that part of the work.

Strengths:

The paper is well-written and the figures are well laid out. The methods are easy to follow, and the rationale and logic for each experiment are easy to follow. The introduction sets the scene well, and the discussion is appropriate. The summary sentences throughout the text help the reader.

Weaknesses:

The ability to hover is described as useful for either feeding or mating. However, several of the North European species studied here would not use hovering for feeding, as they tend to land on the flowers that they feed from. I would therefore argue that the main selection pressure for hovering ability could be courtship and mating. If the authors disagree with this, they could back up their claims with the literature. On that note, a weakness of this paper is that the data for both sexes are merged. If we agree that hovering may be a sexually dimorphic behaviour, then merging flight dynamics from males and females could be an issue in the interpretation. I understand that separating males from females in the movies is difficult, but this could be addressed in the Discussion, to explain why you do not (or do) think that this could cause an issue in the interpretation.

The flight arena is not very big. In my experience, it is very difficult to get hoverflies to fly properly in smaller spaces, and definitely almost impossible to get proper hovering. Do you have evidence that they were flying "normally" and not just bouncing between the walls? How long was each 'flight sequence'? You selected the parts with the slowest flight speed, presumably to get as close to hovering as possible, but how sure are you that this represented proper hovering and not a brief slowdown of thrust?

Your 8 species are evolutionarily well-spaced, but as they were all selected from a similar habitat (your campus), their ecology is presumably very similar. Can this affect your interpretation of your data? I don't think all 6000 species of hoverflies could be said to have similar ecology - they live across too many different habitats. For example, on line 541 you say that wingbeat kinematics were stable across hoverfly species. Could this be caused by their similar habitat?

Reviewer #2 (Public Review):

Summary

Le Roy et al quantify wing morphology and wing kinematics across eight hoverfly species that differ in body mass; the aim is to identify how weight support during hovering is ensured. Wing shape and relative wing size vary significantly with body mass, but wing kinematics are reported to be size-invariant. On the basis of these results, it is concluded that weight support is achieved solely through size-specific variations in wing morphology and that these changes enabled hoverflies to decrease in size throughout their phylogenetic history. Adjusting wing morphology may be preferable compared to the alternative strategy of altering wing kinematics, because kinematics may be under strong evolutionary and ecological constraints, dictated by the highly specialised flight and ecology of the hoverflies.

Strengths

The study deploys a vast array of challenging techniques, including flight experiments, morphometrics, phylogenetic analysis, and numerical simulations; it so illustrates both the power and beauty of an integrative approach to animal biomechanics. The question is well motivated, the methods appropriately designed, and the discussion elegantly and convincingly places the results in broad biomechanical, ecological, evolutionary, and comparative contexts.

Weaknesses

(1) In assessing evolutionary allometry, it is key to identify the variation expected from changes in size alone. The null hypothesis for wing morphology is well-defined (isometry), but the equivalent predictions for kinematic parameters remain unclear. Explicit and well-justified null hypotheses for the expected size-specific variation in angular velocity, angle-of-attack, stroke amplitude, and wingbeat frequency would substantially strengthen the paper, and clarify its evolutionary implications.

(2) By relating the aerodynamic output force to wing morphology and kinematics, it is concluded that smaller hoverflies will find it more challenging to support their body mass - a scaling argument that provides the framework for this work. This hypothesis appears to stand in direct contrast to classic scaling theory, where the gravitational force is thought to present a bigger challenge for larger animals, due to their disadvantageous surface-to-volume ratios. The same problem ought to occur in hoverflies, for wing kinematics must ultimately be the result of the energy injected by the flight engine: muscle. Much like in terrestrial animals, equivalent weight support in flying animals thus requires a positive allometry of muscle force output. In other words, if a large hoverfly is able to generate the wing kinematics that suffice to support body weight, an isometrically smaller hoverfly should be, too (but not vice versa). Clarifying the relation between the scaling of muscle force input, wing kinematics, and weight support would resolve the conflict between these two contrasting hypotheses, and considerably strengthen the biomechanical motivation and interpretation.

(3) The main conclusion - that evolutionary miniaturization is enabled by changes in wing morphology - is only weakly supported by the evidence. First, although wing morphology deviates from the null hypothesis of isometry, the difference is small, and hoverflies about an order of magnitude lighter than the smallest species included in the study exist. Including morphological data on these species, likely accessible through museum collections, would substantially enhance the confidence that size-specific variation in wing morphology occurs not only within medium-sized but also in the smallest hoverflies, and has thus indeed played a key role in evolutionary miniaturization. Second, although wing kinematics do not vary significantly with size, clear trends are visible; indeed, the numerical simulations revealed that weight support is only achieved if variations in wing beat frequency across species are included. A more critical discussion of both observations may render the main conclusions less clear-cut, but would provide a more balanced representation of the experimental and computational results.

In many ways, this work provides a blueprint for work in evolutionary biomechanics; the breadth of both the methods and the discussion reflects outstanding scholarship. It also illustrates a key difficulty for the field: comparative data is challenging and time-consuming to procure, and behavioural parameters are characteristically noisy. Major methodological advances are needed to obtain data across large numbers of species that vary drastically in size with reasonable effort, so that statistically robust conclusions are possible.

Reviewer #3 (Public Review):

The paper by Le Roy and colleagues seeks to ask whether wing morphology or wing kinematics enable miniaturization in an interesting clade of agile flying insects. Isometry argues that insects cannot maintain both the same kinematics and the same wing morphology as body size changes. This raises a long-standing question of which varies allometrically. The authors do a deep dive into the morphology and kinematics of eight specific species across the hoverfly phylogeny. They show broadly that wing kinematics do not scale strongly with body size, but several parameters of wing morphology do in a manner different from isometry leading to the conclusion that these species have changed wing shape and size more than kinematics. The authors find no phylogenetic signal in the specific traits they analyze and conclude that they can therefore ignore phylogeny in the later analyses. They use both a quasi-steady simplification of flight aerodynamics and a series of CFD analyses to attribute specific components of wing shape and size to the variation in body size observed. However, the link to specific correlated evolution, and especially the suggestion of enabling or promoting miniaturization, is fraught and not as strongly supported by the available evidence.

The aerodynamic and morphological data collection, modeling, and interpretation are very strong. The authors do an excellent job combining a highly interpretable quasi-steady model with CFD and geometric morphometrics. This allows them to directly parse out the effects of size, shape, and kinematics.

Despite the lack of a relationship between wing kinematics and size, there is a large amount of kinematic variation across the species and individual wing strokes. The absolute differences in Figure 3F - I could have a very large impact on force production but they do indeed not seem to change with body size. This is quite interesting and is supported by aerodynamic analyses.

The authors switch between analyzing their data based on individuals and based on species. This creates some pseudoreplication concerns in Figures 4 and S2 and it is confusing why the analysis approach is not consistent between Figures 4 and 5. In general, the trends appear to be robust to this, although the presence of one much larger species weighs the regressions heavily. Care should be taken in interpreting the statistical results that mix intra- and inter-specific variation in the same trend.

The authors based much of their analyses on the lack of a statistically significant phylogenetic signal. The statistical power for detecting such a signal is likely very weak with 8 species. Even if there is no phylogenetic signal in specific traits, that does not necessarily mean that there is no phylogenetic impact on the covariation between traits. Many comparative methods can test the association of two traits across a phylogeny (e.g. a phylogenetic GLM) and a phylogenetic PCA would test if the patterns of variation in shape are robust to phylogeny.

The analysis of miniaturization on the broader phylogeny is incomplete. The conclusion that hoverflies tend towards smaller sizes is based on an ancestral state reconstruction. This is difficult to assess because of some important missing information. Specifically, such reconstructions depend on branch lengths and the model of evolution used, which were not specified. It was unclear how the tree was time-calibrated. Most often ancestral state reconstructions utilize a maximum likelihood estimate based on a Brownian motion model of evolution but this would be at odds with the hypothesis that the clade is miniaturizing over time. Indeed such an analysis will be biased to look like it produces a lot of changes towards smaller body size if there is one very large taxa because this will heavily weight the internal nodes. Even within this analysis, there is little quantitative support for the conclusion of miniaturization, and the discussion is restricted to a general statement about more recently diverged species. Such analyses are better supported by phylogenetic tests of directedness in the trait over time, such as fitting a model with an adaptive peak or others.

Setting aside whether the clade as a whole tends towards smaller size, there is a further concern about the correlation of variation in wing morphology and changes in size (and the corresponding conclusion about lack of co-evolution in wing kinematics). Showing that there is a trend towards smaller size and a change in wing morphology does not test explicitly that these two are correlated with the phylogeny. Moreover, the subsample of species considered does not appear to recapitulate the miniaturization result of the larger ancestral state reconstruction.

Given the limitations of the phylogenetic comparative methods presented, the authors did not fully support the general conclusion that changes in wing morphology, rather than kinematics, correlate with or enable miniaturization. The aerodynamic analysis across the 8 species does however hold significant value and the data support the conclusion as far as it extends to these 8 species. This is suggestive but not conclusive that the analysis of consistent kinematics and allometric morphology will extend across the group and extend to miniaturization. Nonetheless, hoverflies face many shared ecological pressures on performance and the authors summarize these well. The conclusions of morphological allometry and conserved kinematics are supported in this subset and point to a clade-wide pattern without having to support an explicit hypothesis about miniaturization.

The data and analyses on these 8 species provide an important piece of work on a group of insects that are receiving growing attention for their interesting behaviors, accessibility, and ecologies. The conclusions about morphology vs. kinematics provide an important piece to a growing discussion of the different ways in which insects fly. Sometimes morphology varies, and sometimes kinematics depending on the clade, but it is clear that morphology plays a large role in this group. The discussion also relates to similar themes being investigated in other flying organisms. Given the limitations of the miniaturization analyses, the impact of this study will be limited to the general question of what promotes or at least correlates with evolutionary trends towards smaller body size and at what phylogenetic scale body size is systematically decreasing.

In general, there is an important place for work that combines broad phylogenetic comparison of traits with more detailed mechanistic studies on a subset of species, but a lot of care has to be taken about how the conclusions generalize. In this case, since the miniaturization trend does not extend to the 8 species subsample of the phylogeny and is only minimally supported in the broader phylogeny, the paper warrants a narrower conclusion about the connection between conserved kinematics and shared life history/ecology.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

In "Changes in wing morphology..." Roy et al investigate the potential allometric scaling in wing morphology and wing kinematics in 8 different hoverfly species. Their study nicely combines different new and classic techniques, investigating flight in an important, yet understudied alternative pollinator. I want to emphasize that I have been asked to review this from a hoverfly biology perspective, as I do not work on flight kinematics. I will thus not review that part of the work.

Strengths:

The paper is well-written and the figures are well laid out. The methods are easy to follow, and the rationale and logic for each experiment are easy to follow. The introduction sets the scene well, and the discussion is appropriate. The summary sentences throughout the text help the reader.

We thank the reviewer for these positive comments on our study.

Weaknesses:

The ability to hover is described as useful for either feeding or mating. However, several of the North European species studied here would not use hovering for feeding, as they tend to land on the flowers that they feed from. I would therefore argue that the main selection pressure for hovering ability could be courtship and mating. If the authors disagree with this, they could back up their claims with the literature.

We thank the reviewer for this insight on potential selection pressures on hovering flight. As suggested, we now put the main emphasize on selection related to mating flight (lines 106–111).

On that note, a weakness of this paper is that the data for both sexes are merged. If we agree that hovering may be a sexually dimorphic behaviour, then merging flight dynamics from males and females could be an issue in the interpretation. I understand that separating males from females in the movies is difficult, but this could be addressed in the Discussion, to explain why you do not (or do) think that this could cause an issue in the interpretation.

We acknowledge that not distinguishing sexes in the flight experiment prevents investigating the hypothesis that selection may act especially on male’s flight. This weakness was not addressed in our first manuscript and is now discussed in the revised Discussion section. We nuanced the interpretation and suggested further investigation on flight dimorphism (lines 726–729).

The flight arena is not very big. In my experience, it is very difficult to get hoverflies to fly properly in smaller spaces, and definitely almost impossible to get proper hovering. Do you have evidence that they were flying "normally" and not just bouncing between the walls? How long was each 'flight sequence'? You selected the parts with the slowest flight speed, presumably to get as close to hovering as possible, but how sure are you that this represented proper hovering and not a brief slowdown of thrust?

We very much agree with the reviewer that flight studied in laboratory conditions does not perfectly reflects natural flight behavior. Moreover, having individual hoverflies performing stable hovering in the flight arena, in the intersecting field of view of all three cameras, is quite challenging. Therefore, we do not claim that we studied “true” hovering (i.e. flight speed = 0 m/s), but that we attempted to get as close as possible to true hovering by selecting the flight sections with the lowest flight speeds for our analysis.

In most animal flight studies, hovering is defined as flight with advance ratios J<0.1, i.e. when the forward flight speed is less than 10% of the wingbeat-induced speed of the wingtip (Ellington, 1984a; Fry et al., 2005; Liu and Sun, 2008). By selecting the low flight-speed wingbeats for our analysis, the mean advance ratio in our experiment was 0.08±0.02 (mean±sd), providing evidence that the hoverflies were operating close to a hovering flight mode. This is explained in both the methods and results sections (lines 228–231 and 467–469, respectively).

We however acknowledge that this definition of hovering, although generally accepted, is not perfect. We edited the manuscript to clarify that our experiment does not quantify perfect hovering (lines 186–188). We moreover added the mean±sd duration of the recorded flight sequence from which the slowest wingbeat was selected (line 179), as this info was missing, and we further describe the behaviour of the hoverflies during the experiment (lines 168–169).

Your 8 species are evolutionarily well-spaced, but as they were all selected from a similar habitat (your campus), their ecology is presumably very similar. Can this affect your interpretation of your data? I don't think all 6000 species of hoverflies could be said to have similar ecology - they live across too many different habitats. For example, on line 541 you say that wingbeat kinematics were stable across hoverfly species. Could this be caused by their similar habitat?

We agree with the reviewer that similarity in habitat and ecology might partially explain the similarity in the wingbeat kinematics that we observe. But this similarity in ecology between the eight studied species is in fact a design feature of our study. Here, we aim to study the effect of size on hoverfly flight, and so we designed our study such that we maximize size differences and phylogenetic spread among the eight species, while minimizing variations in habitat, ecology and flight behavior (~hovering). This allows us to best test for the effect of differences in size on the morphology, kinematics and aerodynamics of hovering flight.

Despite this, we agree with the reviewer that it would be interesting to test whether the observed allometric morphological scaling and kinematic similarity is also present beyond the species that we studied. In our revision, we therefore extended our analysis to address this question. Performing additional flight experiments and fluid mechanics simulations was beyond the scope of our current study, but extending the morphological scaling analyses was certainly possible.

In our revised study, we therefore extended our morphological scaling analysis by including the morphology of twenty additional hoverfly species. This extended dataset includes wing morphology data of 74 museum specimens from Naturalis Biodiversity Centre (Leiden, the Netherlands), including two males and two females per species, whenever possible (4.2±1.7 individuals per species (mean±sd)). This extended analysis shows that the allometric scaling of wing morphology with size is robust along the larger sample of species, from a wider range of habitats and ecologies. Nevertheless, we advocate for additional flight measurement in species from different habitats to ascertain the generality of our results (lines 729–732).

Reviewer #2 (Public review):

Summary

Le Roy et al quantify wing morphology and wing kinematics across eight hoverfly species that differ in body mass; the aim is to identify how weight support during hovering is ensured. Wing shape and relative wing size vary significantly with body mass, but wing kinematics are reported to be size-invariant. On the basis of these results, it is concluded that weight support is achieved solely through size-specific variations in wing morphology and that these changes enabled hoverflies to decrease in size throughout their phylogenetic history. Adjusting wing morphology may be preferable compared to the alternative strategy of altering wing kinematics, because kinematics may be under strong evolutionary and ecological constraints, dictated by the highly specialised flight and ecology of the hoverflies.

Strengths

The study deploys a vast array of challenging techniques, including flight experiments, morphometrics, phylogenetic analysis, and numerical simulations; it so illustrates both the power and beauty of an integrative approach to animal biomechanics. The question is well motivated, the methods appropriately designed, and the discussion elegantly and convincingly places the results in broad biomechanical, ecological, evolutionary, and comparative contexts.

We thank the reviewer for appreciating the strengths of our study.

Weaknesses

(1) In assessing evolutionary allometry, it is key to identify the variation expected from changes in size alone. The null hypothesis for wing morphology is well-defined (isometry), but the equivalent predictions for kinematic parameters remain unclear. Explicit and well-justified null hypotheses for the expected size-specific variation in angular velocity, angle-of-attack, stroke amplitude, and wingbeat frequency would substantially strengthen the paper, and clarify its evolutionary implications.

We agree with the reviewer that the expected scaling of wingbeat kinematics with size was indeed unclear in our initial version of the manuscript. In our revised manuscript (and supplement), we now explicitly define how all kinematic parameters should scale with size under kinematic similarity, and how they should scale for maintaining weight support across various sizes. These are explained in the introduction (lines 46–78), method section (lines 316–327), and dedicated supplementary text (see Supplementary Info section “Geometric and kinematic similarity and scaling for weight support”). Here, we now also provide a thorough description of the isometric scaling of morphology, and scaling of the kinematics parameters under kinematic similarity.

(2) By relating the aerodynamic output force to wing morphology and kinematics, it is concluded that smaller hoverflies will find it more challenging to support their body mass - a scaling argument that provides the framework for this work. This hypothesis appears to stand in direct contrast to classic scaling theory, where the gravitational force is thought to present a bigger challenge for larger animals, due to their disadvantageous surface-to-volume ratios. The same problem ought to occur in hoverflies, for wing kinematics must ultimately be the result of the energy injected by the flight engine: muscle. Much like in terrestrial animals, equivalent weight support in flying animals thus requires a positive allometry of muscle force output. In other words, if a large hoverfly is able to generate the wing kinematics that suffice to support body weight, an isometrically smaller hoverfly should be, too (but not vice versa). Clarifying the relation between the scaling of muscle force input, wing kinematics, and weight support would resolve the conflict between these two contrasting hypotheses, and considerably strengthen the biomechanical motivation and interpretation.

The reviewer highlights a crucial aspect of our study: our perspective on the aerodynamic challenges associated with becoming smaller or larger. This comment made us realize that our viewpoint might be unconventional regarding general scaling literature and requires further clarification.

Our approach is focused on the disadvantage of a reduction in size, in contrast with classic scaling theory focusing on the disadvantage of increasing in size. As correctly stated by the reviewer, producing an upward directed force to maintain weight support is often considered as the main challenge, constrained by size. Hereby, researchers often focus on the limitations on the motor system, and specifically muscle force: as animals increase in size, the ability to achieve weight support is limited by muscle force availability. An isometric growth in muscle cannot sustained the increased weight, due to the disadvantageous surface-to-volume ratio.

In animal flight, this detrimental effect of size on the muscular motor system is also present, particularly for large flying birds. But for natural flyers, there is also a detrimental effect of size on the propulsion system, being the flapping wings. The aerodynamic forces produced by a beating wing scales linearly with the second-moment-of-area of the wing. Under isometry, this second-moment-of-area decreases at higher rate than body mass, and thus producing enough lift for weight support becomes more challenging with reducing size. Because we study tiny insects, our study focuses precisely on this constraint on the wing-based propulsion system, and not on the muscular motor system.

We revised the manuscript to better explain how physical scaling laws differentially affect force production by the muscular flight motor system and the wingbeat-induced propulsion system (lines 46–78).

(3) The main conclusion - that evolutionary miniaturization is enabled by changes in wing morphology - is only weakly supported by the evidence. First, although wing morphology deviates from the null hypothesis of isometry, the difference is small, and hoverflies about an order of magnitude lighter than the smallest species included in the study exist. Including morphological data on these species, likely accessible through museum collections, would substantially enhance the confidence that size-specific variation in wing morphology occurs not only within medium-sized but also in the smallest hoverflies, and has thus indeed played a key role in evolutionary miniaturization.

We thank the reviewer for the suggestion to add additional specimens from museum collections to strengthen the conclusions of our work. In our revised study, we did so by adding the morphology of 20 additional hoverfly species, from the Naturalis Biodiversity Centre (Leiden, the Netherlands). This extended dataset includes wing morphology data of 74 museum specimens, and whenever possible we sampled at least two males and two females (4.2±1.7 individuals per species (mean±sd)). This extended analysis shows that the allometric scaling of wing morphology with size is robust along the larger sample of species, including smaller ones. We discuss these additional results now explicitly in the revised manuscript (see Discussion).

Second, although wing kinematics do not vary significantly with size, clear trends are visible; indeed, the numerical simulations revealed that weight support is only achieved if variations in wing beat frequency across species are included. A more critical discussion of both observations may render the main conclusions less clear-cut, but would provide a more balanced representation of the experimental and computational results.

We agree with the reviewer that variations in wingbeat kinematics between species, and specifically wingbeat frequency, are important and non-negligible. As mentioned by the reviewer, this is most apparent for the fact that weight support is only achieved with the species-specific wingbeat frequency. To address this in a more balanced and thorough way, we revised the final section of our analysis approach, by including changes in wingbeat kinematics to that analysis. By doing so, we now explicitly show that allometric changes in wingbeat frequency are important for maintaining weight support across the sampled size range, but that allometric scaling of morphology has a stronger effect. In fact, the relative contributions of morphology and kinematics to maintaining weight-support across sizes is 81% and 22%, respectively (Figure 7). We discuss this new analysis and results now thoroughly in the revised manuscript (lines 621–629, 650–664), resulting in a more balanced discussion and conclusion about the outcome of our study. We sincerely thank the reviewer for suggesting to look closer into the effect of variations in wingbeat kinematics on aerodynamic force production, as the revised analysis strengthened the study and its results.

In many ways, this work provides a blueprint for work in evolutionary biomechanics; the breadth of both the methods and the discussion reflects outstanding scholarship. It also illustrates a key difficulty for the field: comparative data is challenging and time-consuming to procure, and behavioural parameters are characteristically noisy. Major methodological advances are needed to obtain data across large numbers of species that vary drastically in size with reasonable effort, so that statistically robust conclusions are possible.

We thank the reviewer for their encouraging words about the scholarship of our work. We will continue to improve our methods and techniques for performing comparative evolutionary biomechanics research, and are happy to jointly develop this emerging field of research.

Reviewer #3 (Public review):

The paper by Le Roy and colleagues seeks to ask whether wing morphology or wing kinematics enable miniaturization in an interesting clade of agile flying insects. Isometry argues that insects cannot maintain both the same kinematics and the same wing morphology as body size changes. This raises a long-standing question of which varies allometrically. The authors do a deep dive into the morphology and kinematics of eight specific species across the hoverfly phylogeny. They show broadly that wing kinematics do not scale strongly with body size, but several parameters of wing morphology do in a manner different from isometry leading to the conclusion that these species have changed wing shape and size more than kinematics. The authors find no phylogenetic signal in the specific traits they analyze and conclude that they can therefore ignore phylogeny in the later analyses. They use both a quasi-steady simplification of flight aerodynamics and a series of CFD analyses to attribute specific components of wing shape and size to the variation in body size observed. However, the link to specific correlated evolution, and especially the suggestion of enabling or promoting miniaturization, is fraught and not as strongly supported by the available evidence.

We thank the reviewer for the accurate description of our work, and the time and energy put into reviewing our paper. We regret that the reviewer found our conclusions with respect to miniaturization fraught and not strongly supported by the evidence. In our revision, we addressed this by no longer focusing primarily on miniaturization, by extending our morphology analysis to 20 additional species (Figures 4 and 5), improving our analysis of both the kinematics and morphology data (Figure 7), and by discussing our results in a more balanced way (see Discussion). We hope that the reviewer finds the revised manuscript of sufficient quality for publication in eLife.

The aerodynamic and morphological data collection, modeling, and interpretation are very strong. The authors do an excellent job combining a highly interpretable quasi-steady model with CFD and geometric morphometrics. This allows them to directly parse out the effects of size, shape, and kinematics.

We thank the reviewer for assessing our experimental and modelling approach as very strong.

Despite the lack of a relationship between wing kinematics and size, there is a large amount of kinematic variation across the species and individual wing strokes. The absolute differences in Figure 3F - I could have a very large impact on force production but they do indeed not seem to change with body size. This is quite interesting and is supported by aerodynamic analyses.

We agree with the reviewer that there are important and non-negligible variations in wingbeat kinematics between species. As mentioned by the reviewer, although these kinematics do not significant scale with body mass, the interspecific variations are important for maintaining weight support during hovering flight. We thus also agree with the reviewer that these kinematics variations are interesting and deserve further investigations.

In our revised study, we did so by including these wingbeat kinematic variations in our analysis on the effect of variations in morphology and kinematics on aerodynamic force production for maintaining in-flight weight support across the sampled size range (lines 422–444, Figure 7). By doing so, we now explicitly show that variations in wingbeat kinematics are important for maintaining weight across sizes, but that allometric scaling of morphology has a stronger effect. In fact, the relative contributions of adaptations in morphology and kinematics to maintaining weight support across sizes is 81% and 22%, respectively (Figure 7). We discuss these new analysis and results now in the revised manuscript (lines 621–629, 650–664), resulting in a more balanced discussion about the relative importance of adaptations in morphology and kinematics. We hope the reviewer appreciates this newly added analysis.

The authors switch between analyzing their data based on individuals and based on species. This creates some pseudoreplication concerns in Figures 4 and S2 and it is confusing why the analysis approach is not consistent between Figures 4 and 5. In general, the trends appear to be robust to this, although the presence of one much larger species weighs the regressions heavily. Care should be taken in interpreting the statistical results that mix intra- and inter-specific variation in the same trend.

We agree that it was sometimes unclear whether our analysis is performed at the individual or species level. To improve clarity and avoid pseudoreplication, we now analyze all data at the species level, using phylogenetically informed analyses. Because we think that showing within-species variation is nonetheless informative, we included dedicated figures to the supplement (Figures S3 and S5) in which we show data at the individual level, as equivalent to figures 4 and 5 with data at the species level. Note that this cannot be done for flight data due to our experimental procedure. Indeed, we performed flight experiments with multiple individuals in a single experimental setup, pseudoreplication is thus possible for these flight data. This is explained in the manuscript (lines 167–175). All morphological measurements were however done on a carefully organized series of specimens and thus pseudoreplication is hereby not possible.

The authors based much of their analyses on the lack of a statistically significant phylogenetic signal. The statistical power for detecting such a signal is likely very weak with 8 species. Even if there is no phylogenetic signal in specific traits, that does not necessarily mean that there is no phylogenetic impact on the covariation between traits. Many comparative methods can test the association of two traits across a phylogeny (e.g. a phylogenetic GLM) and a phylogenetic PCA would test if the patterns of variation in shape are robust to phylogeny.

After extending our morphological dataset from 8 to 28 species, by including 20 additional species from a museum collection, we increased statistical power and found a significant phylogenetic signal on all morphological traits, except for the second moment of area (lines 458–460, Table S2). Although we do not detect an effect of phylogeny on flight traits, likely due to the limited number of species for which flight was quantified (n=8), we agree with the reviewer’s observation that the absence of a phylogenetic signal does not rule out the potential influence of phylogeny on the covariation between traits. This is now explicitly discussed in the manuscript (lines 599–608). As mentioned in the previous comment, we now test all relationships between body mass and other traits using phylogenetic generalized least squares (PGLS) regressions, therefore accounting for the impact of phylogeny everywhere. The revised analyses produce sensibly similar results as for our initial study, and so the main conclusions remain valid. We sincerely thank the reviewer for their suggestion for revising our statistical analysis, because the revised phylogenetic analysis strengthens our study as a whole.

The analysis of miniaturization on the broader phylogeny is incomplete. The conclusion that hoverflies tend towards smaller sizes is based on an ancestral state reconstruction. This is difficult to assess because of some important missing information. Specifically, such reconstructions depend on branch lengths and the model of evolution used, which were not specified. It was unclear how the tree was time-calibrated. Most often ancestral state reconstructions utilize a maximum likelihood estimate based on a Brownian motion model of evolution but this would be at odds with the hypothesis that the clade is miniaturizing over time. Indeed such an analysis will be biased to look like it produces a lot of changes towards smaller body size if there is one very large taxa because this will heavily weight the internal nodes. Even within this analysis, there is little quantitative support for the conclusion of miniaturization, and the discussion is restricted to a general statement about more recently diverged species. Such analyses are better supported by phylogenetic tests of directedness in the trait over time, such as fitting a model with an adaptive peak or others.

We thank the reviewer for their expert insight in our ancestral state estimate of body size. We agree that the accuracy of this estimate is rather low. Based on the comments by the reviewer we have now revised our main analysis and results, by no longer basing it on the apparent evolutionary miniaturization of hoverflies, but instead on the observed variations in size in our studied hoverfly species. As a result, we removed the figure mapping ancestral state estimates (called figure S1 in the first version) from the manuscript. We now explicitly mention that ascertaining the evolutionary directedness of body size is beyond the scope of our work, but that we nonetheless focus on the aerodynamic challenge of size reduction (lines 609–615).

Setting aside whether the clade as a whole tends towards smaller size, there is a further concern about the correlation of variation in wing morphology and changes in size (and the corresponding conclusion about lack of co-evolution in wing kinematics). Showing that there is a trend towards smaller size and a change in wing morphology does not test explicitly that these two are correlated with the phylogeny. Moreover, the subsample of species considered does not appear to recapitulate the miniaturization result of the larger ancestral state reconstruction.

As also mentioned above, we agree with the reviewer that we cannot ascertain the trajectory of body size evolution in the diversification of hoverflies. We therefore revised our manuscript such that we do no longer focus explicitly on miniaturization; instead, we discuss how morphology and kinematics scale with size, independently of potential trends over the phylogeny. To do so, we revised the title, abstract results and discussion accordingly.

Given the limitations of the phylogenetic comparative methods presented, the authors did not fully support the general conclusion that changes in wing morphology, rather than kinematics, correlate with or enable miniaturization. The aerodynamic analysis across the 8 species does however hold significant value and the data support the conclusion as far as it extends to these 8 species. This is suggestive but not conclusive that the analysis of consistent kinematics and allometric morphology will extend across the group and extend to miniaturization. Nonetheless, hoverflies face many shared ecological pressures on performance and the authors summarize these well. The conclusions of morphological allometry and conserved kinematics are supported in this subset and point to a clade-wide pattern without having to support an explicit hypothesis about miniaturization.

The reviewer argues here fully correct that we should be careful about extending our analysis based on eight species to hoverflies in general, and especially to extend it to miniaturization in this family of insects. As mentioned above, we therefore do no longer specifically focus on miniaturization. Moreover, we extended our analysis by including the morphology of 20 additional species of hoverflies, sampled from a museum collection. We hope that the reviewer agrees with this more balanced and focused discussion of our study.

The data and analyses on these 8 species provide an important piece of work on a group of insects that are receiving growing attention for their interesting behaviors, accessibility, and ecologies. The conclusions about morphology vs. kinematics provide an important piece to a growing discussion of the different ways in which insects fly. Sometimes morphology varies, and sometimes kinematics depending on the clade, but it is clear that morphology plays a large role in this group. The discussion also relates to similar themes being investigated in other flying organisms. Given the limitations of the miniaturization analyses, the impact of this study will be limited to the general question of what promotes or at least correlates with evolutionary trends towards smaller body size and at what phylogenetic scale body size is systematically decreasing.

We thank the reviewer for their encouraging words about the importance of our work on hoverfly flight. As suggested by the reviewer, we narrowed down the main question of our study by no longer focusing on apparent miniaturization, but instead on the correlation between wing morphology, wingbeat kinematics and variations in size.

In general, there is an important place for work that combines broad phylogenetic comparison of traits with more detailed mechanistic studies on a subset of species, but a lot of care has to be taken about how the conclusions generalize. In this case, since the miniaturization trend does not extend to the 8 species subsample of the phylogeny and is only minimally supported in the broader phylogeny, the paper warrants a narrower conclusion about the connection between conserved kinematics and shared life history/ecology.

We truly appreciated the reviewer’s positive assessment of the importance of our work and study. We also thank the reviewer for their advice to generalize the outcome of our work in a more balanced way. Based on the above comments and suggestions of the reviewer, we did so by revising several aspects of our study, including adding additional species to our study, amending the analysis, and revising the title, abstract, results and discussion sections. We hope that the reviewer warrants the revised manuscript of sufficient quality for final publication in eLife.

Recommendations For The Authors:

Reviewer #1 (Recommendations for the authors):

Figure S1 is lovely. I would recommend merging it with Figure 1 so that it does not disappear.

We appreciate the reviewer comment. However, reviewer 3 had several points of concern about the underlying analysis, which made us realize that our ancestral state estimation analysis does not conclusively support a miniaturization trend. We therefore are no longer focusing on miniaturization when interpreting our results.

Figure 4 is beautiful. The consistent color coding throughout is very helpful.

We thank the reviewer for this comment.

Sometimes spaces are missing before brackets, and sometimes there are double brackets, or random line break.

We did our best to remove these typos.

Should line 367 refer to Table S2?

Table S2 is now referred to when mentioning the result of phylogenetic signal (line 460 in the revised manuscript)

Can you also refer to Figure 2 on line 377?

Good suggestion, and so we now do so (line 462 in the revised manuscript).

Lines 497-512: Please refer to relevant figures.

We now refer to figure 4, and its panels (lines 621–629 in the revised manuscript).

Figure legend 1: Do you need to say that the second author took the photos?

We removed this reference.

Figure legend 4: "(see top of A and B)" is not aligned with the figure layout.

We corrected this.

Figure 5 seems to have a double legend, A, B then A, B. Panel A says it's color-coded for body mass, but the figure seems to be color-coded for species.

Thank you for noting this. We corrected this in the figure legend.

Figure 6 legend: Can you confidently say that they were hovering, or do you need to modify this to flying?

The CFD simulations were performed in full hovering (U¥=0 m/s), but any true flying hoverflies will per definition never hover perfectly. But as explained in our manuscript, we define a hovering flight mode as flying with advance ratios smaller than 0.1 (Ellington, 1984a). Based on this we can state that our hoverflies were flying in a hovering mode. We hope that the reviewer agrees with this approach.

Reviewer #2 (Recommendations for the authors):

Below, I provide more details on the arguments made in the public review, as well as a few additional comments and observations; further detailed comments are provided in the word document of the manuscript file, which was shared with the authors via email (I am not expecting a point-by-point reply to all comments in the word document!).

We thank the reviewer for this detailed list of additional comments, here and in the manuscript. As suggested by the reviewer, we did not provide a point-by-point respond to all comments in the manuscript file, but did take them into account when improving our revised manuscript. Most importantly, we now define explicitly kinematic similarity as the equivalent from morphological similarity (isometry), we added a null hypothesis and the proposed references, and we revised the figures based on the reviewer suggestions.

Null hypotheses for kinematic parameters.

Angular amplitudes should be size-invariant under isometry. The angular velocity is more challenging to predict, and two reasonable options exist. Conservation of energy implies:

W = 1/2 I ω2

where I is the mass moment of inertia and W is the muscle work output (I note that this result is approximate, for it ignores external forces; this is likely not a bad assumption to first order. See the reference provided below for a more detailed discussion and more complicated calculations). From this expression, two reasonable hypotheses may be derived.

First, in line with classic scaling theory (Hill, Borelli, etc), it may be assumed that W∝m; isometry implies that I∝m5/3 from which ω ∝m-1/3 follows at once. Note well the implication with respect to eq. 1: isometry now implies F∝m2/3, so that weight support presents a bigger challenge for larger animals; this result is completely analogous to the same problem in terrestrial animals, which has received much attention, but in strong contrast to the argument made by the authors: weight support is more challenging for larger animals, not for smaller animals.

Second, in line with recent arguments, one may surmise that the work output is limited by the muscle shortening speed instead, which, assuming isometry and isophysiology, implies ω ∝m0 = constant; smaller animals would then indeed be at a seeming disadvantage, as suggested by the authors (but see below).

The following references contain a more detailed discussion of the arguments for and against these two possibilities:

Labonte, D. A theory of physiological similarity for muscle-driven motion. PNAS, 2023, 120, e2221217120

Labonte, D.; Bishop, P.; Dick, T. & Clemente, C. J. Dynamics similarity and the peculiar allometry of maximum running speed. Nat Comms., 2024, 15, 2181

Labonte, D. & Holt, N. Beyond power limits: the kinetic energy capacity of skeletal muscle. bioRxiv doi: 10.1101/2024.03.02.583090, 2024

Polet, D. & Labonte, D. Optimising the flow of mechanical energy in musculoskeletal systems through gearing. bioRxiv doi: 10.1101/2024.04.05.588347, 2024

Labonte et al 2024 also highlight that, due to force-velocity effects, the scaling of the velocity that muscle can impart will fall somewhere in between the extremes presented by the two hypotheses introduced above, so that, in general, the angular velocity should decrease with size with a slope of around -1/6 to -2/9 --- very close to the slope estimated in this manuscript, and to data on other flying animals.

We greatly appreciate the reviewer's detailed insights on null hypotheses for kinematics, along with the accompanying references. As noted in the Public Review section (comment/reply 2.3), our study primarily explores how small-sized insects adapt to constraints imposed by the wing-based propulsion system, rather than by the muscular motor system.

In this context, we chose to contrast the observed scaling of morphology and flight traits with a hypothetical scenario of geometric similarity (isometry) and kinematic similarity, where all size-independent kinematic parameters remain constant with body mass. While isometric expectations for morphological traits are well-defined (i.e., ), those for kinematic traits are more debatable (as pointed out by the reviewer). For this reason, we believe that adopting a simple approach based on kinematic similarity across sizes (f~m0, etcetera) enhances the interpretability of our results and strengthens the overall narrative.

Size range

The study would significantly benefit from a larger size range; it is unreasonable to ask for kinematic measurements, as these experiments become insanely challenging as animals get smaller; but it should be quite straightforward for wing shape and size, as this can be measured with reasonable effort from museum specimens. In particular, if a strong point on miniaturization is to be made, I believe it is imperative to include data points for or close to the smallest species.

We appreciate that the reviewer recognizes the difficulty of performing additional kinematic measurements. Collecting additional morphological data to extend the size range was however feasible. In our revised study, we therefore extended our morphological scaling analysis by including the morphology of twenty additional hoverfly species. This extended dataset includes wing morphology data of 74 museum specimens (4.2±1.7 individuals per species (mean±sd)) from Naturalis Biodiversity Centre (Leiden, the Netherlands). This increased the studied mass range of our hoverfly species from 5 100 mg to 3 132 mg, and strengthened our results and conclusions on the morphological scaling in hoverflies.

Is weight support the main problem?

Phrasing scaling arguments in terms of weight support is consistent with the classic literature, but I am not convinced this is appropriate (neither here nor in the classic scaling literature): animals must be able to move, and so, by strict physical necessity, muscle forces must exceed weight forces; balancing weight is thus never really a concern for the vast majority of animals. The only impact of the differential scaling may be a variation in peak locomotor speed (this is unpacked in more detail in the reference provided above). In other words, the very fact that these hoverfly species exist implies that their muscle force output is sufficient to balance weight, and the arguably more pertinent scaling question is how the differential scaling of muscle and weight force influences peak locomotor performance. I appreciate that this is beyond the scope of this study, but it may well be worth it to hedge the language around the presentation of the scaling problem to reflect this observation, and to, perhaps, motivate future work.

We agree with the reviewer that a question focused on muscle force would be inappropriate for this study, as muscle force and power availability is not under selection in the context of hovering flight, but instead in situation where producing increased output is advantageous (for example during take-off or rapid evasive maneuvers). But as explained in our revised manuscript (lines 81-85), we here do not focus on the scaling of the muscular motor with size and throughout phylogeny, but instead we focus on scaling of the flapping wing-based propulsion system. For this system there are known physical scaling laws that predict how this propulsion system should scale with size (in morphology and kinematics) for maintaining weight-support across sizes. In our study, we test in what way hoverflies achieve this weight support in hovering flight.

Of course, it would be interesting to also test how peak thrust is produced by the propulsion system, for example during evasive maneuvers. In the revised manuscript, we now explicitly mention this as potential future research (lines 733–735).

Other relevant literature

Taylor, G. & Thomas, A. Evolutionary biomechanics: selection, phylogeny, and constraint, Oxford University Press, 2014

This book has quite detailed analyses of the allometry of wing size and shape in birds in an explicit phylogenetic context. It was a while ago that I read it, but I think it may provide much relevant information for the discussion in this work.

Schilder, R. J. & Marden, J. H. A hierarchical analysis of the scaling of force and power production by dragonfly flight motors J. Exp. Biol., 2004, 207, 767

This paper also addresses the question of allometry of flight forces (if in dragonflies). I believe it is relevant for this study, as it argues that positive allometry of forces is partially achieved through variation of the mechanical advantage, in remarkable resemblance to Biewener's classic work on EMA in terrestrial animals (this is discussed and unpacked in more detail also in Polet and Labonte, cited above). Of course, the authors should not measure the mechanical advantage of this work, but perhaps this is an interesting avenue for future work.

We thank the reviewer for these valuable literature suggestions and the insights they offer for future work.

More generally, I thought the introduction misses an opportunity to broaden the perspective even further, by making explicit that running and flying animals face an analogous problem (with swimming likely being a curious exception!); some other references related to the role of phylogeny in biomechanical scaling analyses are provided in the comments in the word file.

The introduction has been revised to better emphasize the generality of the scaling question addressed in our study. Specifically, we now explicitly highlight the similar constraints associated with increasing or decreasing size in both terrestrial and flying animals (lines 53–59). We thank the reviewer for this suggestion, which has improved our manuscript.

Numerical results vs measurements

I felt that the paper did not make the strongest possible use of the very nice numerical simulations. Part of the motivation, as I understood it, was to conduct more complex simulations to also probe the validity of the quasi-steady aerodynamics assumption on which eq. 1 is based. All parameters in eq. 1 are known (or can be approximated within reasonable bounds) - if the force output is evaluated analytically, what is the result? Is it comparable to the numerical simulations in magnitude? Is it way off? Is it sufficient to support body mass? The interplay between experiments and numerics is a main potential strength of the paper, which in my opinion is currently sold short.

We agree with the reviewer that we did not make full use of the numerical simulations results. In fact, we did so deliberately because we aim to focus more on the fluid mechanics of hoverfly flight in a future study. That said, we thank the reviewer for suggesting to use the CFD for validating our quasi-steady model. We now do so by correlating the vertical aerodynamic force with variations in morphology and kinematics (revised Figure 7A). The striking similarity between the predicted and empirical fit shows that the quasi-steady model captures the aerodynamic force production during hovering flight surprisingly well.

Statistics

There are errors in the Confidence Intervals in Tab 2 (and perhaps elsewhere). Please inspect all tables carefully, and correct these mistakes. The disagreement between confidence intervals and p-values suggests a significant problem with the statistics; after a brief consultation with the authors, it appears that this result arises because Standard Major Axis regression was used (and not Reduced Major Axis regression, as stated in the manuscript). This is problematic because SMA confidence intervals become unreliable if the variables are uncorrelated, as appears to be the case for some parameters here (see https://cran.r-project.org/web/packages/lmodel2/vignettes/mod2user.pdf for more details on this point). I strongly recommend that the authors avoid SMA, and use MA, RMA or OLS instead. My recommendation would be to use RMA and OLS to inspect if the conclusions are consistent, in which case one can be shown in the SI; this is what I usually do in scaling papers, as there are some colleagues who have very strong and diverging opinions about which technique is appropriate. If the results differ, further critical analysis may be required.

The reviewer correctly identified an error in the statistical approach: a Standard Major Axis was indeed used under inappropriate conditions. Following Reviewer #3’s comments, the expanded sample size and the resulting increase in statistical power to detect phylogenetic signal, our revised analysis now accounts for phylogenetic effects in these regressions. We therefore now report the results from Phylogenetic Least Square (PGLS) regressions (the phylogenetic equivalent of an OLS).

Figures

Please plot 3E-F in log space, add trendlines, and the expectation from isometry/isophysiology, to make the presentation consistent, and comparison of effect strengths across results more straightforward.

The reviewer probably mentioned Figure 3F-I and not E-F (the four panels depicting the relationships between kinematics variables and body mass). As requested, we added the expectation for kinematic similarity to the revised figure, but prefer to not show the non-significant PGLS fits, as they are not used in any analysis. For completeness, we did add the requested figure in log-space with all trendlines to the supplement (Figure S2), and refer to it in the figure legend.

The visual impression of the effect strength in D is a bit misleading, due to the very narrow y-axis range; it took me a moment to figure this out. I suggest either increasing the y-range to avoid this incorrect impression or to notify the reader explicitly in the caption.

We believe the reviewer is referring to Figure 4D. As rightly pointed out, variation in non-dimensional second moment of area() is very low among species, which is consistent with literature (Ellington, 1984b). We agree that the small range on the y-axis might be confusing, and thus we increased it somewhat. More importantly, we now show, next to the trend line, the scaling for isometry (~m0) and for single-metric weight support. Especially the steepness of the last trend line shows the relatively small effect of on aerodynamic force production. This is even further highlighted by the newly added pie charts of the relative allometric scaling factor, where variations in contribute only 5% to maintaining weight support across sizes.

Despite this small variation, these adaptations in wing shape are still significant and are highly interesting in the context of our work. We now discuss this in more detail in the revised manuscript (lines 645–649).

In Figure 7b, one species appears as a very strong outlier, driving the regression result. Data of the same species seems to be consistent with the other species in 7a, c, and d - where does this strong departure come from? Is this data point flagged as an outlier by any typical regression metric (Cook's distance etc) for the analysis in 7b?

We agree with the reviewer: the species in dark green (Eristalis tenax) appears as an outlier on the in Figure 7B ( vs. vertical force) in our original manuscript. This is most likely due to the narrow range of variation in ( — as the reviewer pointed out in the previous comment — which amplifies differences among species. We expanded the y-axis range in the revised Figure 7, so that the point no longer appears as an outlier (see updated graph, now on Figure 7F).

In Figure 1, second species from the top, it reads "Eristalix tenax" when it is "Eristalis tenax" (relayed info by the Editor).

Corrected.

Reviewer #3 (Recommendations for the authors):

I really like the biomechanical and aerodynamic analyses and think that these alone make for a strong paper, albeit with narrower conclusions. I think it is perfectly valid and interesting to analyze these questions within the scope of the species studied and even to say that these patterns may therefore extend to the hoverflies as a whole group given the great discussion about the shared ecology and behavior of much of the clade. However, the extension to miniaturization is too tenuous. This would need much more support, especially from the phylogenetic methods which are not rigorously presented and likely need additional tests.

We thank the reviewer for the positive words about our study. We agree that our attempt to infer the directedness of size evolution was too simplistic, and thus the miniaturization aspect of our study would need more support. As suggested by the reviewer, we therefore do no longer focus on miniaturization, and thus removed these aspects from the title, abstract and main conclusion of our revised manuscript.

There is a lot of missing data about the tree and the parameters used for the phylogenetic methods that should be added (especially branch lengths and models of evolution). Phylogenetic tests for the relationships of traits should go beyond the analysis of phylogenetic signals in the specific traits. My understanding is also that phylogenetic signal is not properly interpreted as a "control" on the effect of phylogeny. The PCA should probably be a phylogenetic PCA with a corresponding morphospace reconstruction.

We agree with the reviewer that our phylogenetic approach based on phylogenetic signal only was incomplete. In our revised manuscript, we not only test for phylogenetic signal but also account for phylogeny in all regressions between traits and body mass using Phylogenetic Generalized Least Squares (PGLS) regressions. Additionally, we have provided more details about the model of evolution and the parameter estimation method in the Methods section (275–278).

Following the reviewer suggestion, in our revised study we now also performed a phylogenetic PCA instead of a traditional PCA on the superimposed wing shape coordinates. The resulting morphospace was however almost identical to the traditional PCA (Figure S4). We nonetheless included it in the revised manuscript for completion. We thank the reviewer for this suggestion, as the revised phylogenetic analysis strengthens our study as a whole.

For the miniaturization conclusion, my suggestion is a more rigorous phylogenetic analysis of directionality in the change in size across the larger phylogeny. However, even given this, I think the conclusion will be limited because it appears this trend does not hold up under the 8 species subsample. To support that morphology is evolutionarily correlated with miniaturization would for me require an analysis of how the change in body size relates to the change in wing shape and kinematics which is beyond what a scaling relationship does. In other words, you would need to test if the changes in body morphology occur in the same location phylogenetically with a shrinking of body size. I think even more would be required to use the words "enable" or "promote" when referring to the relationship of morphology to miniaturization because those imply evolutionary causality to me. To me, this wording would at least require an analysis that shows something like an increase in the ability of the wing morphological traits preceding the reduction in body size. Even that would likely be controversial. Both seem to be beyond the scope of what you could analyze with the given dataset.

As mentioned in reply 3.1, we agree with the reviewer that the miniaturization aspect of our study would need more support. And thus, as suggested by the reviewer, we therefore do no longer focus primarily on miniaturization, by removing these aspects from the title, abstract and main conclusion of our revised manuscript.

The pseudoreplication should be corrected. You can certainly report the data with all individuals, but you should also indicate in all cases if the analysis is consistent if only species are considered.

As mentioned in the Public Review section, our revised approach avoids pseudoreplication by analyzing all data at the species level. Nonetheless, we have included supplementary figures (Figures S3 and S5) to visualize within-species variation.

My overall suggestion is to remove the analysis of miniaturization and cast the conclusions with respect to the sampling you have. Add a basic phylogenetic test for the correlated trait analysis (like a phylogenetic GLM) which will likely still support your conclusions over the eight species and emphasize the specific conclusion about hoverflies' scaling relationships. I think that is still a very good study better supported by the extent of the data.

We thank the reviewer for the positive assessment of our study, and their detailed and constructive feedback. As suggested by the reviewer, miniaturization is no longer the primary focus of our study, and we revised our analysis by extending the morphology dataset to more species, and by using phylogenetic regressions.

References

Ellington C. 1984a. The aerodynamics of hovering insect flight. III. Kinematics. Philosophical Transactions of the Royal Society of London B: Biological Sciences 305:41–78.

Ellington C. 1984b. The aerodynamics of insect flight. II. Morphological parameters. Phil Trans R Soc Lond B 305:17–40.

Fry SN, Sayaman R, Dickinson MH. 2005. The aerodynamics of hovering flight in Drosophila. Journal of Experimental Biology 208:2303–2318. doi:10.1242/jeb.01612

Liu Y, Sun M. 2008. Wing kinematics measurement and aerodynamics of hovering droneflies. Journal of Experimental Biology 211:2014–2025. doi:10.1242/jeb.016931

Version 2

Abstract

Abstract

Due to physical scaling laws, size greatly affects animal locomotor ability and performance. Whether morphological and kinematic traits always jointly respond to size variation is however poorly known. Here, we examine the relative importance of morphological and kinematic changes in mitigating the consequence of size reduction on aerodynamic force production for weight support in flying insects, focusing on hovering flight of hoverflies (Syrphidae). We compared the morphology of 28 hoverfly species, and the flight biomechanics and aerodynamics of eight species with body masses ranging from 5 to 100 mg. Our study reveals no significant effect of body mass on wingbeat kinematics among species, suggesting that morphological rather than kinematic changes may compensate for the reduction in weight support associated with an isometric reduction in wing size. Computational Fluid Dynamics simulations confirmed that adaptations in wing morphology drive the ability of small hoverfly species to generate weight support, although variations in wingbeat kinematics between species cannot be completely neglected. We specifically show that smaller hoverflies have evolved relatively larger wings, and aerodynamically more effective wing shapes, to mitigate the reduction in aerodynamic weight support with isometric size reduction. Altogether, these results suggest that hoverfly flight underpins highly specialised wingbeat kinematics, which have been largely conserved throughout evolution; instead, evolutionary adaptations in wing morphology enabled flight of small hoverflies.

Introduction

Evolutionary changes in size are common over the course of animal diversification (Clauset and Erwin, 2008; Hanken and Wake, 1993), strongly impacting many aspects of morphology, physiology, and behaviour. Investigating how changes in body size affect other traits (i.e., scaling) is thus essential to understand the causes of traits variation in animals. Traditionally rooted in the field of morphology and physiology (Gould, 1966; Hill, 1950; Pélabon et al., 2014), allometric studies have broadened to explore how more complex traits scale with animal size. These include locomotor kinematics (e.g. flight, stride or swimming kinematics, Bejan and Marden, 2006; Riskin et al., 2010), and behaviour (e.g. mating or escape tactics, Dial et al., 2008). Although behavioural or biomechanical traits are often more noisy and challenging to measure, their inclusion in allometric studies is crucial to better understand trait evolution (Cloyed et al., 2021). Indeed, most phenotypes involve both morphological and behavioural traits, each responding in potentially different ways to size variation (Clemente and Dick, 2023). Difference in scaling patterns between body size and other traits is generally most apparent between species, and can be partly explained by neutral divergence during phylogenetic history (Blomberg et al., 2003). Understanding the origin of trait variation among species thus requires controlling for the influence of phylogeny on allometric relationships (Labonte et al., 2016; Peattie and Full, 2007).

In this study, we address the consequences of size variation between species (i.e. evolutionary allometry) and specifically question the relative importance of wing morphology versus wingbeat kinematic changes in mitigating the effect of size on flight ability.

Flight typically combines a suite of morphological and biomechanical traits, providing a relevant context to jointly examine how morphology and locomotor kinematics change with size. Moreover, the physical scaling laws governing flight allow drawing predictions on the scaling of morphology and kinematics in flying animals. These physical laws set both upper and lower bounds on the size of flying animals, resulting from the need to maintain weight-support by producing upward-directed aerodynamic lift forces. In this context, there are two primary limitations for producing these forces: Limitations on (1) the muscular flight motor system, and on (2) the flapping-wing-based aerodynamic thrust-producing system.

Just like terrestrial animals, larger flying animals face increasing constraints on muscle force production as they grow. This limitation arises because muscle force output scales with muscle cross-sectional area, and therefore under isometry it increases at a lower rate than body mass. Consequently, as body size increases, producing the muscle forces for maintaining weight support requires a positive allometry of muscle cross-sectional area (Alexander, 2013). Conversely, smaller animals are not constrained by muscle force production but instead benefit from a favourable surface-to-volume ratio, allowing for a reduction in musculature for locomotion as size decreases.

Next to this constraint on the muscular motor system, flying animals face also constrains stemming from the scaling between their flapping-wing-based propulsion system and body mass. For large flying animals, the aerodynamic lift (L) for weight support scales linearly with the surface area of the wing (S), and quadratically with the flight speed (U∞) as L∼S U∞2 (Norberg, 2007). Under geometric similarity, wing surface area scales with body mass (m) as S∼m2/3 (Schmidt-Nielsen, 1975). Thus, to maintain weight support during flight (L=mg, where g is gravitational acceleration), larger flying animals exhibit disproportionately larger wings and tend to fly faster (Norberg, 2012). This positive allometry in wing surface area and flight speed with body mass is needed to generate enough lift to stay in the air.

Smaller flying animals such as insects generally fly slower, and thus air movements over the wing are governed by wing flapping and less so by their forward movement. The aerodynamic lift forces produced by flapping wings scales linearly with the wing’s second-moment-of-area (S2) and quadratically with angular wing speed as L∼S2 ω2 (Weis-Fogh, 1973). Thus, for small slow flying animals, and particularly during hovering flight, the lift forces required for weight support scale primarily with the wing’s second-moment-of-area, relative to the wing hinge (Muijres et al., 2017). For isometry, this second-moment-of-wing-area scales with body mass as S2∼m4/3. Because this growth factor of 4/3 is larger than one, a wing’s second-moment-of-area decreases at a higher rate than body mass, and therefore wingbeat-induced aerodynamic lift reduces more rapidly than body weight. Thus, to maintain weight support (mg=L∼S2 ω2), smaller flying animals require disproportionately larger wings, and/or they need to beat their wings at higher wingbeat frequencies or amplitudes, resulting in higher angular speeds of the beating wing. Notably, decreasing size also leads to a drop in Reynolds number, increasing the relative influence of viscous over inertial forces. At the low Reynolds numbers typical of small insects, viscous forces dominate, which may affect how lift is generated during flapping flight (Sane, 2003).

In this study, we aim to test how physical scaling laws cause constrains on the flapping-wing-based aerodynamic thrust-producing system in small insects, and how this drives evolutionary adaptations in wing morphology and wingbeat kinematics. Here, we hypothesize that smaller insects have evolved disproportionately larger wings, and beat their wings at higher wingbeat frequencies, allowing them to maintain weight support despite the detrimental scaling effects of size reduction on their thrust-producing system.

The expected negative allometric scaling in wingbeat frequency with body mass across species is supported by the consistent negative relationship between wingbeat frequency and size observed in most flying animals (Norberg, 2007; Rayner, 1988; Riskin et al., 2010; Tercel et al., 2018). A general trend toward disproportionately larger wings in smaller species is however less striking among species (but see Danforth, 1989; Ellington, 1984a; Outomuro et al., 2013), suggesting that morphological changes in response to size variation are not always straightforward.

Although empirical data often align with expectations drawn from physical laws, the scaling of morphology and kinematic with size can sometimes be challenging to predict. Selective pressures associated with ecological specialisation may influence the evolution of morphology and kinematic, and conflict with the constraints imposed by physical scaling laws. Each component may then adjust for the constraints imposed by weight support in varying degree. For example, changes in morphology have been shown to predominantly mitigate the negative consequences of body size reduction: among hummingbirds, a disproportionate increase in wing area – but little change in wingbeat kinematics – enable weight support among species of different sizes (Skandalis et al., 2017). Similarly, a positive allometry in wing area, but no significant change in wingbeat frequency, was found in relation to body size among species of bees (Duell et al., 2022; Grula et al., 2021). Such discrepancies with the general scaling expectation may point at a particular selective regime maintaining specialised flight kinematics and ability.

Here we focused on an ecologically specialised insect group, the hoverflies, to test if the consequences of size variation are mostly mitigated by changes in wingbeat kinematics or in wing morphology.

Hoverflies (Diptera: Syrphidae) are among the most species-rich groups of nectar-foraging insects and exhibit striking variation in size, ranging from 0.5 mg to over 200 mg (Katzourakis et al., 2001). Despite these large size differences, all adult hoverflies show exceptional hovering abilities. This highly specialised flight behaviour may have been promoted because it plays an important role in mating behaviour: male hoverflies aggregate at specific locations such as under a tree and hover steadily, waiting for passing females to pursue (Gilbert, 1984; Heinrich and Pantle, 1975). Hovering abilities may thus be correlated with mating success (Downes, 1969; Gilbert, 1984).

It may also have been favoured in the context of foraging and navigating around flowers, albeit hovering may not directly relate to feeding as most species land on flowers to feed (Gilbert, 1981). Selection acting on the flight of hoverflies may interfere with general scaling expectations between body mass, wing morphology and wingbeat kinematics.

Here, we study the evolutionary response of wing morphology and wingbeat kinematics to weight support constrains in hovering hoverflies, specifically for small species. Due to the negative scaling with mass of the aerodynamic forces produced by the beating-wing-based aerodynamic thrust-producing system, we hypothesize that tiny hoverflies have evolved relatively large wings or high wingbeat frequencies to maintain weight support during hovering. Because flight behaviour is evolutionarily more labile than morphology (Blomberg et al., 2003), we furthermore hypothesize that adaptations in wingbeat kinematics prevail over wing morphological adaptations. Alternatively, if selection for strong hovering performance is limiting kinematic flexibility, we expect weight support to be maintained across sizes through adaptations in wing morphology rather than in wingbeat kinematics.

We test these hypotheses by examining the scaling relationships between wing morphology and body mass in 28 hoverfly species ranging from 3 to 132 mg, and conjointly assessing how wing morphology and wingbeat kinematics scale with body mass for eight species ranging from 5 to 100 mg. We then combined computational fluid dynamic simulations with quasi-steady aerodynamic modelling to estimate aerodynamic forces produced in hovering flight. Based on this, we determined how the relative changes in wingbeat kinematics and wing morphology contribute to producing weight support during hovering flight of hoverflies.

Material and methods

Collecting and identifying hoverflies

We quantified body and wing morphology in 28 hoverfly species, of which we collected eight species in the wild and twenty from museum specimens at the Naturalis Biodiversity centre (Leiden, the Netherlands). The wild caught live hoverflies were also used for flight experiments, to quantity their wingbeat kinematics during hovering flight (Figure 1).

The studied hoverfly species shown with their phylogenetic relationships.

Sample size as number of individuals for each species is in parentheses. Species marked with coloured circles are displayed on the right, and coloured circles with black outlines are of species for which both morphology and flight kinematics were quantified. Body mass is presented as mean values and standard deviation. Phylogeny was obtained from Wong et al. (2023).

From the museum collection, we sampled 74 individual specimens, aiming to include two males and two females per species whenever possible. This resulted in 4.2±1.7 individuals per species (mean±standard deviation). Museum specimens were selected to maximize both phylogenetic coverage and size range. The eight hoverflies from the wild were collected in September 2022 on the campus of Wageningen University, the Netherlands (51°59’01.0”N 5°39’32.7”E; ca. 10 m). We captured a total 44 individuals, resulting in a sample size of 5.5±2.9 individuals per species. Flies were captured with hand-held nets and brought in the lab to record their flight within an hour after capture, after which morphology was quantified.

The wild species were first distinguished visually, but their species identity was subsequently ascertained using DNA barcoding. This was done by sending a piece of leg of each individual to the Canadian Centre for DNA Barcoding (CCDB), where barcoding was performed following standard automated protocols of the BOLD Identification System (http://www.boldsystems.org/) (Ratnasingham and Hebert, 2007). Sequenced DNA was then compared with a reference library to establish a species match. Sexes were determined visually using criteria from Gilbert (2015).

Flight experiments

The flight experiments were carried out in a custom-built octagonal flight arena made of transparent Plexiglas (50×50×48 cm, height×width×length; Figure 2A) (Cribellier et al., 2022). Stereoscopic high-speed videos were recorded using three synchronised high-speed cameras (Photron FASTCAM SA-X2), equipped with Nikon Sigma 135 mm lenses, and recording at a temporal resolution of 5000 frames s−1, a spatial resolution of 1084×1084 pixels, and a shutter speed of 1/10000 s. The cameras were positioned around the arena to provide a top view, and inclined left and right-side view of the flying insects, which resulted in a zone of focus of approximately 12 cm3 in the centre of the arena in which body and wing movement were in view and in focus of all three cameras. To increase contrast and therefore facilitate subsequent tracking, the cameras were back-lit using three infrared light panels, placed behind the bottom and lower side walls of the octagon arena, opposite each camera (Figure 2A).

Quantifying the in-flight wingbeat kinematics and wing morphology of hoverflies.

(A) Hoverflies were released in an octagon-shaped flight arena. We recorded stereoscopic high-speed videos of the flying hoverflies using three synchronised high-speed video cameras; from the videos we reconstructed the three-dimensional body and wingbeat kinematics. Infrared light panels positioned as the bottom of the arena enabled high contrast between the flying insect and the background. (B) Conventional wing angles were measured at each time step (t=0.0004 s) in the body reference frame. ϕ, wing stroke angle within the stroke plane; η, wing deviation angle out of the stroke plane; θ, wing rotation angle along the spanwise axis; U, air velocity vector relative to the wing; α, angle-of-attack of the wing. (C) Wing morphological parameters including their definitions: wingspan R, wing surface area S, radial position along the span r, local wing chord c at distance r, and the second-moment-of-area S2.

Prior to filming, the stereoscopic camera system was calibrated with the direct linear transformation (DLT) technique (Hartley and Sturm, 1995), by digitising a wand of 2.8 cm long moved throughout the zone of focus of the cameras. To track the wand movement, we trained an artificial neural network using the pose estimation Python library DeepLabCut (Nath et al., 2019), achieving a test error of 3.1 pixels. Computation of the DLT coefficients was then done using the MATLAB program easyWand (Theriault et al., 2014). The calibration root mean square error was 1.21 pixels.

During the flight experiment, all the individuals from a single species were released together in the arena to increase the probability that an individual flew through the intersecting fields of view of the three cameras. For each species, the filming experiment was carried out until a minimum of five suitable flight sequences were obtained, whenever possible. A suitable sequence was defined as a flying individual clearly crossing the intersecting fields of view. Although this procedure facilitated the recording, this prevented identifying which specific individual or sex was recorded flying, leaving open the possibility of recording the same individual multiple times. Note that this approach aimed at estimating the average wingbeat kinematics per species as we could not directly assign flight measurements at the individual level.

Wingbeats selection

First, we determined the flight speed throughout each recorded flight sequence. We did this by tracking the hoverfly body centre, estimated from the mean position between the head and abdomen extremities, throughout the flight sequence (mean duration of the tracked trajectories = 0.13±0.11 sec) using a trained DeepLabCut neural network (Nath et al., 2019) (test error: 6.3 pixels). From this, we calculated flight speed using a central temporal differentiation scheme. Based on these speed data, we selected flight sequences with the lowest average speed and, within each sequence, identified the wingbeat occurring during the slowest part of the trajectory. For this selection process, we only included the flight trajectory sections in which the hoverfly was flying straight at a low climb angle, to prevent possible variation in wingbeat kinematics stemming from flight manoeuvres. We digitised a minimum of three wingbeats per species, each taken from a different flight sequence. For six of the species we tracked three wingbeats, and for two species (Melanostoma mellinum and Sphaerophoria scripta) we tracked six wingbeats. Note that our experiment did not allow capturing perfect hovering flight sequences (i.e. flight speed = 0 m.s-1) but aimed at getting as close as possible to this flight mode. This was evaluated using the advance ratio (see next section).

Quantifying wingbeat kinematics

We quantified the wing movement using the manual stereoscopic video tracker Kine in MATLAB (Fontaine et al., 2009; Fry et al., 2003) originally designed to track the wingbeat kinematics of fruit flies. The tracker was tailored to hoverflies by building wing models matching the wings of the studied hoverfly species. Wing models were obtained by digitising the wing outline on high-resolution photographs and consisted of 41 2D coordinates representing the wing shape, with the known hinge and tip positions. To reduce complexity, wings were considered as rigid flat plates, although real hoverflies wings endure deformations during the wing stroke (Walker et al., 2010).

We tracked the wingbeat kinematics on down-sampled videos (from 5000 to 2500 frames s−1), as this preserved high enough temporal resolution to capture the wing movement (n=27±5 frames per wingbeat). The tracking provided us with the position and orientation of the body and wings in each consecutive video frame. The position and orientation of the body were defined in the world reference frame, as a 3D position vector (X=[x,y,z]) and the consecutive Euler angles body yaw, pitch and roll. The angular orientation of the wings was expressed in the hoverfly body reference frame (Muijres et al., 2014) (Figure 2B), as Euler angles relative to the stroke plane: the stroke angle (ϕ), deviation angle (η) and rotation angle (θ) of each wing (Figure 2B). The stroke plane of each wing was defined as the plane at a 45° angle to the long axis of the body passing through the hinge position of that wing. The three wing kinematics angles (stroke, deviation and rotation angle) were measured separately on the left and right wing and then averaged between the two wings. The time history of the averaged angles was then fitted with a 4th order Fourier series for all three angles (Muijres et al., 2014), and their temporal derivatives were computed analytically (i.e. stroke, deviation and rotation rate). We estimated the absolute angular speed of the beating wing as where 𝛟̇ and η̇ are the stroke rate and deviation rate, respectively. The angle of attack (α) was computed as the angle between the wing plane and the velocity vector of the wing (Figure 2B).

Based on the temporal dynamic of the wing kinematics angles, we derived the following wingbeat-average kinematics parameters: wingbeat frequency, defined as f=1/Δt, where Δt is the duration of the wingbeat; wing stroke, deviation, and rotation amplitudes, defined as Aϕ=|ϕmax −ϕmin|, Aη=|ηmax −ηmin|, and Aθ =|θmax −θmin|, respectively; we estimated the wingbeat-average angular speed as ω) = 2 𝑓 𝐴𝛟; we estimated the mean angle-of-attack α) as the mean value at mid forward and backward wing stroke, where aerodynamic force production is close to maximum (Tian et al., 2013).

We quantified the wingbeat-average body kinematics during each wingbeat using the mean flight speed, climb angle, and the pitch angle of both body and stroke-plane. Flight speed U was estimated as the temporal derivative of body positions, and climb angle as γclimb=atan(Uver/Uhor), where Uver and Uhor are the vertical and horizontal component of the flight velocity. Body pitch angle βbody was calculated as the angle between the body axis and the horizontal. The equivalent stroke-plane pitch angle was defined as the pitch angle of the stroke plane relative to the horizontal, calculated as βstroke-plane=βbody–45°.

To verify whether the studied wingbeats were representative of the overall recorded flight behaviour, we calculated the mean wingbeat frequency over the full flight sequence as f̂ nwingbeats/Tsequence, where nwingbeats and Tsequence are the number of wingbeats in a sequence and the sequence duration, respectively. This mean wingbeat frequency was then compared with that of the studied wingbeats. To estimate whether the analysed wingbeats were representative for hovering flight we calculated the advance ratio of all wingbeats as 𝐽 = U/(ω) 𝑅), where R is the span of a single wing. Hovering is generally defined as flight with advance ratios J<0.1, i.e. when the forward flight speed is less than 10% of the wingbeat-induced speed of the wingtip (Ellington, 1984a).

Finally, although hoverfly flight speed was minimised in the wingbeat selection procedure, remaining variation in body kinematics may still affect the wingbeat kinematics and therefore confound with changes in wingbeat kinematics resulting from scaling effect. We therefore assessed if the flight kinematics metrics flight speed and climb angle affected wingbeat kinematics by performing multiple regression analyses. Here, each wingbeat kinematic metric was treated as dependent variable and flight speed and climb angle as model effects.

Quantifying morphology

We quantified body and wing morphology in both the wild-caught hoverflies and the museum specimens. After being filmed flying, wild-caught individuals were sacrificed by freezing them at -20°C, and weighed with a resolution of 0.1 mg using an analytical balance (XSR204 Mettler Toledo). Wings were then detached from the body and photographed using a digital microscope (Zeiss Stemi SV11). From the museum collection, we only selected specimens with properly flattened (i.e., non-folded) wings, and photographed them using a Nikon D600 camera equipped with a 60mm lens. Here, we photographed the complete animal from above, and the separate wing by it placing parallel to the camera lens. We estimated body mass of museum specimens using the correlation between thorax width and the fresh mass in the wild caught specimens (Figure S1).

From the wing photographs of both the wild and museum specimens, we measured the primary wing morphology parameters using WingImageProcessor in MATLAB (available at https://biomech.web.unc.edu/wing-image-analysis/). These include single-wing area S, wingspan R, mean wing chord c̄, and second-moment-of-area S2. From these we calculated weight-normalised wingspan as R*=R/m1/3, and the normalised second-moment-of-area as 𝑆2∗ = 𝑆2/(𝑅3𝑐̄). The weight-normalized wingspan is a parameter describing relative wingspan, and the normalised second-moment-of-area is a non-dimensional parameter defining wing shape changes independent of changes in wing size. Both parameters allowed us to identify variations in relative wingspan and relative wing shape between species of different sizes, respectively.

Next, we used a landmark-based geometric morphometric method to quantify more precisely the variation in wing shape between species (Bookstein, 1997). Wing shape outline was determined using 300 semi-landmarks equidistantly spaced along the wing outline. Semi-landmarks are commonly used to describe curvy shapes such as wing outlines, which lack typical identifiable landmarks (Gunz and Mitteroecker, 2013). Semi-landmarks are allowed to slide along the curve to establish geometric correspondence (Gunz and Mitteroecker, 2013). After this alignment, the landmarks are treated as regular ones in subsequent analyses. We placed one fixed landmark at the base of the wing, fixing the overall landmark configuration with respect to this homologous position available for all specimens. All landmarks were digitised using TpsDig2 (Rohlf, 2015). Wing outlines were superimposed by performing a generalised Procrustes analysis (Rohlf and Slice, 1990) using the geomorph R package (Adams and Otárola-Castillo, 2013). This procedure isolates the shape information from the extraneous variations, namely: the size, position, and orientation. To visualize wing shape variation, we then performed a phylogenetic Principal Component Analysis (phyloPCA) on the superimposed coordinates using the phytools R package (Revell, 2012). Shape changes carried on the PCA axes were visualised using the plotRefToTarget function in geomorph.

To assess the strength of morphological sexual dimorphism and whether it may interfere with interspecific differences, we tested the effect of species and sex on the morphological parameter using analyses of variance (ANOVAs).

Testing the influence of phylogeny on flight kinematics and morphological traits

To assess whether shared evolutionary history among the studied species influenced variation in morphology and wingbeat kinematics, we tested if closely related species showed more similar values in the measured parameters than distantly related ones. This was done by computing phylogenetic signals with the Blomberg’s K-statistics (Blomberg et al., 2003) on each morphological and flight parameters using the R package phytools (Revell, 2012). Furthermore, we accounted for the effect of phylogeny when assessing the scaling relationships between size and traits using Phylogenetic Generalized Least Squares (PGLS) regression (Symonds and Blomberg, 2014). For the PGLS regressions, we assumed a Brownian motion model of evolution, and the model parameters were estimated using maximum likelihood. Tests were performed on the mean values per species using the most recent phylogeny of hoverflies (Wong et al., 2023). The phylogeny was pruned to align with our sample of 28 species for the morphological analysis and further pruned to include only the eight species used in the flight analysis.

Modelling the aerodynamics of hoverfly flight

To estimate how changes in wing morphology and wingbeat kinematics affect the aerodynamic forces required for flight, we modelled the aerodynamic lift force produced by flapping wings in hovering flight using a simplified quasi-steady approach as (Ellington, 1984b) where F is the wingbeat-average upward-directed aerodynamic lift force produced by a beating wing. This aerodynamic force thus varies quadratically with the angular speed of the beating wing (ω2), and linearly with air density (ρ), the second-moment-of-area of the wing (S2), and the lift force coefficient (CF).

The second-moment-of-area S2 is the only morphological parameter in this equation, thus capturing how wing morphology affects aerodynamic force production, from a first principle. But this second-moment-of-area parameter combines both wing size and shape characteristics (i.e., wing area and its distribution over the spanwise axis (Ellington, 1984c)). To disentangle the respective effect of wing shape and size on aerodynamic force production, the second moment-of-area can be decomposed as 𝑆2 = 𝑆2∗ 𝑅3 𝑐̄, where R is the wingspan, 𝑐̄ is the mean wing chord, and 𝑆2∗is the normalised second-moment-of-area of the wing. Here, wingspan and average wing chord are size dependent parameters; the non-dimensional second-moment-of-area describes wing shape as the distribution of area along the spanwise wing axis, independent of wing size.

The lift force coefficient depends on the angle-of-attack, and for fruit flies this dependency can be modelled as CF = sin(α) CFα, where CFα is the angle-of-attack-specific lift force coefficient (Dickinson and Muijres, 2016). For hoverflies, we here assume a similar scaling of lift with angle-of-attack. Both angular speed (ω) and the angle-of-attack (α) of the beating wing are wingbeat kinematics parameters that directly affect aerodynamic force production. A flying insect can vary the angular speed of its beating wings by changing both its wingbeat frequency f and wing stroke amplitude Aϕ, as angular speed scales linearly with the product of these (ω∼f Aϕ). The decomposed model for estimating the effect of wing morphology and kinematics of aerodynamic lift force production in flapping insect flight is thus We used this model to decompose the relative effect of variations in wingbeat kinematic and wing morphology between species on aerodynamic force production.

Physical scaling of kinematics and morphology parameters with size

In hovering flight, the aerodynamic lift force F should be directed vertically up and equal to the weight of the animal. Thus, to maintain weight support among different sizes of hoverflies, the aerodynamic force should scale linearly with body mass. This cannot be achieved through isometric scaling, i.e. the expected scaling if geometric similarity is preserved.

For isometric scaling of morphology with mass, all length parameters scale with mass as liso∼m1/3, resulting in an isometric scale factor under morphological similarity asim=1/3. As a result, the morphological parameters of interest scale with body mass as R∼m1/3, 𝑐̄∼m1/3 and 𝑆2∗∼m0, and thus both S2 and F scale with body mass as F∼S2∼m4/3 (see supplementary text for detailed derivation). This value is larger than one, and so with an isometric reduction in size the aerodynamic force for weight support reduces more quickly than body mass. Thus, compared to large species of hoverflies, small species need relatively larger wings or adjust their wingbeat kinematics to maintain weight support.

For the wingbeat kinematics, we define kinematic similarity when the kinematics parameters of interest do not scale with size, and should thus remain constant across body mass. Specifically, wingbeat frequency (f), stroke amplitude (Aϕ), angular speed (ω), and angle of attack (α) are all expected to scale under kinematic similarity as f∼m0, Aϕ∼m0, ω∼m0, and α∼m0, respectively. As hoverflies become smaller, deviations from this kinematic similarity might occur to adjust for the reduction in weight support due to isometric size reduction.

Using our aerodynamic force model (Eqns 1-2) we can estimate how the combination of wing morphology and wingbeat kinematics should scale with size, while maintaining weight support for hovering flight across sizes (F=mg). Here, we do so for two extreme cases: (1) weight support is maintained across sizes using allometric scaling of wing morphology only, and thus wingbeat kinematics are kept constant (kinematic similarity); (2) weight support is maintained across sizes using allometric scaling of wingbeat kinematics, while wing morphology scales isometrically.

To maintain weight support with non-varying wingbeat kinematics, the second-moment-of-area should scale linearly with body mass (S2∼m), resulting in negative allometry of S2 relative to body mass (S2∼m1<m4/3). This could be achieved by negative allometric scaling in the wingspan, mean chord or normalised second-moment-of-area (R<m1/3, 𝑐̄<m1/3 and 𝑆2∗<m0), or a combination of these. To maintain weight support via negative allometric scaling of only one of these three metrics, while the others scale isometrically, then these should scale with mass as R∼m2/9, 𝑐̄∼m0 and 𝑆2∗∼m-1/3 (see supplementary text for detailed derivations).

To maintain weight support via allometric scaling of kinematics, while wing morphology scales isometrically (S2∼m4/3), the angular speed of the beating wing should scale with body mass as ω∼m-1/6 (see supplementary text for detailed derivations). Again, a hovering hoverfly could achieve this using a comparative adjustment in wingbeat frequency or wing stroke amplitude (f∼m-1/6 or Aϕ∼m-1/6, respectively), or a combination of both (ω∼f Aϕ∼m-1/6). Furthermore, the animal could adjust the aerodynamic force vector using an equivalent adjustment in the wing angle-of-attack (sin(α)∼m-1/6).

To examine how wing morphology and wingbeat kinematics scales with body mass, and whether this scaling is allometric, we regressed the morphology and wingbeat-average kinematics metrics in Eqns 1-2 (𝑆2∗, 𝑐̄, 𝑅, ω), f, Aϕ and α)) against the body mass (m), using Phylogenetic Generalized Least Square (PGLS) regressions (Symonds and Blomberg, 2014). Measurements were all log10-transformed prior to the regression analysis. For each morphology and kinematics metric, we tested if the observed scaling factor significantly deviates from isometry or kinematic similarity, respectively (R∼m1/3; 𝑐̄∼m1/3; 𝑆2∗∼m0 and ω)∼m0; f∼m0; Aϕ∼m0; α)∼m0, respectively), by verifying if its 95% confidence intervals excluded the similarity slope.

To better visualize variations in wing shape and size relative to body mass among species, we scaled all wings with the length-scale for isometry (liso∼m1/3) and calculated the corresponding weight-normalised wingspan as R*=R/m1/3. We performed additional PGLS regressions to test if other kinematic adjustments were associated with changes in body mass, in addition to the ones considered in the aerodynamics model (Eqns 1-2). This includes regressions of body mass versus the amplitudes of the additional measured wingbeat kinematics angles (Aη, Aθ), and the corresponding peak angular rates (𝛟̇peak, η̇peak, θ̇peak).

The relative contribution of allometric scaling in wing morphology and kinematics to maintaining weight-support across sizes

In this study, we hypothesize that small hoverflies have evolved relatively large wings or high wingbeat frequencies to maintain weight support during hovering. Furthermore, we hypothesize that adaptions in wingbeat kinematics will prevail over adaptions in wing morphology, as kinematics is more flexible than morphology (Blomberg et al., 2003). Alternatively, if selection for specialised hovering flight is limiting kinematic flexibility, weight support in small hoverflies might be achieved primarily through changes in wing morphology rather than in wingbeat kinematics.

We tested these hypotheses by assessing how the potential allometric changes in morphology and kinematics contribute to maintaining weight support across the studied range of hoverfly sizes. We did so by comparing, for each relevant morphological and kinematics parameter, its observed mass-specific allometric scaling factor (aallo) with both the scaling factor for morphologic or kinematic similarity (asim), and the scaling factor for achieving weight support fully by allometric scaling of only that parameter (aws). Hereby, we calculated the relative allometric scaling factor for weight-support as This parameter defines the percentage of which the observed allometric scaling of the specific morphological or kinematics parameter contributes to maintaining weight support during hovering flight, across the studied size range of hoverflies (see supplementary text for detailed derivations).

Computational fluid dynamics simulations

We performed Computational Fluid Dynamics (CFD) simulations to quantify aerodynamic force produced by the different flying hoverfly species. These simulations allowed us to explicitly quantify the effect of size, wingbeat kinematics and morphology on aerodynamic force production. Furthermore, we used the CFD results to test the validity of our simplified quasi-steady aerodynamic model (Eqns 1-2).

Each simulation was performed with a single rigid, flat wing with species-specific contours, and the averaged wingbeat kinematics across all species. The computational domain was 8R×8R×8R, and we simulated three consecutive wingbeat cycles. The wing has a thickness of 0.0417 R. We assumed hovering flight mode, so no mean inflow was imposed (U∞ = 0 m s-1).

We performed the simulations using our in-house open-source code WABBIT (Wavelet Adaptive Block-Based solver for Insects in Turbulence)(Engels et al., 2022, 2021). The computational approach is based on explicit finite differences that solve the incompressible Navier–Stokes equation, in an artificial compressibility formulation. Wavelets are used to dynamically adapt the discretisation grid to the actual flow at every given time t. This allows us to focus the computational effort where it is required to ensure a given precision, while the grid is significantly coarsened wherever possible. Starting from a coarse base grid, we allow up to seven refinement levels, each one refining the grid by a factor of two. In this work we use the Cohen-Daubechies-Feauveau wavelet CDF42. The grid is composed of blocks of identical size with Bs=22 points in all three spatial directions, yielding an effective maximum resolution of 352 points along the wingspan. An additional simulation with eight refinement levels confirmed validity of our results.

As demonstrated in numerous validation tests, obtained numerical data can be regarded as quasi-exact in the sense that they are equivalent to an experiment with identical geometry and kinematics. More details on the code can be found in Engels et al., (2021); the required parameter files to reproduces the simulations presented here can be found in the supplementary information. To accelerate the computation, each simulation used up to 600 CPU cores with 2.7 TB of memory in total.

From a fluid dynamics point of view, changing wing length R or wingbeat frequency f changes the Reynolds number Re (Batchelor, 1967). For each species, we performed a simulation using two different values for Re, one at the average Reynolds number for all species, and a second at the species-specific Reynolds number. This was initially done to separate the effect of Reynolds number and wing shape between species. The non-dimensional aerodynamic force was however very similar in both cases, hence the influence of Reynolds number is negligible.

In general, the range of Reynolds numbers for the studied hoverfly species is rather narrow (Re ∈ [324,1310]). We finally assessed the contribution of interspecific variation in wingbeat frequency to force production by rescaling all aerodynamic forces with the square of the species-specific and average wingbeat frequency ratio (f/ f)G 2.

The relative contribution of variations in wing size, shape and kinematics to aerodynamic force production

By combining the CFD simulation results with our aerodynamic force model (Eqns 1-2) and our physical scaling law analysis, we assessed the relative contribution of changes in wing size, shape and kinematics between the studied species on the aerodynamic force production required for weight support.

Because wingbeat kinematics were found to vary little across species and to be uncorrelated with body mass, we based this analysis on the average wingbeat kinematics of all species combined (see CFD methods above for details). We did assess the contribution of interspecific variation in wingbeat frequency on aerodynamic force production, as this varied most between species. We did so by rescaling all aerodynamic forces from CFD with the square of the species-specific and average wingbeat frequency ratio (f/ f)G 2. According to the quasi-steady model, the vertical aerodynamic force from CFD should scales with the product of the wing’s second-moment-of-area and angular wing speed squared (Eqn 1: 𝐹∼𝑆2 ω2). The angular wing speed scales linearly with wingbeat frequency (ω∼𝑓), and the second-moment-of-wing-area can be decomposed into the wing size and shape parameters wing span, mean chord and normalised second-moment-of-area as 𝑆2 = 𝑅3 𝑐̄ 𝑆2∗(Eqn 2).

In our CFD-based scaling analysis, we first tested the validity of our aerodynamic model (Eqns 1-2) by correlating the CFD-based vertical aerodynamic force with the product of second-moment-of-area and wingbeat-average angular speed squared (Eqn 1: 𝐹∼𝑆2 ω̄2). For this, we used an Ordinary Least Squares (OLS) regression on the log-transformed data to quantify the scaling factor between these metrics. We used the regression statistics to test whether the CFD-based force scaled significantly with 𝑆2 ω̄2 (with cut-off at P<0.05). Based on these results, we then quantified the corresponding relative allometric scaling factor a* (Eqn 3), to assess how allometric scaling of 𝑆2 ω̄2 contribute to changes in aerodynamic force production at different sizes.

Secondly, we estimated the relative contribution of the primary variables in 𝑆2 ω̄2 to aerodynamic force production, being second-moment-of-area and wingbeat frequency. We did so by using the same approach as described above, by first determining the scaling of CFD-based aerodynamic force with second-moment-of-area and wingbeat frequency, and secondly by estimating the corresponding relative scaling factors a*. Finally, we used this approach to estimate the relative contributions to aerodynamic force production of the S2-components: wingspan, mean chord length, and normalized second-moment-of-area.

This combined analysis allowed us to estimate the relative contribution of the various morphology and kinematics metrics to variations in vertical force production for weight-support across sizes, and test how hoverflies might have adapted specific wing morphology and kinematics parameters to maintain weight support across sizes. Moreover, comparing the calculated weight-support percentages in our model with 100% for full weight support, allowed us to test the validity of our analysis approach and aerodynamic force model.

Results

In this study, we examined the morphology of 28 hoverfly species, with body masses ranging from 3 to 132 mg. Among them, we analyzed the hovering flight biomechanics and aerodynamics of eight species, with body masses ranging from 5 to 100 mg. The sampled species were well spread throughout the phylogeny of hoverflies (Figure 1). Among the eight species for which we studied flight biomechanics, body mass variation was unevenly distributed. Only one species, Eristalis tenax, weighed more than 50 mg, while the two closely related species Syrphus nigrilinearus and Eupeodes nielseni had highly similar body masses (Figure 1).

Sexual dimorphism in morphological traits was observed in body mass, wing chord, and wing second-moment-of-area (Table S1). However, the direction of this difference varied across species, with males exhibiting larger values in some cases and females in others. Interspecific difference in morphology was much stronger (Table S1), making sexual dimorphism negligible in comparison. While sexual differences in flight kinematics could not be assessed in this study (see Methods), they are likely to be similarly outweighed by interspecies variation.

When testing the phylogenetic signal on wing morphology and body size among the 28 studied species, we detected a significant effect of phylogeny on all traits except for the second-moment-of-area. In contrast, the effect of phylogeny on flight, tested on eight species, was non-significant for all flight traits (Table S2).

No correlation between wingbeats kinematics and body mass

We quantified a total of 33 wingbeats across eight hoverfly species, including at least three wingbeats per species (Figures 2 and 3). The wingbeat frequencies of the studied wingbeats were highly correlated with the average wingbeat frequency calculated on the entire flight sequence (f=G 186±43 Hz; f=< 178±40 Hz; mean±standard deviation; n=33 flight sequences; r=0.98, P<0.001). Here, the number of wingbeats per flight sequence was 26±28. This suggests that the digitised wingbeats reliably reflect those of the full flight sequences. Furthermore, the mean advance ratio for the digitised wingbeats was below the generally accepted threshold of 0.1 (J=G 0.08±0.02), confirming that the hoverflies were operating approximately in a hovering flight mode (Ellington, 1984a).

Wingbeat kinematics during hovering flight of the eight studied hoverfly species.

(A-E) Temporal dynamics of the wingbeat kinematics throughout the wingbeat cycle of all digitised wingbeats. Separate wingbeats are colour coded by body mass (see legend on top), and the black lines show the average wingbeat kinematics for all wingbeats combined. (A-C) Temporal dynamics of the three conventional wingbeat kinematics angle (stroke, deviation, and rotation angle, respectively; see Figure 2 for definitions). (D-E) Angular speed and angle-of-attack of the wings throughout the wingbeat cycle, derived from wing stroke, deviation and rotation angles. (F-I) Derived wingbeat-average wingbeat kinematics parameters for each studied species versus the body mass of that species. The kinematics parameters are wing stroke amplitude Aϕ, wingbeat frequency f, wingbeat-average angular wing speed ω), and mean angle-of-attack at mid wing stroke)α, respectively. Each data point shows mean ± standard error per species, and is colour-coded according to the legend on the top. None of the wingbeat-average wingbeat kinematic parameters were significantly associated with body mass (Table 1). Horizontal dashed lines show the mean parameter values, as expected under kinematic similarity.

The eight hoverfly species exhibited similar wingbeat kinematics (Figures 2 and 3A–E). Here, they all moved their wings forward and backwards sinusoidally within the stroke plane (Figure 3A), with a wing stroke amplitude of Aϕ=100°±18° (Figure 3H). Movements out of the stroke-plane, as expressed by the deviation angle, are minimal (Figure 3B). Therefore, the oscillatory angular speed of the wingbeats (Figure 3D) is primarily caused by the wing stroke movement, and peak at approximately 5°×104 s-1 (Figure 3F). During the majority of each wingbeat (forward and backward wing stroke), the wing operates at approximately constant wing rotation angle and angle-of-attack (Figs 3C and 3E, respectively). At the end of the forward and backward wing stroke, the wings rapidly supinate and pronate, respectively. This causes the wing to flip upside down at stroke reversal, allowing it to operate again at the constant wing angle-of-attack of ᾱ=42°±10° (Figure 3I). The hoverflies flew with a mean body pitch angle of βbody=30°±16°. The corresponding pitch angle of the wing stroke-plane was βstroke-plane=−15°±16°, showing that these hoverflies hover with an inclined stroke-plane.

The temporal dynamics of wingbeat kinematics was apparently not associated with variation in body mass (Figures 3A-E). Accordingly, none of the derived wingbeat-average wing kinematics parameters varied significantly with body mass, although a small negative trend was observed for the wingbeat frequency (Table 1, Figures 3F-I and S2). These include the parameters that directly affect aerodynamic force production (Eqn 1-2, wingbeat-average angular speed ω), angle-of-attack α), stroke amplitude Aϕ and wingbeat frequency f), and the additionally estimated wingbeat kinematic parameters (rotation amplitude Aθ, deviation amplitude Aη, and peak stroke rate 𝛟̇peak, rotation rate θ̇peak and deviation rate η̇peak) (Table S4).

Furthermore, variations in wingbeat kinematics were little affected by the flight kinematics metrics flight speed and climb angle (Table S3). Only flight speed had a positive significant effect on the angular speed of the beating wings, suggesting that faster flying hoverflies beat their wings at higher angular speeds. When controlling for this effect of flight speed on angular wing speed, the regression between angular wing speed and body mass remained non-significant (r2=0.02; P=0.41). Additionally, flight speed was not correlated with body mass (r2=0.01, P=0.43).

These combined results suggest that variation in wingbeat kinematics between species cannot be explained by the demands of maintaining weight support throughout (isometric) size changes (Figure 3). Allometric adjustments for maintaining weight support across sizes among hoverfly species thus more likely occur through changes in wing morphology.

Wing morphology covaries with body mass

Wing morphology strongly covaried with body mass: the wing’s second-moment-of-area (S2), span (R), mean chord (c̄), and non-dimensional second-moment-of-area (𝑆2∗) all significantly correlated with body mass (Figure 4, see also Figure S3). Covariation between the wing morphological parameters and body mass all deviated from the isometric expectation, except for mean chord (Table 1, Figure 4). Second-moment-of-area, wingspan and 𝑆2∗ all showed a significant negative allometry, i.e., they were all disproportionately larger in smaller species (Figure 4A–C, Table 1). Scaling of the second-moment-of-area with size differed the most from isometric scaling for morphological similarity (aallo=1.01 versus asim=1.33) and almost perfectly aligned with the scaling for metric-specific weight-support (aws=1). The corresponding relative scaling factor confirms this (Eqn 3), as it is close to 100% (a*S2=98%; Figure 4E). This suggests that hoverfly species of different size should be able to maintain weight-support during hovering flight almost purely through variations in the second-moment-of-area. By estimating the relative scaling factors of the various morphological parameters that compose the second-moment-of-area (Figure 4F), we find that adaptations in wingspan contribute the most to the variations in second-moment-of-area (a*R=71%; a*c=12%; a*S2*=5%, Figure 4F). Allometric scaling in the non-dimensional second-moment-of-area had the smallest yet significant effect, indicating a small but notable change in wing shape associated with body mass variation.

The scaling of wing morphology with body mass for all 28 studied hoverfly species (A-D), and how allometric scaling of wing morphology contributes to maintaining weight support across the hoverfly size range (E,F).

(A-D) body mass (abscissa) versus on the ordinate the second-moment-of-area S2, wingspan R, mean chord 𝑐̅, and normalized second-moment-of-area 𝑆2∗, respectively. Each data point shows a species average value, where data points of species for which flight was studied are colour-coded (see top row), and species with only quantified morphology are shown in grey (see the corresponding individual-specific data in Figure S3). The black, red and grey trendlines show the best fitting line from a PGLS regression, isometric scaling, and the expected scaling for metric-specific maintenance of weight support, respectively (see legend above A). All fitted PGLS regression slopes, except for wing chord, are significantly lower than expected under isometry, suggesting negative allometric scaling of wing size and shape with respect to body mass (see also Table 1). (E,F) The relative contributions of different wing morphology metrics to maintaining weight-support across the studied range of hoverfly sizes, expressed by the relative allometric scaling factor a*. (E) the relative allometric scaling factor for the second-moment-of-area S2, and (F) the relative allometric scaling factor for the separate morphological components of S2: wingspan R, mean chord 𝑐̅, and normalized second-moment-of-area 𝑆2∗.

Our phylogenetic geometric morphometrics analysis revealed two main axes of wing shape variation among the studied species associated with the principal components PC1 and PC2, explaining 69.21% and 14.64% of the variation, respectively (Figure S4 and S5). Shape variation carried on PC1 was correlated with the 𝑆2∗ (r=-0.49; P=0.009) as well as body mass (r=0.55; P=0.002) (Figure 5), depicting the changes in wing shape associated with body mass variation among hoverflies species. This change in 𝑆2∗ reflects a wing surface area mostly located distally in smaller species, whereas most wing surface area is located near the wing base in larger hoverflies species (Figure 5).

Changes in wing shape and size associated with body mass variation.

Each data point shows a species-average value (see the individual-specific data in Figure S5), and are colour-coded for species of which both flight and morphology were studied (see right column of B), and in grey for species with only morphology quantified. (A) Normalised second-moment-of-area (S2*) versus the primary Principal Component (PC1) of the phylogenetic Principal Component Analysis (phyloPCA) performed on the mean wing shape coordinates of all 28 studied hoverfly species. On the right, we show the 28 wing outlines colour-coded and normalised with body mass (with coordinates X∗=X/m1/3). (B) body mass (m) versus PC1 (left), and mass-normalised wing outlines versus body mass for the eight species used in our flight experiments (right). The left and right gritted wing shapes on the PC1 axis show the theoretical wing shapes at maximum and minimum value of PC1. PC1 explains 65.24% of the variations in wing shape (see also Figures S4 and S5). (A-B) The combined results show that in larger species, wing surface area was located more proximally (lower PC1 and S*2 values) than in smaller species, in which wing area tend to be located more distally (higher PC1 and S*2 values). Combined with the changes in wing shape (left), weight-normalised wingspan (R∗) and mean chord (𝑐̅) tend to be larger in smaller species (right).

The combined correlative and morphometrics analyses thus shows that as their size decreases, hoverflies exhibit relatively larger wings that are also shaped such that their normalised second-moment-of-area is maximized (Figures 4 and 5). These combined morphological adaptations allow smaller hoverflies to maintain in-hovering weight-support without the need to adjust their wingbeat kinematics.

Aerodynamic modelling confirms the main role of wing morphology for weight support

Because among the eight hoverfly species, wingbeat kinematics did not vary significantly with body mass (Table 1), we performed all CFD simulations with the average wingbeat kinematics of all species combined (Figure 6A). We did vary the wing shape and size between species (top row of Figure 6), and we modelled aerodynamic force production with both the species-specific wingbeat frequency and mean wingbeat frequency of all species combined (Figure 6B).

Aerodynamic forces produced by hovering hoverflies, as estimated using Computational Fluid Dynamic (CFD) simulations.

(A-B) For our simulations, we used the species-specific wing shapes and sizes (top row), but average wingbeat kinematics and frequency (𝑓̅) across all eight studied hoverfly species (A,B). (C) The resulting temporal dynamic of vertical forces throughout the wingbeat cycle, coloured by species (see legend on top). (D) The wingbeat-average vertical force versus body mass, for all simulated hoverfly species operating at the average wingbeat frequency for all species 𝑓̅ (square data points and dashed trend line), and forces scaled to the species-specific wingbeat frequency f (round data points in B and D, and solid trend line). The trendline for weight support (F=mg) is shown with a grey dashed line. With the species-specific wingbeat frequency f (solid trend line and circles in B and D) hoverflies are closer to producing weight support than for the simulations at the average wingbeat frequency 𝑓̅ (dashed lines in B and D).

The resulting aerodynamic forces differed between species mostly in magnitude, but the temporal dynamics of force production were strikingly similar (Figure 6C and S6). For all species, the temporal dynamics of the vertical aerodynamic force showed a large force peak preceded by a small peak, during each wing stroke (forward and backward stroke). The back stroke produces a larger vertical force and a more distinct higher harmonic. These dynamics are similar between species, despite their different wing shapes. This suggests that wing shape does not strongly affect the characteristics of aerodynamic force production.

The wingbeat-average vertical aerodynamic forces showed a strong scaling with body mass (Figure 6D), as expected due to the differences in wing size between species (top row in Figure 6). For each studied species, we modelled aerodynamic forces, both at the species-specific wingbeat frequency and at the mean wingbeat frequency of all species (Figure 6B,D). For the simulations with species-specific wingbeat frequency, the hoverflies produced on average weight-support, whereas for the mean wingbeat frequency cases weight-support was not achieved (Figure 6D). Variation in wingbeat frequency thus appeared necessary for individual hoverflies to achieve weight support, although between species wingbeat frequencies do not correlate with body mass.

We continued to analyse the relative effect of variations in wing morphology and wingbeat-frequency on the aerodynamic force production, as quantified from CFD (Figure 7, Table 2). Here, all the tested morphology and kinematics parameters varied significantly with aerodynamic force magnitude, except for wingbeat-frequency and normalized second-moment-of-area.

The relative contribution of allometric variations in wing morphology and wingbeat kinematics to CFD-derived vertical aerodynamic force production, and how this contributes to maintaining in-hovering weight-support across the sampled range of hoverfly sizes.

(A-C, D-F) Wingbeat-average vertical force at species-specific wingbeat frequency f (abscissa) vs. on the ordinate various wing morphology and kinematics traits. In each panel, data points are results for different species, colour-coded according to the legend on the top. The solid and dashed black lines show the significant and non-significant OLS regression fits, respectively. The dashed red and grey lines show the expected slopes for scaling under morphological and kinematic similarity and for 100% metric-specific weight-support, respectively. (G,H) The relative contributions of allometric variations in wing morphology and kinematics to maintaining weight-support across the studied range of hoverfly sizes, expressed by the relative allometric scaling factor a* (Eqn 3). Contributions based on significant and non-significant OLS regressions are shown in dark and light grey, respectively. (A) The wingbeat-average vertical force scales linearly with the product of second-moment-of-area and wingbeat-average angular speed squared (Eqn 1: 𝐹∼𝑆2 ω̄2), resulting in weight-support across all sizes. (B-C) Separating this product into its main components (B and C, respectively) shows that the second-moment-of-area and wingbeat frequency contribute 81% and 22% to maintaining weight support across sizes, respectively (D). (E-H) Parting the contribution of second-moment-of-area into its components (D-F) shows that allometric variations in wingspan, mean chord and nondimensional second-moment-of-area contribute 55%, 19% and 5% to maintaining weight support across sizes, respectively (H).

First, we correlated the vertical forces estimated using CFD with the product of second-moment-of-area and wingbeat-average angular speed squared (Figure 7A). According to our quasi-steady aerodynamic model, this should yield a linear correlation (Eqn 1: 𝐹∼𝑆2 ω̄2), which was also the case as the allometric scaling factor of aall=0.99 [0.97 – 1.01] (n=8) is not significantly different from linear (Table 2). The corresponding allometric scaling factor equals 103%, suggesting that the allometric adaptations that drive variations in 𝑆2 ω̄2 fully explain variations in aerodynamic force production required for maintaining weight support across sizes. It also shows that our quasi-steady aerodynamic model predicts the effect of morphology and kinematics on aerodynamic force production well.

Separating this dependency of aerodynamic forces on 𝑆2 ω̄2 into its primary components, second-moment-of-area and wingbeat frequency (Figure 7B-D), shows that allometric adaptations in second-moment-of-area and wingbeat frequency contribute 81% and 22% to maintaining weight support across sizes, respectively. Hereby, the aerodynamic forces do not scale significantly with wingbeat frequency (Figure 7B; Table 2), as was also the case for the equivalent correlation of wingbeat frequency with body mass (Figure 3F; Table 1). These combined results confirm that adaptations in morphology drive variations in aerodynamic force production between hoverflies species of various sizes, although variations in wingbeat frequency cannot be ignored.

Further parting the effect of S2 on aerodynamic force production into its components (Figure 7E-G) shows that allometric adaptations in wingspan R, mean chord c̄ and normalized second-moment-of-area 𝑆2∗ contribute 55%, 19% and 5% to variations in aerodynamic force production across sizes, respectively (Figure 7H). Here, normalized second-moment-of-area did not significantly scale with vertical force production (Figure 7G; Table 2), as opposed to the significant scaling found in the corresponding analysis with body mass for the extended 28 species dataset (Figure 4D; Table 1).

The sum of the relative scaling factors of all metrics contributing to vertical force production (Eqn 2) equals 101% (Figure 7H), suggesting that these allometric adaptations combined fully explain variations in aerodynamic force production required for maintaining weight support across sizes. The here-identified effects of morphological traits on CFD-based forces are strikingly similar to those from the comparative analysis between morphology and body mass (Figure 4). This shows that our kinematics analysis based on eight species captures the major effects of wing morphology on aerodynamic force production, as expected based on the morphological analysis with the large sample size of 28 species.

Discussion

Allometric changes in wing morphology enables small hoverflies to maintain weight support during hovering

In this study, we investigated whether variation in body mass among hoverfly species is associated with changes in wingbeat kinematics or in wing morphology. Both components directly influence the aerodynamic forces produced by beating wings during flapping flight (Ellington, 1984b), and therefore could accommodate for maintaining weight support during evolutionary changes in size occurring throughout the diversification of hoverfly species. Comparing how wingbeat kinematics and wing morphology scale with body mass across hoverfly species, we showed that wing size and shape rather than wingbeat kinematics scale with body mass. This suggests that wing morphology might respond stronger than wingbeat kinematics to the constraints imposed for maintaining weight support during flight in hoverflies.

Changes in wing morphology associated with body mass variation are commonly observed in flying animals (Le Roy et al., 2019; Norberg, 2007; Rayner, 1988; Taylor and Thomas, 2014; Tercel et al., 2018), indicating that wing and body morphology evolve together as an integrated phenotype. Increasing size is usually not only associated with relatively larger wings, but also accompanied by changes in wing shape (Danforth, 1989; Outomuro et al., 2013). Teasing apart the changes in wing morphology resulting from constraints imposed by body mass from that of other selective pressures is however not straightforward. It is generally facilitated by comparing species with similar ecology, because this limits the effect of contrasted ecological pressures at play in different species (García and Sarmiento, 2012). Hoverfly species present a similar ecology in that all adults are specialised flower visitors (Klecka et al., 2018). Comparable selective regimes are thus likely acting on the evolution of flight performance and associated flight kinematics and morphology in adult hoverflies.

Shared ancestry can also affect the evolution of wing morphology and its effect should thus be controlled to understand how body mass may constrain the evolution of other traits (Blomberg et al., 2003). Among the 28 species studied here, variation in the measured morphological traits was significantly influenced by phylogenetic history. In contrast, among the sub-set of eight species for which we studied flight kinematics, phylogenetic signal in flight traits was non-significant. This may suggest either a higher evolutionary lability in flight compared to morphological traits, or results from limited statistical power due to the small number of species with quantified flight data. Note that the lack of phylogenetic signal does not necessarily preclude an impact of phylogeny on the covariation between traits and body mass. As such, changes in wing morphology (and possibly in wingbeat kinematics) resulting from constraints imposed by body mass are confounded with the effect of shared ancestry. Our phylogenetically informed regressions of morphological and flight traits against body mass nonetheless allow disentangling such effects.

The large size variation among hoverfly species results from 90 million years of evolutionary diversification (Wong et al., 2023). While some taxa may have evolved larger body sizes than their ancestors, others may have undergone size reduction. Determining the exact trajectory of body size evolution in the species studied is beyond the scope of our study. Addressing the aerodynamic challenges of size reduction may not fully represent the evolutionary constraints acting on all hoverfly species. However, we chose to focus on the adaptation of the wing-based aerodynamic thrust-producing system in this context, as it provides a valuable framework for understanding the evolution of flight in small flying insects.

As size decreases, the wings become relatively less effective in producing the aerodynamic forces required for weight support (see Supplementary Text). The constraints imposed by weight support during hovering flight should then promote the evolution of disproportionately large wings, or an increased wingbeat frequency or amplitude in smaller hoverflies species. We found that none of wingbeat kinematic parameters significantly scaled with body mass but that wings indeed tended to be disproportionately larger in smaller species (i.e. negative allometry). Here, size-dependent adaptations in the primarily wing morphology parameter second-moment-of-areas (𝑆2; Eqn 1) almost fully explained the required adaptations for maintaining weight-support across sizes (a*S2=98%; Figure 4E). Parsing the second-moment-of-areas in its separate wing size and shape components shows that both the primary wing size wing shape metrics (wingspan R and dimensionless second-moment-of-areas 𝑆2∗, respectively) exhibit significant negative allometric scaling (Figure 4B, D). These allometric adaptations in wing size and shape are predicted to contribute to maintaining weight-support across sizes for 71% and 5%, respectively (Figure 4F). Thus, changes in multiple aspects of the wing morphology may contribute to mitigating the negative effect of size reduction on flight ability, whereby wing shape effects are small but significant (Figure 5).

A negative allometry between wing size and body size has been found in other insect groups: for solitary bees this was found at both the interspecific and intraspecific level (Grula et al., 2021), and at the intraspecific level in rhinoceros beetles (Kawano, 1995). In contrast, disproportionately larger wing were found in larger species of dragonfly, suggesting positive allometry in these larger insects (May, 1981). This opposite allometric scaling of wing size in dragonflies and hoverflies may be the result of the use of contrasted mechanisms for producing aerodynamic forces, involving both morphological or kinematics variations, and contrasted flight modes (e.g., gliding flight in dragonflies).

The observed negative allometry in wing shape with mass was further emphasised in our geometric morphometrics analysis (Figures 5, S4 and S5). Indeed, in our phylogenetic Principal Component Analysis performed on superimposed wing shape coordinates, the primary axis PC1 explained a striking 69% of the variations in wing shape, and was strongly correlated with normalised second-moment-of-area 𝑆2∗. This PC1 shape variation depicts a shift in the spanwise distribution of wing area, from more proximally located in larger species to more distally located in smaller species (Figure 5). Interestingly, similar changes in wing shape associated with decreasing body size have been found among species of Hymenoptera (Danforth, 1989). An illustration of the most pronounced end of this pattern is the highly spatulated wing shape found in tiny wasp species (e.g. in the Mymaridae or Mymarommatidae families). This shows that although the relative contribution of allometric scaling of wing shape to maintaining weight support is relatively small (5%; Figure 4F), physical constraints associated with size reduction are strong enough to drive these adaptations. Evolutionary changes in the body size of hoverflies may thus have favoured wing shapes in smaller species that produce aerodynamic forces more effectively.

We continued to study how allometric variations in wing morphology and wingbeat kinematics affected aerodynamic force production across the studied range of hoverfly sizes (Figures 6 and 7). For this, we combined flight experiments on eight hoverfly species with both analytical quasi-steady aerodynamic modelling (Eqns 1,2) and numerical aerodynamics modelling (CFD). By combining these results, we first tested the validity of our quasi-steady aerodynamics model by correlating the wingbeat-average vertical forces estimated from CFD with the product of second-moment-of-area and wingbeat-average angular wing speed (𝐹∼𝑆2 ω̄2; Eqn 1; Figure 7A). The resulting linear correlation shows that our quasi-steady aerodynamic model captures the effect of wing morphology and kinematics on aerodynamic force production well. By separating the combined effect of variations in morphology and kinematics on aerodynamic force production into its components (Eqn 2), we confirm that adaptations in morphology drive the variations in aerodynamic force production between hoverfly species of different sizes. Here, wing size variations as expressed by wingspan have the largest effect on aerodynamic force production, but variations in wing shape and wingbeat frequency cannot be completely ignored (Figure 7C,G,H). In fact, the respective 5% and 22% contribution of wing shape and wingbeat frequency to variations in aerodynamic force production, need to be included in our model to maintain 100% weight support across the size range of our hovering hoverflies (Figure 7H).

Factors decoupling wingbeat kinematics from the influence of body mass

Our study revealed that wingbeat kinematics is relatively stable across hoverfly species, with none of the measured kinematics parameters significantly correlated with body mass (Figure 3; Table 1). Performing computational fluid dynamic simulations using the average wingbeat kinematics for all species, but incorporating differences in wing morphology, demonstrated that weight support was indeed achieved across species. Important however is that variation in wingbeat frequency – albeit uncorrelated with body mass – has proven crucial for correctly estimating the aerodynamic force necessary to achieve weight support (Figures 6D and 7C,D). Adjustments in wingbeat frequency thus appear to play a role in achieving weight support during hovering for individual hoverflies. The hovering hoverflies might thus primarily use wingbeat frequency to finely tune the vertical aerodynamic force required for weight support. Unlike birds and bats, flying insects cannot actively morph their wings to control aerodynamic force production (Harvey et al., 2022). As a result, flying insects rely purely on wingbeat kinematics changes to trim aerodynamic forces and torques during steady flight (Muijres et al., 2017), and to adjust these forces and torques during manoeuvring flight (Hedrick et al., 2009; Muijres et al., 2014). Our CFD simulation results suggest that hovering hoverflies use relatively simple wingbeat frequency modulations to compensate for the continuous daily fluctuations in body mass, as a result of food and water intake and excretion. The lack of covariation between in-flight wing movement and body mass goes against the prediction that wingbeat kinematics is more likely to be adjusted with varying body mass due to the greater flexibility of flight kinematics as compared to morphology. This finding is also surprising given that size variation is generally accompanied with changes in wingbeat kinematics in other flying animals (Norberg, 2007; Rayner, 1988; Riskin et al., 2010; Taylor and Thomas, 2014; Tercel et al., 2018). A previous study dissecting the flapping wing kinematics of miniaturised insects found an increased wing deviation in smaller species enabling higher wing velocity and hence aerodynamic force production (Lyu et al., 2019). Such a change in wing kinematics associated with decreasing size can be explained by the increased viscous drag constraint that tiny flying insects experience at low Reynolds numbers, requiring higher aerodynamic force generation. Here we did not find an increased wing deviation angle in smaller hoverfly species, most-likely because our study focusses on a too narrow range of sizes and consequently too similar Reynolds numbers. Our CFD analysis confirms this, as the effect of Reynold number aerodynamic force production was negligible.

While only a handful studies have examined the effect of body mass on detailed wingbeat kinematics in insects (such as wingbeat amplitudes and angle-of-attack Belyaev et al., 2014; Lyu et al., 2019; Meresman and Ribak, 2017), the scaling between body mass and wingbeat frequency has been frequently assessed (Bartholomew and Casey, 1978; Byrne et al., 1988; Casey et al., 1985; De Nadai et al., 2021; Dudley, 1990; Tercel et al., 2018). In theory, an increase in wingbeat frequency is expected in smaller insects because flapping wing-based aerodynamic force production decreases faster than body mass. As such, it is puzzling that the reduction in size among the studied hoverfly species is not associated with increased wingbeat frequency.

Interestingly, inconsistent results on the scaling between body mass and wingbeat frequency have been found among other groups of flying animals. Although predominantly negative (e.g. in butterflies (Dudley, 1990), in orchid bees (Darveau et al., 2005), in mosquitoes (De Nadai et al., 2021), and in a comparative study across insect orders (Tercel et al., 2018)), several studies found weak or no relationship between body mass and wingbeat frequency (e.g. in moths (Bartholomew and Casey, 1978), whiteflies (Byrne et al., 1988), bees (Duell et al., 2022; Grula et al., 2021)). Consistent with our findings, a study comparing the wingbeat kinematics during tethered flight among hoverfly species spanning a range of body mass similar to that investigated here found no significant effect of body mass on wingbeat kinematics (Belyaev et al., 2014).

A possible explanation to the heterogeneous trends found among insect groups is that the extent to which body mass constrains the evolution of wingbeat kinematics depends on species ecology. Adaptation to a particular ecological niche may drive the evolution of specialised flight abilities, promoting a specific wingbeat kinematics. Selective pressures exerted on these wingbeat kinematics may sometimes overcome the constraints imposed by weight support. In that case, relatively constant wingbeat kinematics would be maintained over a range of body mass, and changes in wing morphology alone are expected to adjust for weight support. Consistent with this hypothesis is the decoupling of wingbeat kinematics from body mass variation, which has been highlighted in species where hovering flight ability is crucial for food acquisition. These include bees (Duell et al., 2022; Grula et al., 2021), sphingid moth (Bartholomew and Casey, 1978), hummingbirds (Skandalis et al., 2017), and hoverflies (Belyaev et al., 2014) as also in the present study. The absence of correlation between body mass and wingbeat kinematics observed in hoverflies may thus be due to their highly specialised hovering flight abilities.

The selective pressures promoting the evolution of unique hovering flight abilities in hoverflies are not precisely known but hovering steadily is likely advantageous for navigating between flowers when foraging for nectar. Good hovering flight performance may also have been promoted because it increases mating opportunities during the leck gathering of males (Downes, 1969; Gilbert, 1984; Heinrich and Pantle, 1975). Thus, selection for increased foraging efficacy and/or mating opportunities may have maintained consistent wingbeat kinematics across hoverfly species despite variation in body mass.

To further ascertain our findings, quantifying sexual dimorphism in hoverfly wingbeat kinematics would be a relevant future research direction. This would clarify the role of mating behaviour on the evolution of hovering flight performance. In this study, sexes could not be distinguished during our flight experiments, preventing us to test whether male hoverflies exhibit better hovering performance. Furthermore, additional wingbeat kinematics measurements in species from different habitats would help verifying the generality of our results in assessing the possible effect of habitat. Contrasted environmental conditions (e.g., temperature, humidity, or altitude) can affect flight performance (Altshuler and Dudley, 2003; Farnworth, 1972), and thus interfere with the scaling between flight and morphology. Future research should also focus on how the variation in scaling highlighted here influences peak locomotor performance. The effect of morphology on performance is generally most pronounced when individuals are pushed to their limits (Losos et al., 2002). For instance, investigating how the differential scaling between wing and body size affects peak thrust production during evasive flight manoeuvres would be a relevant experiment for future study.

Conclusion

Altogether, our results suggest that the hovering flight abilities of hoverflies may stem from highly specialised wingbeat kinematics that have been largely conserved throughout their diversification. In contrast, changes in wing morphology have evolved with the aerodynamic constraints associated with evolutionary changes in size. The existence of selective pressures maintaining stable wingbeat kinematics despite variation in body mass implies that high hovering flight performance can only be achieved over a limited range of wingbeat kinematic parameters, which remains to be ascertained.

Our study moreover highlights that unlike many other Diptera, hoverflies make use of a highly stereotyped inclined stroke-plane wingbeat kinematics for hovering, where during the forward dorsal-ventral wing stroke the wing moves not only from back to front but also downwards (Mou et al., 2011). This wingbeat kinematics pattern may contribute to the specialised hovering abilities of hoverflies, as it is consistently observed among the eight species studied here, with their stroke-plane pitch angle during hovering being approximately orientated at −30°. Furthermore, despite significant changes in wing morphology and wingbeat frequency among the studied species, the temporal dynamics of aerodynamic force production is strikingly similar between them (Figures 6C and S6). This suggests that not only the wingbeat kinematics is conserved between species, but also the underlying aerodynamics including the relative use of different unsteady aerodynamic mechanisms (Sane, 2003).

Thus, the apparent highly specialised flight style of hoverflies is maintained among several species with a large range of body masses, and robust against variations in wing shape and wingbeat frequency.

eLife Assessment

This important study addresses how wing morphology and kinematics change across hoverflies of different body sizes. The authors provide convincing evidence that there is no significant correlation between body size and wing kinematics across 28 species and instead argue that non-trivial changes in wing size and shape evolved to support flight across the size range. Overall, this paper illustrates the power and beauty of an integrative approach to animal biomechanics and will be of broad interest to biologists, physicists and engineers.

Reviewer #1 (Public review):

The paper is well written and the figures well laid out. The methods are easy to follow, and the rational and logic for each experiment easy to follow. The introduction sets the scene well, and the discussion is appropriate. The summary sentences throughout the text help the reader.

The authors have done a lot of work addressing my previous concerns and those of the other Reviewers.

Reviewer #2 (Public review):

Summary

Le Roy et al quantify wing morphology and wing kinematics across twenty eight and eight hoverfly species, respectively; the aim is to identify how weight support during hovering is ensured across body sizes. Wing shape and relative wing size vary non-trivially with body mass, but wing kinematics are reported to be size-invariant. On the basis of these results, it is concluded that weight support is achieved solely through size-specific variations in wing morphology, and that these changes enabled hoverflies to decrease in size. Adjusting wing morphology may be preferable compared to the alternative strategy of altering wing kinematics, because kinematics may be subject to stronger evolutionary and ecological constraints, dictated by the highly specialised flight and ecology of the hoverflies.

Strengths

The study deploys a vast array of challenging techniques, including flight experiments, morphometrics, phylogenetic analyses, and numerical simulations; it so illustrates both the power and beauty of an integrative approach to animal biomechanics. The question is well motivated, the methods appropriately designed, and the discussion elegantly places the results in broad biomechanical, ecological, and evolutionary context.

Weaknesses

(1) In assessing evolutionary allometry, it is key to pinpoint the variation expected from changes in size alone. The null hypothesis for wing morphology is well-defined (isometry), but the equivalent predictions for kinematic parameters, although specified, are insufficiently justified, and directly contradict classic scaling theory. A detailed justification of the "kinematic similarity" assumption, or a change in the null hypothesis, would substantially strengthen the paper, and clarify its evolutionary implications.

(2) By relating the aerodynamic output force to wing morphology and kinematics, it is concluded that smaller hoverflies will find it more challenging to support their body mass--a scaling argument that provides the framework for this work. This hypothesis appears to stand in direct contrast to classic scaling theory, where the gravitational force is thought to present a bigger challenge for larger animals, due to their disadvantageous surface-to-volume ratios. The same problem ought to occur in hoverflies, for wing kinematics must ultimately be the result of the energy injected by the flight engine: muscle. Much like in terrestrial animals, equivalent weight support in flying animals thus requires a positive allometry of muscle force output. In other words, if a large hoverfly is able to generate the wing kinematics that suffice to support body weight, an isometrically smaller hoverfly should be, too (but not vice versa). Clarifying the relation between the scaling of muscle mechanical input, wing kinematics, and weight support would help resolve the conflict between these two contrasting hypotheses, and considerably strengthen the biomechanical motivation and evolutionary interpretation.

(3) One main conclusion-- that miniaturization is enabled by changes in wing morphology--is insufficiently supported by the evidence. Is it miniaturization or "gigantism" that is enabled by (or drives) the non-trivial changes in wing morphology? To clarify this question, the isolated treatment of constraints on the musculoskeletal system vs the "flapping-wing based propulsion" system needs to be replaced by an integrated analysis: the propulsion of the wings, is, after all, due to muscle action. Revisiting the scaling predictions by assessing what the engine (muscle) can impart onto the system (wings) will clarify whether non-trivial adaptations in wing shape or kinematics are necessary for smaller or larger hovering insects (if at all!).

In many ways, this work provides a blueprint for work in evolutionary biomechanics; the breadth of both the methods and the discussion reflects outstanding scholarship.

Reviewer #3 (Public review):

This paper addresses an important question about how changes in wing morphology vs. wing kinematics change with body size across an important group of high-performance insects, the hoverflies. The biomechanics and morphology convincingly support the conclusions that there is no significant correlation between wing kinematics and size across the eight specific species analyzed in depth and that instead wing morphology changes allometrically. The morphological analysis is enhanced with phylogenetically appropriate tests across a larger data set incorporating museum specimens.

The authors have made very extensive revisions that have significantly improved the manuscript and brought the strength of conclusions in line with the excellent data. Most significantly, they have expanded their morphological analysis to include museum specimens and removed the conclusions about evolutionary drivers of miniaturization. As a result, the conclusion about morphological changes scaling with body size rather than kinematic properties is strongly supported and very nicely presented with a strong complementary set of data. I only have minor textual edits for them to consider.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer #1 (Public review):

The paper is well written and the figures well laid out. The methods are easy to follow, and the rational and logic for each experiment easy to follow. The introduction sets the scene well, and the discussion is appropriate. The summary sentences throughout the text help the reader.

The authors have done a lot of work addressing my previous concerns and those of the other Reviewers.

We are pleased that the revised manuscript satisfactorily addresses the previous concerns of the reviewer.

Reviewer #2 (Public review):

Summary

Le Roy et al quantify wing morphology and wing kinematics across twenty eight and eight hoverfly species, respectively; the aim is to identify how weight support during hovering is ensured across body sizes. Wing shape and relative wing size vary non-trivially with body mass, but wing kinematics are reported to be size-invariant. On the basis of these results, it is concluded that weight support is achieved solely through size-specific variations in wing morphology, and that these changes enabled hoverflies to decrease in size. Adjusting wing morphology may be preferable compared to the alternative strategy of altering wing kinematics, because kinematics may be subject to stronger evolutionary and ecological constraints, dictated by the highly specialised flight and ecology of the hoverflies.

Strengths

The study deploys a vast array of challenging techniques, including flight experiments, morphometrics, phylogenetic analyses, and numerical simulations; it so illustrates both the power and beauty of an integrative approach to animal biomechanics. The question is well motivated, the methods appropriately designed, and the discussion elegantly places the results in broad biomechanical, ecological, and evolutionary context.

We thank the reviewer for appreciating the strengths of our study.

Weaknesses

(1) In assessing evolutionary allometry, it is key to pinpoint the variation expected from changes in size alone. The null hypothesis for wing morphology is well-defined (isometry), but the equivalent predictions for kinematic parameters, although specified, are insufficiently justified, and directly contradict classic scaling theory. A detailed justification of the "kinematic similarity" assumption, or a change in the null hypothesis, would substantially strengthen the paper, and clarify its evolutionary implications.

We agree with the reviewer that a clearly articulated null hypothesis is crucial for interpreting scaling relationships. In fact, when carefully reviewing our manuscript, we realized that we nowhere did so, and which might have led to a misinterpretation of this. In the revised manuscript, we therefore now explicitly state our newly defined null hypotheses (lines 120–125, 340-352), and how we tested these (lines 359-360).

In fact, we define two alternative null hypotheses: (1) weight support is maintained across sizes using allometric scaling of wing morphology only, and thus wingbeat kinematics are kept constant (kinematic similarity); (2) weight support is maintained across sizes using allometric scaling of wingbeat kinematics, while wing morphology scales isometrically (morphological similarity).

According to the first null hypothesis, the second-moment-of-area of the wing should scale linearly with body mass, resulting in negative allometry of S2 relative to body mass (S2∼m1 <m4/3). According to the second null hypothesis, the product of wingbeat frequency and amplitude should scale with mass under negative allometry (ω∼ƒ Aϕ∼m-1/6). We test these alternative null hypotheses using Phylogenetic Generalized Least Square (PGLS) regressions of the morphology and kinematics metrics against the body mass.

Furthermore, in our revised manuscript, we now also better explain the use of "kinematic similarity" assumption as a theoretical scenario, that is physically, biomechanically nor physiological sustainable across sizes, but that we merely use to define our null hypotheses (lines 340-351). This is made particularly explicit in a new subsection named “Theoretical considerations” (lines 448–461). Note that our second null hypothesis is thus not that hoverflies fly under "kinematic similarity", but that wingbeat kinematics scales under negative allometry (ω∼ƒ Aϕ∼m-1/6), which we assume is in line with the classic scaling theory that the reviewer refers to.

We sincerely thank the reviewer for making us aware that we did not explicitly state our null hypotheses, and that introducing these new null hypotheses removed the confusion about the assumptions in our study.

(2) By relating the aerodynamic output force to wing morphology and kinematics, it is concluded that smaller hoverflies will find it more challenging to support their body mass--a scaling argument that provides the framework for this work. This hypothesis appears to stand in direct contrast to classic scaling theory, where the gravitational force is thought to present a bigger challenge for larger animals, due to their disadvantageous surface-to-volume ratios. The same problem ought to occur in hoverflies, for wing kinematics must ultimately be the result of the energy injected by the flight engine: muscle. Much like in terrestrial animals, equivalent weight support in flying animals thus requires a positive allometry of muscle force output. In other words, if a large hoverfly is able to generate the wing kinematics that suffice to support body weight, an isometrically smaller hoverfly should be, too (but not vice versa). Clarifying the relation between the scaling of muscle mechanical input, wing kinematics, and weight support would help resolve the conflict between these two contrasting hypotheses, and considerably strengthen the biomechanical motivation and evolutionary interpretation.

We agree with the reviewer that, due to disadvantageous surface-to-volume ratios, larger animals are more challenged to maintain weight-support, and that this is also the case for hovering hoverflies. In the current manuscript, we do not aim to challenge this universal scaling law of muscle force with body mass.

Instead, we here focus merely on how the flight propulsion system (wing morphology and kinematics) scale with size, and how this allows hovering hoverflies to maintain weight support. We also fully agree with the reviewer that in theory, “if a large hoverfly is able to generate the wing kinematics that suffice to support body weight, an isometrically smaller hoverfly should be, too”. This aligns in fact with our second null hypothesis where wingbeat frequency should scale as ƒ∼m-1/6, to maintain weight support under morphological isometry.

In our study, we show that this null hypothesis is rejected (lines 511-517, and line 525), and thus hoverflies primarily adjust their wing morphology to maintain in-hovering weight-support across sizes, and wingbeat kinematics is in fact highly conserved. Why this specific flight kinematics is so strongly conserved is not known, and thus a key topic in the discussion section of our manuscript.

We agree with the reviewer that muscle physiology might be an important driver for this conserved kinematics, but also aerodynamic efficiency and maneuverability could be key aspects here. In our revised manuscript, we now discuss these three aspects in more detail (lines 762-775). Also, we here now also mention that we aim to address this outstanding question in future studies, by including muscle physiology in our animal flight studies, and by studying the aerodynamics and maneuver kinematic of hoverflies in more detail.

Moreover, in our revised introduction section, we now also mention explicitly that the capability for maintaining in-flight weight-support scales inversely with animal size, due to the negative isometric scaling of muscle force with body mass (line 52-56). Furthermore, we removed all statements that might suggest the opposite. We hope that these adjustments helped resolve the apparent conflict between our null hypotheses and general muscle scaling laws.

Finally, in the Discussion section (lines 770-775), we now more explicitly acknowledge that wing motion is ultimately driven by the flight motor musculature, and that a full biomechanical interpretation must consider the scaling of muscle mechanical input alongside wing kinematics and morphology. While we decided to keep the focus primarily on aerodynamic constraints in this study, we agree that future work integrating both aerodynamic and physiological scaling will be essential to fully resolve these contrasting perspectives.

(3) One main conclusion-- that miniaturization is enabled by changes in wing morphology--is insufficiently supported by the evidence. Is it miniaturization or "gigantism" that is enabled by (or drives) the non-trivial changes in wing morphology? To clarify this question, the isolated treatment of constraints on the musculoskeletal system vs the "flapping-wing based propulsion" system needs to be replaced by an integrated analysis: the propulsion of the wings, is, after all, due to muscle action. Revisiting the scaling predictions by assessing what the engine (muscle) can impart onto the system (wings) will clarify whether non-trivial adaptations in wing shape or kinematics are necessary for smaller or larger hovering insects (if at all!).

In many ways, this work provides a blueprint for work in evolutionary biomechanics; the breadth of both the methods and the discussion reflects outstanding scholarship.

In response to the first review round, we have removed all references to “miniaturization,” as our data does not allow us to infer evolutionary trajectories of body size (i.e., whether lineages have become smaller or larger over time). We now frame our conclusion more conservatively: that changes in wing morphology enable small hoverflies to maintain weight support despite the aerodynamic disadvantages imposed by isometric scaling.

We fully agree that an integrated biomechanical framework, explicitly linking muscle mechanical output with wing kinematics and morphology, would significantly strengthen the study. However, we believe that performing an integrated analysis assessing the scaling of muscle input into the wing is beyond the current scope, which focuses specifically on the aerodynamic consequences of morphological and kinematic variation (see reply above).

Reviewer #3 (Public review):

This paper addresses an important question about how changes in wing morphology vs. wing kinematics change with body size across an important group of high-performance insects, the hoverflies. The biomechanics and morphology convincingly support the conclusions that there is no significant correlation between wing kinematics and size across the eight specific species analyzed in depth and that instead wing morphology changes allometrically. The morphological analysis is enhanced with phylogenetically appropriate tests across a larger data set incorporating museum specimens.

The authors have made very extensive revisions that have significantly improved the manuscript and brought the strength of conclusions in line with the excellent data. Most significantly, they have expanded their morphological analysis to include museum specimens and removed the conclusions about evolutionary drivers of miniaturization. As a result, the conclusion about morphological changes scaling with body size rather than kinematic properties is strongly supported and very nicely presented with a strong complementary set of data. I only have minor textual edits for them to consider.

We thank the reviewer for this positive feedback. We are pleased to hear that the revised manuscript is satisfactory.

Reviewer #2 (Recommendations For The Authors):

My main remaining qualm remains the null hypothesis for the scaling of kinematic parameters - all weaknesses come back to this point. I appreciate that the authors now specify an expectation, but they offer no justification. This is a problem, because the expectation dictates the interpretation of the results and is thus crucial to some of the key claims (including one in the paper title!): the choice made by the authors indeed implies that hovering is harder for small hoverflies, so that the reported changes in size-specific wing morphology are to be interpreted as an adaptation that enables miniaturization. However, why is this choice appropriate over alternatives that would predict the exact opposite, namely that hovering is harder for larger hoverflies?

In my original review, I suggested that the authors may address this key question by considering the scaling of muscle mechanical output, and provided a quick sketch of what such an argument would look like, both in classic textbook scaling theory, and in the framework of more recent alternative approaches. The authors have decided against an implementation of this suggestion, providing various version of the following justification in their reply: "our study focuses precisely on this constraint on the wing-based propulsion system, and not on the muscular motor system." I am puzzled by this distinction, which also appears in the paper: muscle is the engine responsible for wing propulsion. How can one be assessed independent of the other? The fact that the two must be linked goes straight to the heart of the difficulty in determining the null hypotheses for the allometry of kinematic and dynamic parameters: they must come from assertions on how muscle mechanical output is expected to vary with size, and so couple muscle mechanical output to the geometry of the wing-based propulsion system. What if not muscle output dictates wing kinematics?

I fully agree with the authors that null hypotheses on kinematic parameters are debatable. But then the authors should debate their choice, and at least assess the plausibility of its implications (note that the idea of "similarity" in scaling does not translate to equal or invariant, but is tied closely to dimensional analysis - so one cannot just proclaim that kinematic similarity implies no change in kinematic parameters). I briefly return to the same line of argument I laid out in the initial review to provide such an assessment:

Conservation of energy implies:

W = 1/2 I ω2

where I is the mass moment of inertia and W is the muscle work output. Under isometry, I ∝m5/3, the authors posit ω ∝m0, and it follows at once that they predict W ∝m5/3. That is, the "kinematic similarity" hypothesis presented in the paper implies that larger animals can do substantially more work per unit body mass than small animals (unless the author have an argument why wing angular velocity is independent of muscle work capacity, and I cannot think of one). This increase in work output is in contradiction with the textbook prediction, going all the way back to Borelli and Hill: isogeometric and isophysiological animals ought to have a constant mass-specific work output. So why, according to the authors, is this an incorrect expectation, ie how do they justify the assumption ω ∝m0 and its implication W ∝m5/3? How can larger animals do more mass-specific work, or, equivalently, what stops smaller animals from delivering the same mass-specific work? If non-trivial adaptations such as larger relative muscle mass enable larger animals to do more work, how does this fit within the interpretation suggested by the authors that the aerodynamics of hovering require changes in small animals?

A justification of the kinematic similarity hypothesis, alongside answers to the above questions, is necessary, not only to establish a relation to classic scaling theory, but also because a key claim of the paper hinges on the assumed scaling relationship: that changes in wing morphology enable hovering in small hoverflies. If I were to believe Borelli, Hill and virtually all biomechanics textbooks, the opposite should be the case: combing constant mass-specific work output with eq. 1, one retrieves F∝m2/3, so that weight support presents a bigger challenge for larger animals; the allometry of wing morphology should then be seen as an adaptation that enables hovering in larger hoverflies - the exact opposite of the interpretation offered by the authors.

Now, as it so happens, I disagree with classic scaling theory on this point, and instead believe that there are good reasons to assume that muscle work output varies non-trivially with size. The authors can find a summary of the argument for this disagreement in the initial review, or in any of the following references:

Labonte, D. A theory of physiological similarity for muscle-driven motion. PNAS, 2023, 120, e2221217120

Labonte, D.; Bishop, P.; Dick, T. & Clemente, C. J. Dynamics similarity and the peculiar allometry of maximum running speed. Nat Comms., 2024, 15, 2181

Labonte, D. & Holt, N. Beyond power limits: the kinetic energy capacity of skeletal muscle. J Exp Bio, 2024, 227, jeb247150

Polet, D. & Labonte, D. Optimal gearing of musculoskeletal systems. Integr Org Biol, 2024, 64, 987-10062024

I am asking neither that the authors agree with the above references nor that they cite them. But I do expect that they critically discuss and justify their definition of kinematic similarity, its relation to expectation from classic scaling theory, and the implications for their claim that hovering is harder for small animals. I do note that the notion of "physiological similarity" introduced in the above references predicts a size-invariant angular velocity for small animals, that small animals should be able to do less mass-specific work, and that average muscle force output can grow with positive allometry even for isogeometric systems. These predictions appear to be consistent with the data presented by the authors.

We agree with the reviewer that our null hypothesis was not clearly articulated in our previous version of the manuscript, and that this might have led to a misinterpretation of the merits and limitations of our study. In the revised manuscript, we therefore now explicitly introduce our null hypotheses in the Introduction (lines 120–125), we define these in the Methods section (lines 340–360), test these in the Results section (lines 511–517), and reflect on the results in the Discussion (lines 602–610). We thank the reviewer for pointing out this unclarity in our manuscript, because revising it clarified the study significantly. See our replies in the “Public Review” section for details.

Minor points

L56: This is somewhat incomplete and simplistic; to just give one alternative option, weight support with equivalent muscle effort could also be ensured by a change in gearing (see eg Biewener's work). It is doubtful whether weight support is a strong selective force, as any animal that can move will be able to support its weight. The impact of scaling on dynamics is thus arguably more relevant.

We thank the reviewer for pointing out that our original sentence may be too simplistic. We now briefly mention alternative mechanisms (suggested by the reviewer) to provide more nuance (line 56-58).

L58: I am not aware of any evidence that smaller animals have reduced the musculature dedicated to locomotion beyond what is expected from isometry; please provide a reference for this claim or remove it.

We removed that claim.

The authors use both isometry and geometric similarity. As they also talk about muscle, solely geometric similarity (or isogeometry) may be preferable, to avoid confusion with isometric muscle contractions.

To avoid confusion, we now use “geometric similarity” wherever the use of isometry might be ambiguous.

L86: negative allometry only makes sense if there is a justified expectation for isometry - I suggest to change to "The assumed increase in wingbeat frequency in smaller animals" or similar, or to clarify the kinematic similarity hypothesis.

We edited the sentence as suggested.

L320: This assertion is somewhat misleading. Musculoskeletal systems are unlikely to be selected for static weight support. Instead, they need to allow movement. Where movement is possible, weight support is trivially possible, and so weight support should rarely, if ever, be a relevant constraint. At most, the negative consequence of isometry on weight support would be that a larger fraction of the muscle mass needs to be active in larger animals to support the weight.

We fully agree with the reviewer that musculoskeletal systems are unlikely not selected for static loads, as the ability to move dynamically in the real world is crucial for survival. That said, we here look at hovering flight, which is far from static. In fact, hovering flight is among the energetic most costly movement patterns found in nature, due to the required high-frequency wingbeat motions (Dudley 2002). Rapid maneuvers are of course more power demanding, but hovering is a good proxy for this. For example, in fruit flies maximum force production in rapid evasive maneuvers are only two times the force produced during hovering (Muijres et al., 2014).

We agree with the reviewer that it is important to explicitly mention the differences in functional demands on the motor system in hovering and maneuvering flight, and thus we now do so in both the introduction and discussion sections (lines 116-118 and 762-765, respectively).

Dudley, Robert. The biomechanics of insect flight: form, function, evolution. Princeton university press, 2002.Muijres, F. T., et al. "Flies evade looming targets by executing rapid visually directed banked turns." Science 344.6180 (2014): 172-177.

Reviewer #3 (Recommendations For The Authors):

Throughout, check use of "constrains" vs. "constraints"

Thank you for pointing this out. We have corrected these errors.

Line 52 do you mean lift instead of thrust?

We agree with the reviewer that the use of “thrust” might be confusing in the context of hovering flight, and thus we replaced “flapping-wing-based aerodynamic thrust-producing system” with the “flapping-wing-based propulsion system”. This way, we no longer use the word thrust in this context, and only use lift as the upward-directed force required for weight-support.

Line 60 "face also constrains" wording

Corrected.

Line 79 Viscous forces only "dominate" at Re<1 and so this statement only refers to very very small insects which I suspect are far below the scale of the hoverflies considered (likely Re ~100) although maybe not for the smallest 3 mg ones?

Indeed, viscous forces do not “dominate” force production at the Reynolds numbers of our flying insects. We thank the reviewer for pointing out this incorrect statement, which we corrected in the revised manuscript.

Line 85 again thrust doesn't seem to be right

Agreed. See reply 3.2.

533 "maximized" should probably be "increased"

We now use “increased”.

Line 705-710 The new study by Darveau might help resolve this a bit because of the reliability of this relationship across and between orders. Darveau, C.-A. (2024). Insect Flight Energetics And the Evolution of Size, Form, And Function. Integrative And Comparative Biology icae028.

We thank the reviewer for this highly relevant reference, which was unfortunately not included in the original manuscript. In connection with this work, we now further discuss the relationship between wing size allometry and deviations from the expected scaling of wingbeat frequency (lines 730-735).

Version 3

Abstract

Abstract

Due to physical scaling laws, size greatly affects animal locomotor ability and performance. Whether morphological and kinematic traits always jointly respond to size variation is however poorly known. Here, we examine the relative importance of morphological and kinematic changes in mitigating the consequence of size reduction on aerodynamic force production for weight support in flying insects, focusing on hovering flight of hoverflies (Syrphidae). We compared the morphology of 28 hoverfly species, and the flight biomechanics and aerodynamics of eight species with body masses ranging from 5 to 100 mg. Our study reveals no significant effect of body mass on wingbeat kinematics among species, suggesting that morphological rather than kinematic changes may compensate for the reduction in weight support associated with an isometric reduction in wing size. Computational Fluid Dynamics simulations confirmed that adaptations in wing morphology drive the ability of small hoverfly species to generate weight support, although variations in wingbeat kinematics between species cannot be completely neglected. We specifically show that smaller hoverflies have evolved relatively larger wings, and aerodynamically more effective wing shapes, to mitigate the reduction in aerodynamic weight support with isometric size reduction. Altogether, these results suggest that hoverfly flight underpins highly specialised wingbeat kinematics, which have been largely conserved throughout evolution; instead, evolutionary adaptations in wing morphology enabled flight of small hoverflies.

Introduction

Evolutionary changes in size are common over the course of animal diversification (Clauset and Erwin, 2008; Hanken and Wake, 1993), strongly impacting many aspects of morphology, physiology, and behaviour. Investigating how changes in body size affect other traits (i.e., scaling) is thus essential to understand the causes of traits variation in animals. Traditionally rooted in the field of morphology and physiology (Gould, 1966; Hill, 1950; Pélabon et al., 2014), allometric studies have broadened to explore how more complex traits scale with animal size. These include locomotor kinematics (e.g. flight, stride or swimming kinematics, Bejan and Marden, 2006; Riskin et al., 2010), and behaviour (e.g. mating or escape tactics, Dial et al., 2008). Although behavioural or biomechanical traits are often more noisy and challenging to measure, their inclusion in allometric studies is crucial to better understand trait evolution (Cloyed et al., 2021). Indeed, most phenotypes involve both morphological and behavioural traits, each responding in potentially different ways to size variation (Clemente and Dick, 2023). Difference in scaling patterns between body size and other traits is generally most apparent between species, and can be partly explained by neutral divergence during phylogenetic history (Blomberg et al., 2003). Understanding the origin of trait variation among species thus requires controlling for the influence of phylogeny on allometric relationships (Labonte et al., 2016; Peattie and Full, 2007).

In this study, we address the consequences of size variation between species (i.e. evolutionary allometry) and specifically question the relative importance of wing morphology versus wingbeat kinematic changes in mitigating the effect of size on flight ability.

Flight typically combines a suite of morphological and biomechanical traits, providing a relevant context to jointly examine how morphology and locomotor kinematics change with size. Moreover, the physical scaling laws governing flight allow drawing predictions on the scaling of morphology and kinematics in flying animals. These physical laws set both upper and lower bounds on the size of flying animals, resulting from the need to maintain weight-support by producing upward-directed aerodynamic lift forces. In this context, there are two primary limitations for producing these forces: Limitations on (1) the muscular flight motor system, and on (2) the flapping-wing-based propulsion system.

Just like terrestrial animals, larger flying animals face increasing constraints on muscle force production as they grow. This limitation arises because muscle force output scales with muscle cross-sectional area, and therefore under geometric similarity it increases at a lower rate than body mass. Consequently, as body size increases, producing the muscle forces for maintaining weight support may require a positive allometry of muscle cross-sectional area (Alexander, 2013); however, alternative strategies such as changes in musculoskeletal gearing or adjustments in flight kinematics and behaviour can also help offset these scaling constraints (Azizi et al., 2008; Norberg, 2007). Conversely, smaller animals are not constrained by muscle force production but instead benefit from a favourable surface-to-volume ratio.

Next to this constraint on the muscular motor system, flying animals also face constraints stemming from the scaling between their flapping-wing-based propulsion system and body mass. For large flying animals, the aerodynamic lift (L) for weight support scales linearly with the surface area of the wing (S), and quadratically with the flight speed (U∞) as L∼S U∞2 (Norberg, 2007). Under geometric similarity, wing surface area scales with body mass (m) as S∼m2/3 (Schmidt-Nielsen, 1975). Thus, to maintain weight support during flight (L=mg, where g is gravitational acceleration), larger flying animals exhibit disproportionately larger wings and tend to fly faster (Norberg, 2012). This positive allometry in wing surface area and flight speed with body mass is needed to generate enough lift to stay in the air.

Smaller flying animals such as insects generally fly slower, and thus air movements over the wing are governed by wing flapping and less so by their forward movement. In the limiting case of hovering flight (U∞=0 m/s), the airflow is generated entirely by the flapping wings, and the animal relies solely on this flapping-based propulsion to produce an upward aerodynamic force that balances its weight. This aerodynamic lift forces produced by flapping wings scale linearly with the wing’s second-moment-of-area (S2) and quadratically with angular wing speed (ω) as L∼S2 ω2 (Weis-Fogh, 1973). Thus, for small slow flying animals, and particularly during hovering flight, the lift forces required for weight support scale primarily with the wing’s second-moment-of-area, relative to the wing hinge (Muijres et al., 2017). For geometric similarity, this second-moment-of-wing-area scales with body mass as S2∼m4/3. Because this scaling factor of 4/3 is larger than one, a wing’s second-moment-of-area decreases at a higher rate than body mass, and therefore wingbeat-induced aerodynamic lift reduces more rapidly than body weight. Thus, to maintain weight support (mg=L∼S2 ω2), smaller flying animals require disproportionately larger wings, and/or they need to beat their wings at higher wingbeat frequencies or amplitudes, resulting in higher angular speeds of the beating wing. Notably, decreasing size also leads to a drop in Reynolds number, increasing the relative influence of viscous over inertial forces. Consequently, at the low Reynolds numbers typical of small insects, viscous forces may strongly affect both lift and drag forces produced during flapping flight (Sane, 2003).

In this study, we aim to test how physical scaling laws cause constraints on the flapping-wing-based propulsion system in small insects, and how this drives evolutionary adaptations in wing morphology and wingbeat kinematics. Here, we hypothesize that smaller insects have evolved disproportionately larger wings, and beat their wings at higher wingbeat frequencies, allowing them to maintain weight support despite the detrimental scaling effects of size reduction on their propulsion system.

The assumed increase in wingbeat frequency as size decreases is supported by the consistent negative relationship between wingbeat frequency and size observed in most flying animals (Norberg, 2007; Rayner, 1988; Riskin et al., 2010; Tercel et al., 2018). A general trend toward disproportionately larger wings in smaller species is however less striking among species (but see Danforth, 1989; Ellington, 1984a; Outomuro et al., 2013), suggesting that morphological changes in the propulsion system in response to size variation are not always straightforward.

Although empirical data often align with expectations drawn from physical laws, the scaling of morphology and kinematic with size can sometimes be challenging to predict. Selective pressures associated with ecological specialisation may influence the evolution of morphology and kinematic, and conflict with the constraints imposed by physical scaling laws. Each component may then adjust for the constraints imposed by weight support in varying degree. For example, changes in morphology have been shown to predominantly mitigate the negative consequences of body size reduction: among hummingbirds, a disproportionate increase in wing area – but little change in wingbeat kinematics – enable weight support among species of different sizes (Skandalis et al., 2017). Similarly, a positive allometry in wing area, but no significant change in wingbeat frequency, was found in relation to body size among species of bees (Duell et al., 2022; Grula et al., 2021). Such discrepancies with the general scaling expectation may point at a particular selective regime maintaining specialised flight kinematics and ability. Here we focused on an ecologically specialised insect group, the hoverflies, to test if the consequences of size variation are mostly mitigated by changes in wingbeat kinematics or in wing morphology. Hoverflies (Diptera: Syrphidae) are among the most species-rich groups of nectar-foraging insects and exhibit striking variation in size, ranging from 0.5 mg to over 200 mg (Katzourakis et al., 2001). Despite these large size differences, all adult hoverflies show exceptional hovering abilities. This highly specialised flight behaviour may have been promoted because it plays an important role in mating behaviour: male hoverflies aggregate at specific locations such as under a tree and hover steadily, waiting for passing females to pursue (Gilbert, 1984; Heinrich and Pantle, 1975). Hovering abilities may thus be correlated with mating success (Downes, 1969; Gilbert, 1984). It may also have been favoured in the context of foraging and navigating around flowers, albeit hovering may not directly relate to feeding as most species land on flowers to feed (Gilbert, 1981). Selection acting on the flight of hoverflies may interfere with general scaling expectations between body mass, wing morphology and wingbeat kinematics.

Here, we study the evolutionary response of wing morphology and wingbeat kinematics to weight support constraints in hovering hoverflies, specifically for small species. We focus on hovering flight as it is well-defined (U∞=0 m/s), and because it is a functionally and energetically demanding flight mode (Ellington, 1991). During such hovering, the aerodynamic forces produced by the beating-wing-based propulsion system scales negative with mass. We therefore hypothesize that tiny hoverflies have evolved relatively large wings or high wingbeat frequencies to maintain weight support during hovering. To test whether the potential adaptations in morphology or kinematics are dominant, we furthermore define two alternative null hypotheses: (1) by assuming that selection for strong hovering performance is limiting kinematic flexibility, we hypothesize that adaptations in wing morphology prevail over adaptations in wingbeat kinematics. (2) Alternatively, if flight behaviour is evolutionarily more labile than morphology (Blomberg et al., 2003), we expect that adaptations in wingbeat kinematics prevail over wing morphological adaptations.

We test these two null hypotheses by examining the scaling relationships between wing morphology and body mass in 28 hoverfly species ranging from 3 to 132 mg, and conjointly assessing how wing morphology and wingbeat kinematics scale with body mass for eight species ranging from 5 to 100 mg. We then combined computational fluid dynamic simulations with quasi-steady aerodynamic modelling to estimate aerodynamic forces produced in hovering flight. Based on this, we determined how the relative changes in wingbeat kinematics and wing morphology contribute to producing weight support during hovering flight of hoverflies.

Material and methods

Collecting and identifying hoverflies

We quantified body and wing morphology in 28 hoverfly species, of which we collected eight species in the wild and twenty from museum specimens at the Naturalis Biodiversity centre (Leiden, the Netherlands). The wild caught live hoverflies were also used for flight experiments, to quantity their wingbeat kinematics during hovering flight (Figure 1).

The studied hoverfly species shown with their phylogenetic relationships.

Sample size as number of individuals for each species is in parentheses. Species marked with coloured circles are displayed on the right, and coloured circles with black outlines are of species for which both morphology and flight kinematics were quantified. Body mass is presented as mean values and standard deviation. Phylogeny was obtained from Wong et al. (2023).

From the museum collection, we sampled 74 individual specimens, aiming to include two males and two females per species whenever possible. This resulted in 4.2±1.7 individuals per species (mean±standard deviation). Museum specimens were selected to maximize both phylogenetic coverage and size range. The eight hoverflies from the wild were collected in September 2022 on the campus of Wageningen University, the Netherlands (51°59’01.0”N 5°39’32.7”E; ca. 10 m). We captured a total 44 individuals, resulting in a sample size of 5.5±2.9 individuals per species. Flies were captured with hand-held nets and brought in the lab to record their flight within an hour after capture, after which morphology was quantified.

The wild species were first distinguished visually, but their species identity was subsequently ascertained using DNA barcoding. This was done by sending a piece of leg of each individual to the Canadian Centre for DNA Barcoding (CCDB), where barcoding was performed following standard automated protocols of the BOLD Identification System (http://www.boldsystems.org/) (Ratnasingham and Hebert, 2007). Sequenced DNA was then compared with a reference library to establish a species match. Sexes were determined visually using criteria from Gilbert (2015).

Flight experiments

The flight experiments were carried out in a custom-built octagonal flight arena made of transparent Plexiglas (50×50×48 cm, height×width×length; Figure 2A) (Cribellier et al., 2022). Stereoscopic high-speed videos were recorded using three synchronised high-speed cameras (Photron FASTCAM SA-X2), equipped with Nikon Sigma 135 mm lenses, and recording at a temporal resolution of 5000 frames s−1, a spatial resolution of 1084×1084 pixels, and a shutter speed of 1/10000 s. The cameras were positioned around the arena to provide a top view, and inclined left and right-side view of the flying insects, which resulted in a zone of focus of approximately 12 cm3 in the centre of the arena in which body and wing movement were in view and in focus of all three cameras. To increase contrast and therefore facilitate subsequent tracking, the cameras were back-lit using three infrared light panels, placed behind the bottom and lower side walls of the octagon arena, opposite each camera (Figure 2A).

Quantifying the in-flight wingbeat kinematics and wing morphology of hoverflies.

(A) Hoverflies were released in an octagon-shaped flight arena. We recorded stereoscopic high-speed videos of the flying hoverflies using three synchronised high-speed video cameras; from the videos we reconstructed the three-dimensional body and wingbeat kinematics. Infrared light panels positioned as the bottom of the arena enabled high contrast between the flying insect and the background. (B) Conventional wing angles were measured at each time step (t=0.4 ms) in the body reference frame. ϕ, wing stroke angle within the stroke plane; η, wing deviation angle out of the stroke plane; θ, wing rotation angle along the spanwise axis; , air velocity vector relative to the wing; α, angle-of-attack of the wing. (C) Wing morphological parameters including their definitions: wingspan R, wing surface area S, radial position along the span r, local wing chord c at distance r, the second-moment-of-area S2 and its dimensionless homologue .

Prior to filming, the stereoscopic camera system was calibrated with the direct linear transformation (DLT) technique (Hartley and Sturm, 1995), by digitising a wand of 2.8 cm long moved throughout the zone of focus of the cameras. To track the wand movement, we trained an artificial neural network using the pose estimation Python library DeepLabCut (Nath et al., 2019), achieving a test error of 3.1 pixels. Computation of the DLT coefficients was then done using the MATLAB program easyWand (Theriault et al., 2014). The calibration root mean square error was 1.21 pixels.

During the flight experiment, all the individuals from a single species were released together in the arena to increase the probability that an individual flew through the intersecting fields of view of the three cameras. Most individuals tended to remain on the sidewalls, occasionally making short flights across the arena. For each species, the filming experiment was carried out until a minimum of five suitable flight sequences were obtained, whenever possible. A suitable sequence was defined as a flying individual clearly crossing the intersecting fields of view. Although this procedure facilitated the recording, this prevented identifying which specific individual or sex was recorded flying, leaving open the possibility of recording the same individual multiple times. Note that this approach aimed at estimating the average wingbeat kinematics per species as we could not directly assign flight measurements at the individual level.

Wingbeats selection

First, we determined the flight speed throughout each recorded flight sequence. We did this by tracking the hoverfly body centre, estimated from the mean position between the head and abdomen extremities, throughout the flight sequence (mean duration of the tracked trajectories = 0.13±0.11 sec) using a trained DeepLabCut neural network (Nath et al., 2019) (test error: 6.3 pixels). From this, we calculated flight speed using a central temporal differentiation scheme. Based on these speed data, we selected flight sequences with the lowest average speed and, within each sequence, identified the wingbeat occurring during the slowest part of the trajectory. For this selection process, we only included the flight trajectory sections in which the hoverfly was flying straight at a low climb angle, to prevent possible variation in wingbeat kinematics stemming from flight manoeuvres. We digitised a minimum of three wingbeats per species, each taken from a different flight sequence. For six of the species we tracked three wingbeats, and for two species (Melanostoma mellinum and Sphaerophoria scripta) we tracked six wingbeats. Note that our experiment did not allow capturing perfect hovering flight sequences (i.e. flight speed = 0 m.s−1) but aimed at getting as close as possible to this flight mode. This was evaluated using the advance ratio (see next section).

Quantifying wingbeat kinematics

We quantified the wing movement using the manual stereoscopic video tracker Kine in MATLAB (Fontaine et al., 2009; Fry et al., 2003) originally designed to track the wingbeat kinematics of fruit flies. The tracker was tailored to hoverflies by building wing models matching the wings of the studied hoverfly species. Wing models were obtained by digitising the wing outline on high-resolution photographs and consisted of 41 2D coordinates representing the wing shape, with the known hinge and tip positions. To reduce complexity, wings were considered as rigid flat plates, although real hoverflies wings endure deformations during the wing stroke (Walker et al., 2010).

We tracked the wingbeat kinematics on down-sampled videos (from 5000 to 2500 frames s−1), as this preserved high enough temporal resolution to capture the wing movement (n=27±5 frames per wingbeat). The tracking provided us with the position and orientation of the body and wings in each consecutive video frame. The position and orientation of the body were defined in the world reference frame, as a 3D position vector (X=[x,y,z]) and the consecutive Euler angles body yaw, pitch and roll. The angular orientation of the wings was expressed in the hoverfly body reference frame (Muijres et al., 2014) (Figure 2B), as Euler angles relative to the stroke plane: the stroke angle (ϕ), deviation angle (η) and rotation angle (θ) of each wing (Figure 2B). The stroke plane of each wing was defined as the plane at a 45° angle to the long axis of the body passing through the hinge position of that wing. The three wing kinematics angles (stroke, deviation and rotation angle) were measured separately on the left and right wing and then averaged between the two wings. The time history of the averaged angles was then fitted with a 4th order Fourier series for all three angles (Muijres et al., 2014), and their temporal derivatives were computed (i.e. stroke, deviation and rotation rate). We estimated the absolute angular speed of the beating wing as where and are the stroke rate and deviation rate, respectively. The angle of attack (α) was computed as the angle between the wing plane and the velocity vector of the wing (Figure 2B). Based on the temporal dynamic of the wing kinematics angles, we derived the following wingbeat-average kinematics parameters: wingbeat frequency, defined as f=1/Twingbeat, where Twingbeat is the duration of the wingbeat; wing stroke, deviation, and rotation amplitudes, defined as Aϕ=|ϕmax −ϕmin|, Aη=|ηmax −ηmin|, and Aθ =|θmax −θmin|, respectively; we estimated the wingbeat-average angular speed as ; we estimated the mean angle-of-attack as the mean value at mid forward and backward wing stroke, where aerodynamic force production is close to maximum (Tian et al., 2013).

We quantified the wingbeat-average body kinematics during each wingbeat using the mean flight speed, climb angle, and the pitch angle of both body and stroke-plane. Flight speed U was estimated as the temporal derivative of body positions, and climb angle as γclimb=atan(Uver/Uhor), where Uver and Uhor are the vertical and horizontal component of the flight velocity. Body pitch angle βbody was calculated as the angle between the body axis and the horizontal. The equivalent stroke-plane pitch angle was defined as the pitch angle of the stroke plane relative to the horizontal, calculated as βstroke-plane=βbody–45°.

To verify whether the studied wingbeats were representative of the overall recorded flight behaviour, we calculated the mean wingbeat frequency over the full flight sequence as , where nwingbeats and Tsequence are the number of wingbeats in a sequence and the sequence duration, respectively. This mean wingbeat frequency was then compared with that of the studied wingbeats. To estimate whether the analysed wingbeats were representative for hovering flight we calculated the advance ratio of all wingbeats as , where R is the span of a single wing. Hovering is generally defined as flight with advance ratios J<0.1, i.e. when the forward flight speed is less than 10% of the wingbeat-induced speed of the wingtip (Ellington, 1984a).

Finally, although hoverfly flight speed was minimised in the wingbeat selection procedure, remaining variation in body kinematics may still affect the wingbeat kinematics and therefore confound with changes in wingbeat kinematics resulting from scaling effect. We therefore assessed if the flight kinematics metrics flight speed and climb angle affected wingbeat kinematics by performing multiple regression analyses. Here, each wingbeat kinematic metric was treated as dependent variable and flight speed and climb angle as model effects.

Quantifying morphology

We quantified body and wing morphology in both the wild-caught hoverflies and the museum specimens. After being filmed flying, wild-caught individuals were sacrificed by freezing them at −20°C, and weighed with a resolution of 0.1 mg using an analytical balance (XSR204 Mettler Toledo). Wings were then detached from the body and photographed using a digital microscope (Zeiss Stemi SV11). From the museum collection, we only selected specimens with properly flattened (i.e., non-folded) wings, and photographed them using a Nikon D600 camera equipped with a 60mm lens. Here, we photographed the complete animal from above, and the separate wing by it placing parallel to the camera lens. We estimated body mass of museum specimens using the correlation between thorax width and the fresh mass in the wild caught specimens (Figure S1).

From the wing photographs of both the wild and museum specimens, we measured the primary wing morphology parameters using WingImageProcessor in MATLAB (available at https://biomech.web.unc.edu/wing-image-analysis/). These include single-wing area S, wingspan R, mean wing chord , and second-moment-of-area S2. From these we calculated weight-normalised wingspan as R*=R/m1/3, and the normalised second-moment-of-area as . The weight-normalized wingspan is a parameter describing relative wingspan, and the normalized second-moment-of-area is defining wing shape changes independent of changes in wing size. Both parameters allowed us to identify variations in relative wingspan and relative wing shape between species of different sizes, respectively.

Next, we used a landmark-based geometric morphometric method to quantify more precisely the variation in wing shape between species (Bookstein, 1997). Wing shape outline was determined using 300 semi-landmarks equidistantly spaced along the wing outline. Semi-landmarks are commonly used to describe curvy shapes such as wing outlines, which lack typical identifiable landmarks (Gunz and Mitteroecker, 2013). Semi-landmarks are allowed to slide along the curve to establish geometric correspondence (Gunz and Mitteroecker, 2013). After this alignment, the landmarks are treated as regular ones in subsequent analyses. We placed one fixed landmark at the base of the wing, fixing the overall landmark configuration with respect to this homologous position available for all specimens. All landmarks were digitised using TpsDig2 (Rohlf, 2015). Wing outlines were superimposed by performing a generalised Procrustes analysis (Rohlf and Slice, 1990) using the geomorph R package (Adams and Otárola-Castillo, 2013). This procedure isolates the shape information from the extraneous variations, namely: the size, position, and orientation. To visualize wing shape variation, we then performed a phylogenetic Principal Component Analysis (phyloPCA) on the superimposed coordinates using the phytools R package (Revell, 2012). Shape changes carried on the PCA axes were visualised using the plotRefToTarget function in geomorph. To assess the strength of morphological sexual dimorphism and whether it may interfere with interspecific differences, we tested the effect of species and sex on the morphological parameter using analyses of variance (ANOVAs).

Testing the influence of phylogeny on flight kinematics and morphological traits

To assess whether shared evolutionary history among the studied species influenced variation in morphology and wingbeat kinematics, we tested if closely related species showed more similar values in the measured parameters than distantly related ones. This was done by computing phylogenetic signals with the Blomberg’s K-statistics (Blomberg et al., 2003) on each morphological and flight parameters using the R package phytools (Revell, 2012). Furthermore, we accounted for the effect of phylogeny when assessing the scaling relationships between size and traits using Phylogenetic Generalized Least Squares (PGLS) regression (Symonds and Blomberg, 2014). For the PGLS regressions, we assumed a Brownian motion model of evolution, and the model parameters were estimated using maximum likelihood. Tests were performed on the mean values per species using the most recent phylogeny of hoverflies (Wong et al., 2023). The phylogeny was pruned to align with our sample of 28 species for the morphological analysis and further pruned to include only the eight species used in the flight analysis.

Modelling the aerodynamics of hoverfly flight

To estimate how changes in wing morphology and wingbeat kinematics affect the aerodynamic forces required for flight, we modelled the aerodynamic vertical force produced by flapping wings in hovering flight using a simplified quasi-steady approach as (Ellington, 1984b) where is the wingbeat-average upward-directed aerodynamic lift force produced by a beating wing. This aerodynamic force thus varies quadratically with the angular speed of the beating wing , and linearly with air density (ρ), the second-moment-of-area of the wing (S2), and the lift force coefficient .

The second-moment-of-area S2 is the only morphological parameter in this equation, thus capturing how wing morphology affects aerodynamic force production, from a first principle. But this second-moment-of-area parameter combines both wing size and shape characteristics (i.e., wing area and its distribution over the spanwise axis (Ellington, 1984c)). To disentangle the respective effect of wing shape and size on aerodynamic force production, the second-moment-of-area can be decomposed as , where R is the wingspan, is the mean wing chord, and is the normalised second-moment-of-area of the wing. Here, wingspan and average wing chord are size dependent parameters; the normalized second-moment-of-area describes wing shape as the distribution of area along the spanwise wing axis, independent of wing size.

The lift force coefficient depends on the angle-of-attack, and for fruit flies this dependency can be modelled as CF = sin(α) CFα, where CFα is the angle-of-attack-specific lift force coefficient (Dickinson and Muijres, 2016). For hoverflies, we here assume a similar scaling of lift with angle-of-attack. Both angular speed (ω) and the angle-of-attack (α) of the beating wing are wingbeat kinematics parameters that directly affect aerodynamic force production. A flying insect can vary the angular speed of its beating wings by changing both its wingbeat frequency f and wing stroke amplitude Aϕ, as angular speed scales linearly with the product of these (ω ∼ f Aϕ). The decomposed model for estimating the effect of wing morphology and kinematics of aerodynamic lift force production in flapping insect flight is thus We used this model to decompose the relative effect of variations in wingbeat kinematic and wing morphology between species on aerodynamic force production.

Physical scaling of kinematics and morphology parameters with size

In hovering flight, the aerodynamic vertical force F should equal the weight of the animal. Thus, to maintain weight support among different sizes of hoverflies, the aerodynamic force should scale linearly with body mass. This cannot be achieved through isometric scaling, i.e. the expected scaling if geometric similarity is preserved.

For isometric scaling of morphology with mass, all length parameters scale with mass as liso∼m1/3, resulting in an isometric scale factor under morphological similarity asim=1/3. As a result, the morphological parameters of interest scale with body mass as , and thus both S2 and F scale with body mass as F∼S2∼m4/3 (see supplementary text for detailed derivation). This value is larger than one, and so with an isometric reduction in size the aerodynamic force for weight support reduces more quickly than body mass. Thus, compared to large species of hoverflies, small species need relatively larger wings or adjust their wingbeat kinematics to maintain weight support.

For the wingbeat kinematics, we define kinematic similarity when the size-independent kinematics parameters of interest also do not scale with size, and should thus remain constant across body mass. Specifically, wingbeat frequency (f), angular stroke amplitude (Aϕ), angular speed (ω), and angle of attack (α) are all expected to scale under kinematic similarity as f∼m0, Aϕ∼m0, ω∼m0, and α∼m0, respectively. As hoverflies become smaller, deviations from this kinematic similarity might occur to adjust for the reduction in weight support due to isometric size reduction. We emphasize that kinematic similarity functions here as a pure hypothetical scenario used as a baseline prediction for comparing variations in flight kinematics between species and sizes of hoverflies. This approach facilitates the identification of deviations from scale-invariant wingbeat kinematics required for producing weight support during hovering flight (see the “Theoretical consideration” section for further details).

Using our aerodynamic force model (Eqns 1-2) we can estimate how the combination of wing morphology and wingbeat kinematics should scale with size, while maintaining weight support for hovering flight across sizes . Here, we do so for two extreme cases: (1) weight support is maintained across sizes using allometric scaling of wing morphology only, and thus wingbeat kinematics are kept constant (kinematic similarity); (2) weight support is maintained across sizes using allometric scaling of wingbeat kinematics, while wing morphology scales isometrically. These two scenarios are thus two alternative null hypotheses to be tested in this study:

To maintain weight support with non-varying wingbeat kinematics, the second-moment-of-area should scale linearly with body mass (S2∼m), resulting in negative allometry of S2 relative to body mass (S2∼m1<m4/3). This could be achieved by negative allometric scaling in the wingspan, mean chord or normalised second-moment-of-area , or a combination of these. To maintain weight support via negative allometric scaling of only one of these three metrics, while the others scale isometrically, then these should scale with mass as (see supplementary text for detailed derivations).

To maintain weight support via allometric scaling of kinematics, while wing morphology scales isometrically (S2∼m4/3), the angular speed of the beating wing should scale with body mass as ω∼m−1/6 (see supplementary text for detailed derivations). Again, a hovering hoverfly could achieve this using a comparative adjustment in wingbeat frequency or wing stroke amplitude (f∼m−1/6 or Aϕ∼m−1/6, respectively), or a combination of both (ω∼f Aϕ∼m−1/6). Furthermore, the animal could adjust the aerodynamic force vector using an equivalent adjustment in the wing angle-of-attack (sin(α)∼m−1/6).

To test these two alternative null hypotheses, we examined how wing morphology and wingbeat kinematics scale with body mass for our hovering hoverflies, and whether these scalings were isometric or allometric. We did so using Phylogenetic Generalized Least Square (PGLS) regressions of the morphology and wingbeat-average kinematics metrics in Eqns 1-2 against the body mass (m) (Symonds and Blomberg, 2014). Measurements were all log10-transformed prior to the regression analysis. For each morphology and kinematics metric, we tested if the observed scaling factor significantly deviates from geometric similarity or kinematic similarity, respectively , respectively), by verifying if its 95% confidence intervals excluded the similarity slope.

To better visualize variations in wing shape and size relative to body mass among species, we scaled all wings with the length-scale for geometric similarity (liso∼m1/3) and calculated the corresponding weight-normalised wingspan as R*=R/m1/3. We performed additional PGLS regressions to test if other kinematic adjustments were associated with changes in body mass, in addition to the ones considered in the aerodynamics model (Eqns 1-2). This includes regressions of body mass versus the amplitudes of the additional measured wingbeat kinematics angles (Aη, Aθ), and the corresponding peak angular rates .

The relative contribution of allometric scaling in wing morphology and kinematics to maintaining weight-support across sizes

We quantitatively tested our two alternative null hypotheses by assessing how the potential allometric changes in morphology and kinematics contribute to maintaining weight support across the studied range of hoverfly sizes. We did so by comparing, for each relevant morphological and kinematics parameter, its observed mass-specific allometric scaling factor (aallo) with both the scaling factor for morphologic or kinematic similarity (asim), and the scaling factor for achieving weight support fully by allometric scaling of only that parameter (aws). Hereby, we calculated the relative allometric scaling factor for weight-support as This parameter defines the percentage of which the observed allometric scaling of the specific morphological or kinematics parameter contributes to maintaining weight support during hovering flight, across the studied size range of hoverflies (see supplementary text for detailed derivations).

Computational fluid dynamics simulations

We performed Computational Fluid Dynamics (CFD) simulations to quantify aerodynamic force produced by the different flying hoverfly species. These simulations allowed us to explicitly quantify the effect of size, wingbeat kinematics and morphology on aerodynamic force production. Furthermore, we used the CFD results to test the validity of our simplified quasi-steady aerodynamic model (Eqns 1-2).

Each simulation was performed with a single rigid, flat wing with species-specific contours, and the averaged wingbeat kinematics across all species. The computational domain was 8R×8R×8R, and we simulated three consecutive wingbeat cycles. The wing has a thickness of 0.0417 R. We assumed hovering flight mode, so no mean inflow was imposed (U∞ = 0 m s−1).

We performed the simulations using our in-house open-source code WABBIT (Wavelet Adaptive Block-Based solver for Insects in Turbulence) (Engels et al., 2022, 2021). The computational approach is based on explicit finite differences that solve the incompressible Navier–Stokes equation, in an artificial compressibility formulation. Wavelets are used to dynamically adapt the discretisation grid to the actual flow at every given time t. This allows us to focus the computational effort where it is required to ensure a given precision, while the grid is significantly coarsened wherever possible. Starting from a coarse base grid, we allow up to seven refinement levels, each one refining the grid by a factor of two. In this work we use the Cohen-Daubechies-Feauveau wavelet CDF42. The grid is composed of blocks of identical size with Bs=22 points in all three spatial directions, yielding an effective maximum resolution of 352 points along the wingspan. An additional simulation with eight refinement levels confirmed validity of our results.

As demonstrated in numerous validation tests, obtained numerical data can be regarded as quasi-exact in the sense that they are equivalent to an experiment with identical geometry and kinematics. More details on the code can be found in Engels et al., (2021); the required parameter files to reproduces the simulations presented here can be found in the supplementary information. To accelerate the computation, each simulation used up to 600 CPU cores with 2.7 TB of memory in total.

From a fluid dynamics point of view, changing wing length R or wingbeat frequency f changes the Reynolds number Re (Batchelor, 1967). For each species, we performed a simulation using two different values for Re, one at the average Reynolds number for all species, and a second at the species-specific Reynolds number. This was initially done to separate the effect of Reynolds number and wing shape between species. The non-dimensional aerodynamic force was however very similar in both cases, hence the influence of Reynolds number is negligible. In general, the range of Reynolds numbers for the studied hoverfly species is rather narrow (Re ∈ [324,1310]). We finally assessed the contribution of interspecific variation in wingbeat frequency to force production by rescaling all aerodynamic forces with the square of the species-specific and average wingbeat frequency ratio .

The relative contribution of variations in wing size, shape and kinematics to aerodynamic force production

By combining the CFD simulation results with our aerodynamic force model (Eqns 1-2) and our physical scaling law analysis, we assessed the relative contribution of changes in wing size, shape and kinematics between the studied species on the aerodynamic force production required for weight support.

Because wingbeat kinematics were found to vary little across species and to be uncorrelated with body mass, we based this analysis on the average wingbeat kinematics of all species combined (see CFD methods above for details). We did assess the contribution of interspecific variation in wingbeat frequency on aerodynamic force production, as this varied most between species. We did so by rescaling all aerodynamic forces from CFD with the square of the species-specific and average wingbeat frequency ratio . According to the quasi-steady model, the vertical aerodynamic force from CFD should scale with the product of the wing’s second-moment-of-area and the wingbeat-mean angular wing speed squared (Eqn 1: ). The angular wing speed scales linearly with wingbeat frequency (ω∼f), and the second-moment-of-wing-area can be decomposed into the wing size and shape parameters wing span, mean chord and normalised second-moment-of-area as (Eqn 2).

In our CFD-based scaling analysis, we first tested the validity of our aerodynamic model (Eqns 1-2) by correlating the CFD-based vertical aerodynamic force with the product of second-moment-of-area and wingbeat-average angular speed squared (Eqn 1: ). For this, we used an Ordinary Least Squares (OLS) regression on the log-transformed data to quantify the scaling factor between these metrics. We used the regression statistics to test whether the CFD-based force scaled significantly with (with cut-off at P<0.05). Based on these results, we then quantified the corresponding relative allometric scaling factor a* (Eqn 3), to assess how allometric scaling of contribute to changes in aerodynamic force production at different sizes.

Secondly, we estimated the relative contribution of the primary variables in to aerodynamic force production, being second-moment-of-area and wingbeat frequency. We did so by using the same approach as described above, by first determining the scaling of CFD-based aerodynamic force with second-moment-of-area and wingbeat frequency, and secondly by estimating the corresponding relative scaling factors a*. Finally, we used this approach to estimate the relative contributions to aerodynamic force production of the S2-components: wingspan, mean chord length, and normalized second-moment-of-area.

This combined analysis allowed us to estimate the relative contribution of the various morphology and kinematics metrics to variations in vertical force production for weight-support across sizes, and test how hoverflies might have adapted specific wing morphology and kinematics parameters to maintain weight support across sizes. Moreover, comparing the calculated weight-support percentages in our model with 100% for full weight support, allowed us to test the validity of our analysis approach and aerodynamic force model.

Theoretical considerations

Our approach uses the concepts of morphological and kinematic similarity to define our working null hypotheses for two extreme scenarios in scaling of the flight propulsion system with size. Here, morphological similarity is defined as the case where all morphological parameters scale isometrically with size; kinematic similarity dictates that all wingbeat kinematics parameters that are not size related also do not scale with size (i.e., f∼m0, Aϕ∼m0, ω∼m0, and α∼m0). Based on these concepts, we defined two alternative null hypotheses: (1) weight support is maintained across sizes using allometric scaling of wing morphology only, and thus wingbeat kinematics are kept constant (kinematic similarity); (2) weight support is maintained across sizes using allometric scaling of wingbeat kinematics, while wing morphology scales isometrically (morphological similarity).

It is important to note that both these scenarios of morphological and kinematic similarity are pure theoretical concept that are physically, biomechanically nor physiological sustainable across sizes (Borelli, 1680; Hill, 1950). We here merely use these theoretical scenarios to define and test our null hypotheses. We do so by comparing the variations in morphology and kinematics for the differently sized hoverfly species with those based on our null hypotheses and based on morphological and kinematic similarity.

Results

In this study, we examined the morphology of 28 hoverfly species, with body masses ranging from 3 to 132 mg. Among them, we analyzed the hovering flight biomechanics and aerodynamics of eight species, with body masses ranging from 5 to 100 mg. The sampled species were well spread throughout the phylogeny of hoverflies (Figure 1). Among the eight species for which we studied flight biomechanics, body mass variation was unevenly distributed. Only one species, Eristalis tenax, weighed more than 50 mg, while the two closely related species Syrphus nigrilinearus and Eupeodes nielseni had highly similar body masses (Figure 1).

Sexual dimorphism in morphological traits was observed in body mass, wing chord, and wing second-moment-of-area (Table S1). However, the direction of this difference varied across species, with males exhibiting larger values in some cases and females in others. Interspecific difference in morphology was much stronger (Table S1), making sexual dimorphism negligible in comparison. While sexual differences in flight kinematics could not be assessed in this study (see Methods), they are likely to be similarly outweighed by interspecies variation.

When testing the phylogenetic signal on wing morphology and body size among the 28 studied species, we detected a significant effect of phylogeny on all traits except for the second-moment-of-area. In contrast, the effect of phylogeny on flight, tested on eight species, was non-significant for all flight traits (Table S2).

No correlation between wingbeats kinematics and body mass

We quantified a total of 33 wingbeats across eight hoverfly species, including at least three wingbeats per species (Figures 2 and 3). The wingbeat frequencies of the studied wingbeats were highly correlated with the average wingbeat frequency calculated on the entire flight sequence ; mean±standard deviation; n=33 flight sequences; r=0.98, P<0.001). Here, the number of wingbeats per flight sequence was 26±28. This suggests that the digitised wingbeats reliably reflect those of the full flight sequences. Furthermore, the mean advance ratio for the digitised wingbeats was below the generally accepted threshold of 0.1 , confirming that the hoverflies were operating approximately in a hovering flight mode (Ellington, 1984a).

Wingbeat kinematics during hovering flight of the eight studied hoverfly species.

(A-E) Temporal dynamics of the wingbeat kinematics throughout the wingbeat cycle of all digitised wingbeats. Separate wingbeats are colour coded by body mass (see legend on top), and the black lines show the average wingbeat kinematics for all wingbeats combined. (A-C) Temporal dynamics of the three conventional wingbeat kinematics angle (stroke, deviation, and rotation angle, respectively; see Figure 2 for definitions). (D-E) Angular speed and angle-of-attack of the wings throughout the wingbeat cycle, derived from wing stroke, deviation and rotation angles. (F-I) Derived wingbeat-average wingbeat kinematics parameters for each studied species versus the body mass of that species. The kinematics parameters are wing stroke amplitude Aϕ, wingbeat frequency f, wingbeat-average angular wing speed , and mean angle-of-attack at mid wing stroke , respectively. Each data point shows mean ± standard error per species, and is colour-coded according to the legend on the top. None of the wingbeat-average wingbeat kinematic parameters were significantly associated with body mass (Table 1). Horizontal dashed lines show the mean parameter values, as expected under kinematic similarity.

The eight hoverfly species exhibited similar wingbeat kinematics (Figures 2 and 3A-E). Here, they all moved their wings forward and backwards sinusoidally within the stroke plane (Figure 3A), with a wing stroke amplitude of Aϕ=100°±18° (Figure 3H). Movements out of the stroke-plane, as expressed by the deviation angle, are minimal (Figure 3B). Therefore, the oscillatory angular speed of the wingbeats (Figure 3D) is primarily caused by the wing stroke movement, and peak at approximately 5°×104 s−1 (Figure 3F). During the majority of each wingbeat (forward and backward wing stroke), the wing operates at approximately constant wing rotation angle and angle-of-attack (Figs 3C and 3E, respectively). At the end of the forward and backward wing stroke, the wings rapidly supinate and pronate, respectively. This causes the wing to flip upside down at stroke reversal, allowing it to operate again at the constant wing angle-of-attack of (Figure 3I). The hoverflies flew with a mean body pitch angle of βbody=30°±16°. The corresponding pitch angle of the wing stroke-plane was βstroke-plane=−15°±16°, showing that these hoverflies hover with an inclined stroke-plane.

The temporal dynamics of wingbeat kinematics was apparently not associated with variation in body mass (Figures 3A-E). Accordingly, none of the derived wingbeat-average wing kinematics parameters varied significantly with body mass, although a small negative trend was observed for the wingbeat frequency (Table 1, Figures 3F-I and S2). These include the parameters that directly affect aerodynamic force production (Eqn 1-2, wingbeat-average angular speed , angle-of-attack stroke amplitude Aϕ and wingbeat frequency f), and the additionally estimated wingbeat kinematic parameters (rotation amplitude Aθ, deviation amplitude Aη, and peak stroke rate rotation rate and deviation rate ) (Table S4).

Furthermore, variations in wingbeat kinematics were little affected by the flight kinematics metrics flight speed and climb angle (Table S3). Only flight speed had a positive significant effect on the angular speed of the beating wings, suggesting that faster flying hoverflies beat their wings at higher angular speeds. When controlling for this effect of flight speed on angular wing speed, the regression between angular wing speed and body mass remained non-significant (r2=0.02; P=0.41). Additionally, flight speed was not correlated with body mass (r2=0.01, P=0.43).

Altogether, these findings contradict our second null hypothesis, which proposed that hoverflies maintain weight support through allometric scaling of wingbeat kinematics (with wing morphology scaling isometrically), i.e., that the product of wingbeat frequency and amplitude should scale with mass under negative allometry (ω∼f Aϕ∼m−1/6). Variation in wingbeat kinematics between species thus cannot be explained by the demands of maintaining weight support throughout (isometric) size changes (Figure 3). Allometric adjustments for maintaining weight support across sizes among hoverfly species thus more likely occur through changes in wing morphology, as formulated in our first null hypothesis.

Wing morphology covaries with body mass

Wing morphology strongly covaried with body mass: the wing’s second-moment-of-area (S2), span (R), mean chord (), and normalized second-moment-of-area () all significantly correlated with body mass (Figure 4, see also Figure S3). Covariation between the wing morphological parameters and body mass all deviated from the isometric expectation, except for mean chord (Table 1, Figure 4). Second-moment-of-area, wingspan and all showed a significant negative allometry, i.e., they were all disproportionately larger in smaller species (Figure 4A-C, Table 1), supporting our first null hypothesis that weight support is maintained across sizes using allometric scaling of wing morphology.

The scaling of wing morphology with body mass for all 28 studied hoverfly species (A-D), and how allometric scaling of wing morphology contributes to maintaining weight support across the hoverfly size range (E,F).

(A-D) body mass (abscissa) versus on the ordinate the second-moment-of-area S2, wingspan R, mean chord , and normalized second-moment-of-area , respectively. Each data point shows a species average value. Data points of species for which flight was studied are colour-coded (see top row), and species with only quantified morphology are shown in grey (see the corresponding individual-specific data in Figure S3). The black, red and grey trendlines show the best fitting line from a PGLS regression, isometric scaling, and the expected scaling for metric-specific maintenance of weight support, respectively (see legend above A). All fitted PGLS regression slopes, except for wing chord, are significantly lower than expected under geometric similarity, suggesting negative allometric scaling of wing size and shape with respect to body mass (see also Table 1). (E,F) The relative contributions of different wing morphology metrics to maintaining weight-support across the studied range of hoverfly sizes, expressed by the relative allometric scaling factor a*. (E) the relative allometric scaling factor for the second-moment-of-area S2, and (F) the relative allometric scaling factor for the separate morphological components of S2: wingspan R, mean chord , and normalized second-moment-of-area .

Scaling of the second-moment-of-area with size differed the most from isometric scaling for morphological similarity (aallo=1.01 versus asim=1.33) and almost perfectly aligned with the scaling for metric-specific weight-support (aws=1). The corresponding relative scaling factor confirms this (Eqn 3), as it is close to 100% (a*S2=103%; Figure 4E). This suggests that hoverfly species of different size should be able to maintain weight-support during hovering flight almost purely through variations in the second-moment-of-area. By estimating the relative scaling factors of the various morphological parameters that compose the second-moment-of-area (a*R=71%; a*c=12%; a*S2*=10%, Figure 4F), we find that adaptations in wingspan contribute the most to the variations in second-moment-of-area. Allometric scaling in the normalized second-moment-of-area were smaller yet significant, indicating a notable contribution of change in wing shape to the adaptations in second-moment-of-area across sizes.

Our phylogenetic geometric morphometrics analysis revealed two main axes of wing shape variation among the studied species associated with the principal components PC1 and PC2, explaining 69.21% and 14.64% of the variation, respectively (Figure S4 and S5). Shape variation carried on PC1 was positively correlated with the (r=−0.49; P=0.009) and negatively with body mass (r=0.55; P=0.002) (Figure 5A and 5B, respectively), depicting the changes in wing shape associated with body mass variation among hoverflies species. This change in reflects a wing surface area mostly located distally in smaller species, whereas most wing surface area is located near the wing base in larger hoverflies species (Figure 5).

Changes in wing shape and size associated with body mass variation.

Each data point shows a species-average value (see the individual-specific data in Figure S5), and are colour-coded for species of which both flight and morphology were studied (see right column of B), and in grey for species with only morphology quantified. (A) Normalised second-moment-of-area (S2*) versus the primary Principal Component (PC1) of the phylogenetic Principal Component Analysis (phyloPCA) performed on the mean wing shape coordinates of all 28 studied hoverfly species. On the right, we show the 28 wing outlines colour-coded and normalised with body mass (with coordinates X∗=X/m1/3). (B) body mass (m) versus PC1 (left), and mass-normalised wing outlines versus body mass for the eight species used in our flight experiments (right). The left and right gritted wing shapes on the PC1 axis show the theoretical wing shapes at maximum and minimum value of PC1. PC1 explains 65.24% of the variations in wing shape (see also Figures S4 and S5). (A-B) The combined results show that in larger species, wing surface area was located more proximally (lower PC1 and S*2 values) than in smaller species, in which wing area tend to be located more distally (higher PC1 and S*2 values). Combined with the changes in wing shape (left), weight-normalised wingspan (R∗) and mean chord () tend to be larger in smaller species (right).

The combined correlative and morphometrics analyses thus shows that as their size decreases, hoverflies exhibit relatively larger wings that are also shaped such that their normalised second-moment-of-area is increased (Figures 4 and 5). These combined morphological adaptations allow smaller hoverflies to maintain in-hovering weight-support without the need to adjust their wingbeat kinematics.

Aerodynamic modelling confirms the main role of wing morphology for weight support

Because among the eight hoverfly species, wingbeat kinematics did not vary significantly with body mass (Table 1), we performed all CFD simulations with the average wingbeat kinematics of all species combined (Figure 6A). We did vary the wing shape and size between species (top row of Figure 6), and we modelled aerodynamic force production with both the species-specific wingbeat frequency and mean wingbeat frequency of all species combined (Figure 6B).

Aerodynamic forces produced by hovering hoverflies, as estimated using Computational Fluid Dynamic (CFD) simulations.

(A-B) For our simulations, we used the species-specific wing shapes and sizes (top row), but average wingbeat kinematics and frequency () across all eight studied hoverfly species (A,B). (C) The resulting temporal dynamic of vertical forces throughout the wingbeat cycle, coloured by species (see legend on top). (D) The wingbeat-average vertical force versus body mass, for all simulated hoverfly species operating at the average wingbeat frequency for all species (square data points and dashed trend line), and forces scaled to the species-specific wingbeat frequency f (round data points in B and D, and solid trend line). The trendline for weight support (F=mg) is shown with a grey dashed line. With the species-specific wingbeat frequency f (solid trend line and circles in B and D) hoverflies are closer to producing weight support than for the simulations at the average wingbeat frequency (dashed lines in B and D).

The resulting aerodynamic forces differed between species mostly in magnitude, but the temporal dynamics of force production were strikingly similar (Figure 6C and S6). For all species, the temporal dynamics of the vertical aerodynamic force showed a large force peak preceded by a small peak, during each wing stroke (forward and backward stroke). The back stroke produces a larger vertical force and a more distinct higher harmonic. These dynamics are similar between species, despite their different wing shapes. This suggests that wing shape does not strongly affect the characteristics of aerodynamic force production.

The wingbeat-average vertical aerodynamic forces showed a strong scaling with body mass (Figure 6D), as expected due to the differences in wing size between species (top row in Figure 6). For each studied species, we modelled aerodynamic forces, both at the species-specific wingbeat frequency and at the mean wingbeat frequency of all species (Figure 6B,D). For the simulations with species-specific wingbeat frequency, the hoverflies produced on average weight-support, whereas for the mean wingbeat frequency cases weight-support was not achieved (Figure 6D). Variation in wingbeat frequency thus appeared necessary for individual hoverflies to achieve weight support, although between species wingbeat frequencies do not correlate with body mass.

We continued to analyse the relative effect of variations in wing morphology and wingbeat-frequency on the aerodynamic force production, as quantified from CFD (Figure 7, Table 2). Here, all the tested morphology and kinematics parameters varied significantly with aerodynamic force magnitude, except for wingbeat-frequency and normalized second-moment-of-area.

The relative contribution of allometric variations in wing morphology and wingbeat kinematics to CFD-derived vertical aerodynamic force production, and how this contributes to maintaining in-hovering weight-support across the sampled range of hoverfly sizes.

(A-C, E-G) Wingbeat-average vertical force at species-specific wingbeat frequency f (abscissa) vs. on the ordinate various wing morphology and kinematics traits. In each panel, data points are results for different species, colour-coded according to the legend on the top. The solid and dashed black lines show the significant and non-significant OLS regression fits, respectively. The dashed red and grey lines show the expected slopes for scaling under morphological and kinematic similarity and for 100% metric-specific weight-support, respectively. (D,H) The relative contributions of allometric variations in wing morphology and kinematics to maintaining weight-support across the studied range of hoverfly sizes, expressed by the relative allometric scaling factor a* (Eqn 3). Contributions based on significant and non-significant OLS regressions are shown in dark and light grey, respectively. (A) The wingbeat-average vertical force scales linearly with the product of second-moment-of-area and wingbeat-average angular speed squared (Eqn 1: ), resulting in weight-support across all sizes. (B-D) Separating this product into its main components (B and C, respectively) shows that the second-moment-of-area and wingbeat frequency contribute 81% and 22% to maintaining weight support across sizes, respectively (D). (E-H) Parting the contribution of second-moment-of-area into its components (E-G) shows that allometric variations in wingspan, mean chord and normalized second-moment-of-area contribute 55%, 19% and 9% to maintaining weight support across sizes, respectively (H).

First, we correlated the vertical forces computed using CFD with the product of second-moment-of-area and wingbeat-average angular speed squared (Figure 7A). According to our quasi-steady aerodynamic model, this should yield a linear correlation (Eqn 1: ), which was also the case as the allometric scaling factor of aall=0.99 [0.97 – 1.01] (n=8) is not significantly different from linear (Table 2). The corresponding allometric scaling factor equals 103%, suggesting that the allometric adaptations that drive variations in fully explain variations in aerodynamic force production required for maintaining weight support across sizes. It also shows that our quasi-steady aerodynamic model predicts the effect of morphology and kinematics on aerodynamic force production well.

Separating this dependency of aerodynamic forces on into its primary components, second-moment-of-area and wingbeat frequency (Figure 7B-D), shows that allometric adaptations in second-moment-of-area and wingbeat frequency contribute 81% and 22% to maintaining weight support across sizes, respectively. Hereby, the aerodynamic forces do not scale significantly with wingbeat frequency (Figure 7B; Table 2), as was also the case for the equivalent correlation of wingbeat frequency with body mass (Figure 3F; Table 1). These combined results confirm that adaptations in morphology drive variations in aerodynamic force production between hoverflies species of various sizes, although variations in wingbeat frequency cannot be ignored.

Further parting the effect of S2 on aerodynamic force production into its components (Figure 7E-G) shows that allometric adaptations in wingspan R, mean chord and normalized second-moment-of-area contribute 55%, 19% and 9% to variations in aerodynamic force production across sizes, respectively (Figure 7H). Here, normalized second-moment-of-area did not significantly scale with vertical force production (Figure 7G; Table 2), as opposed to the significant scaling found in the corresponding analysis with body mass for the extended 28 species dataset (Figure 4D; Table 1).

The sum of the relative scaling factors of all metrics contributing to vertical force production (Eqn 2) equals 105% (Figure 7H), suggesting that these allometric adaptations combined fully explain variations in aerodynamic force production required for maintaining weight support across sizes. The here-identified effects of morphological traits on CFD-based forces are strikingly similar to those from the comparative analysis between morphology and body mass (Figure 4). This shows that our kinematics analysis based on eight species captures the major effects of wing morphology on aerodynamic force production, as expected based on the morphological analysis with the large sample size of 28 species.

Discussion

Allometric changes in wing morphology enables small hoverflies to maintain weight support during hovering

In this study, we investigated whether variation in body mass among hoverfly species is associated with changes in wing morphology (null hypothesis one) or in wingbeat kinematics (null hypothesis two). Both components directly influence the aerodynamic forces produced by beating wings during flapping flight (Ellington, 1984b), and therefore could accommodate for maintaining weight support during evolutionary changes in size occurring throughout the diversification of hoverfly species. Comparing how wingbeat kinematics and wing morphology scale with body mass across hoverfly species, we showed that wing size and shape rather than wingbeat kinematics scale with body mass. This supports the first null hypothesis while rejecting the second, suggests that wing morphology might respond stronger than wingbeat kinematics to the constraints imposed for maintaining weight support during flight in hoverflies.

Changes in wing morphology associated with body mass variation are commonly observed in flying animals (Le Roy et al., 2019; Norberg, 2007; Rayner, 1988; Taylor and Thomas, 2014; Tercel et al., 2018), indicating that wing and body morphology evolve together as an integrated phenotype. Increasing size is usually not only associated with relatively larger wings, but also accompanied by changes in wing shape (Danforth, 1989; Outomuro et al., 2013). Teasing apart the changes in wing morphology resulting from constraints imposed by body mass from that of other selective pressures is however not straightforward. It is generally facilitated by comparing species with similar ecology, because this limits the effect of contrasted ecological pressures at play in different species (García and Sarmiento, 2012). Hoverfly species present a similar ecology in that all adults are specialised flower visitors (Klecka et al., 2018). Comparable selective regimes are thus likely acting on the evolution of flight performance and associated flight kinematics and morphology in adult hoverflies.

Shared ancestry can also affect the evolution of wing morphology and its effect should thus be controlled to understand how body mass may constrain the evolution of other traits (Blomberg et al., 2003). Among the 28 species studied here, variation in the measured morphological traits was significantly influenced by phylogenetic history. In contrast, among the sub-set of eight species for which we studied flight kinematics, phylogenetic signal in flight traits was non-significant. This may suggest either a higher evolutionary lability in flight compared to morphological traits, or results from limited statistical power due to the small number of species with quantified flight data. Note that the lack of phylogenetic signal does not necessarily preclude an impact of phylogeny on the covariation between traits and body mass. As such, changes in wing morphology (and possibly in wingbeat kinematics) resulting from constraints imposed by body mass are confounded with the effect of shared ancestry. Our phylogenetically informed regressions of morphological and flight traits against body mass nonetheless allow disentangling such effects.

The large size variation among hoverfly species results from 90 million years of evolutionary diversification (Wong et al., 2023). While some taxa may have evolved larger body sizes than their ancestors, others may have undergone size reduction. Determining the exact trajectory of body size evolution in the species studied is beyond the scope of our study. Addressing the aerodynamic challenges of size reduction may not fully represent the evolutionary constraints acting on all hoverfly species. However, we chose to focus on the adaptation of the wing-based propulsion system in this context, as it provides a valuable framework for understanding the evolution of flight in small flying insects.

As size decreases, the wings become relatively less effective in producing the aerodynamic forces required for weight support (see Supplementary Text). The constraints imposed by weight support during hovering flight should then promote the evolution of disproportionately large wings, or an increased wingbeat frequency or amplitude in smaller hoverflies species. We found that none of wingbeat kinematic parameters significantly scaled with body mass but that wings indeed tended to be disproportionately larger in smaller species (i.e. negative allometry). Here, size-dependent adaptations in the primarily wing morphology parameter second-moment-of-areas (S2; Eqn 1) almost fully explained the required adaptations for maintaining weight-support across sizes (a*S2=98%; Figure 4E). Parsing the second-moment-of-areas in its separate wing size and shape components shows that both the primary wing size wing shape metrics (wingspan R and normalized second-moment-of-areas , respectively) exhibit significant negative allometric scaling (Figure 4B, D). These allometric adaptations in wing size and shape are predicted to contribute to maintaining weight-support across sizes for 71% and 10%, respectively (Figure 4F). Thus, changes in multiple aspects of the wing morphology may contribute to mitigating the negative effect of size reduction on flight ability, whereby wing shape effects are small but significant (Figure 5).

A negative allometry between wing size and body size has been found in other insect groups: for solitary bees this was found at both the interspecific and intraspecific level (Grula et al., 2021), and at the intraspecific level in rhinoceros beetles (Kawano, 1995). In contrast, disproportionately larger wing were found in larger species of dragonfly, suggesting positive allometry in these larger insects (May, 1981). This opposite allometric scaling of wing size in dragonflies and hoverflies may be the result of the use of contrasted mechanisms for producing aerodynamic forces, involving both morphological or kinematics variations, and contrasted flight modes (e.g., gliding flight in dragonflies).

The observed negative allometry in wing shape with mass was further emphasised in our geometric morphometrics analysis (Figures 5, S4 and S5). Indeed, in our phylogenetic Principal Component Analysis performed on superimposed wing shape coordinates, the primary axis PC1 explained a striking 69% of the variations in wing shape, and was strongly correlated with normalised second-moment-of-area . This PC1 shape variation depicts a shift in the spanwise distribution of wing area, from more proximally located in larger species to more distally located in smaller species (Figure 5). Interestingly, similar changes in wing shape associated with decreasing body size have been found among species of Hymenoptera (Danforth, 1989). An illustration of the most pronounced end of this pattern is the highly spatulated wing shape found in tiny wasp species (e.g. in the Mymaridae or Mymarommatidae families). This shows that although the relative contribution of allometric scaling of wing shape to maintaining weight support is relatively small (10%; Figure 4F), physical constraints associated with size reduction are strong enough to drive these adaptations. Evolutionary changes in the body size of hoverflies may thus have favoured wing shapes in smaller species that produce aerodynamic forces more effectively.

We continued to study how allometric variations in wing morphology and wingbeat kinematics affected aerodynamic force production across the studied range of hoverfly sizes (Figures 6 and 7). For this, we combined flight experiments on eight hoverfly species with both analytical quasi-steady aerodynamic modelling (Eqns 1,2) and numerical aerodynamics modelling (CFD). By combining these results, we first tested the validity of our quasi-steady aerodynamics model by correlating the wingbeat-average vertical forces estimated from CFD with the product of second-moment-of-area and wingbeat-average angular wing speed (; Eqn 1; Figure 7A). The resulting linear correlation shows that our quasi-steady aerodynamic model effectively captures the effect of wing morphology and kinematics on aerodynamic force production. By separating the combined effect of variations in morphology and kinematics on aerodynamic force production into its components (Eqn 2), we confirm that adaptations in morphology drive the variations in aerodynamic force production between hoverfly species of different sizes. Here, wing size variations as expressed by wingspan have the largest effect on aerodynamic force production, but variations in wing shape and wingbeat frequency cannot be completely ignored (Figure 7C,G,H). In fact, the respective 9% and 22% contribution of wing shape and wingbeat frequency to variations in aerodynamic force production, need to be included in our model to maintain 100% weight support across the size range of hovering hoverflies (Figure 7H).

Factors decoupling wingbeat kinematics from the influence of body mass

Our study revealed that wingbeat kinematics is relatively stable across hoverfly species, with none of the measured kinematics parameters significantly correlated with body mass (Figure 3; Table 1). Performing computational fluid dynamic simulations using the average wingbeat kinematics for all species, but incorporating differences in wing morphology, demonstrated that weight support was indeed achieved across species. Important however is that variation in wingbeat frequency – albeit uncorrelated with body mass – has proven crucial for correctly estimating the aerodynamic force necessary to achieve weight support (Figures 6D and 7C,D). Adjustments in wingbeat frequency thus appear to play a role in achieving weight support during hovering for individual hoverflies. The hovering hoverflies might thus primarily use wingbeat frequency to finely tune the vertical aerodynamic force required for weight support. Unlike birds and bats, flying insects cannot actively morph their wings to control aerodynamic force production (Harvey et al., 2022). As a result, flying insects rely purely on wingbeat kinematics changes to trim aerodynamic forces and torques during steady flight (Muijres et al., 2017), and to adjust these forces and torques during manoeuvring flight (Hedrick et al., 2009; Muijres et al., 2014). Our CFD simulation results suggest that hovering hoverflies use relatively simple wingbeat frequency modulations to compensate for the continuous daily fluctuations in body mass, as a result of food and water intake and excretion.

The lack of covariation between in-flight wing movement and body mass goes against the prediction that wingbeat kinematics is more likely to be adjusted with varying body mass due to the greater flexibility of flight kinematics as compared to morphology. This finding is also surprising given that size variation is generally accompanied with changes in wingbeat kinematics in other flying animals (Norberg, 2007; Rayner, 1988; Riskin et al., 2010; Taylor and Thomas, 2014; Tercel et al., 2018). A previous study dissecting the flapping wing kinematics of miniaturised insects found an increased wing deviation angle in smaller species, enabling higher wing velocity and hence aerodynamic force production (Lyu et al., 2019). Such a change in wing kinematics associated with decreasing size can be explained by the increased viscous drag constraint that tiny flying insects experience at low Reynolds numbers, requiring higher aerodynamic force generation. Here we did not find an increased wing deviation angle in smaller hoverfly species, most-likely because our study focusses on a too narrow range of sizes and consequently too similar Reynolds numbers. Our CFD analysis confirms this, as the effect of Reynold number aerodynamic force production was negligible.

While only a handful studies have examined the effect of body mass on detailed wingbeat kinematics in insects (such as wingbeat amplitudes and angle-of-attack Belyaev et al., 2014; Lyu et al., 2019; Meresman and Ribak, 2017), the scaling between body mass and wingbeat frequency has been frequently assessed (Bartholomew and Casey, 1978; Byrne et al., 1988; Casey et al., 1985; De Nadai et al., 2021; Dudley, 1990; Tercel et al., 2018). In theory, an increase in wingbeat frequency is expected in smaller insects because flapping wing-based aerodynamic force production decreases faster than body mass. As such, it is puzzling that the reduction in size among the studied hoverfly species is not associated with increased wingbeat frequency.

Interestingly, inconsistent results on the scaling between body mass and wingbeat frequency have been found among other groups of flying animals. Although predominantly negative (e.g. in butterflies (Dudley, 1990), in orchid bees (Darveau et al., 2005), in mosquitoes (De Nadai et al., 2021), and in a comparative study across insect orders (Tercel et al., 2018)), several studies found weak or no relationship between body mass and wingbeat frequency (e.g. in moths (Bartholomew and Casey, 1978), whiteflies (Byrne et al., 1988), bees (Duell et al., 2022; Grula et al., 2021)). Consistent with our findings, a study comparing the wingbeat kinematics during tethered flight among hoverfly species spanning a range of body mass similar to that investigated here found no significant effect of body mass on wingbeat kinematics (Belyaev et al., 2014).

Interestingly, multiple studies have highlighted that deviation from the expected scaling of wingbeat frequency with body mass is often associated with allometry in wing area (see literature review, Darveau, 2024). Flying animals exhibiting negative allometry in wing area tend to have lower wingbeat frequencies than would be predicted based on their body mass alone (Bartholomew and Casey, 1978; Dorsett, 1962; Norberg, 2012). In line with this pattern, the lack of increased wingbeat frequency in small hoverflies is here associated with disproportionately large wings.

Beyond morphological constraints, ecological factors may also contribute to explaining the heterogeneous scaling trends found among insect groups. The extent to which body mass constrains the evolution of wingbeat kinematics may depend on species ecology. Adaptation to a particular ecological niche may drive the evolution of specialised flight abilities, promoting a specific wingbeat kinematics. Selective pressures exerted on these wingbeat kinematics may sometimes overcome the constraints imposed by weight support. In that case, relatively constant wingbeat kinematics would be maintained over a range of body mass, and changes in wing morphology alone are expected to adjust for weight support. Consistent with this hypothesis is the decoupling of wingbeat kinematics from body mass variation, which has been highlighted in species where hovering flight ability is crucial for food acquisition. These include bees (Duell et al., 2022; Grula et al., 2021), sphingid moth (Bartholomew and Casey, 1978), hummingbirds (Skandalis et al., 2017), and hoverflies (Belyaev et al., 2014) as also in the present study. The absence of correlation between body mass and wingbeat kinematics observed in hoverflies may thus be due to their highly specialised hovering flight abilities.

The selective pressures promoting the evolution of unique hovering flight abilities in hoverflies are not precisely known but hovering steadily is likely advantageous for navigating between flowers when foraging for nectar. Good hovering flight performance may also have been promoted because it increases mating opportunities during the leck gathering of males (Downes, 1969; Gilbert, 1984; Heinrich and Pantle, 1975). Thus, selection for increased foraging efficacy and/or mating opportunities may have maintained consistent wingbeat kinematics across hoverfly species despite variation in body mass.

To further ascertain our findings, quantifying sexual dimorphism in hoverfly wingbeat kinematics would be a relevant future research direction. This would clarify the role of mating behaviour on the evolution of hovering flight performance. In this study, sexes could not be distinguished during our flight experiments, preventing us to test whether male hoverflies exhibit better hovering performance. Furthermore, additional wingbeat kinematics measurements in species from different habitats would help verifying the generality of our results in assessing the possible effect of habitat. Contrasted environmental conditions (e.g., temperature, humidity, or altitude) can affect flight performance (Altshuler and Dudley, 2003; Farnworth, 1972), and thus interfere with the scaling between flight and morphology.

Here, we focussed on how variations in wing morphology and kinematics allow hoverflies of different sizes to produce weight-support in hovering flight. Hovering flight is a functionally and energetically demanding type of locomotion (Ellington, 1991), but other flight behaviours might be even more so (Alexander and Vogel, 2004; Muijres et al., 2014). Future research should also focus on how the variation in scaling highlighted here influences such peak locomotor performance, especially because the effect of morphology on performance is generally most pronounced when individuals are pushed to their limits (Losos et al., 2002). For instance, investigating how the differential scaling of wing morphology and kinematics with size affects peak thrust production during evasive flight manoeuvres would be a relevant experiment for future study.

Finally, a critical but overlooked aspect of our study is the interplay between wingbeat kinematics and muscle physiology. Although our analysis focuses on the wing-based propulsion system, wing motion is ultimately powered by the muscular motor. The scaling of muscle mechanical output is linked to wingbeat kinematics, yet the findings discussed above are largely decoupled from the muscle constraints that may accompany changes in body size. Future studies that integrate muscle physiology with aerodynamic modelling are thus needed to complement the current approach.

Conclusion

Altogether, our results suggest that the hovering flight abilities of hoverflies may stem from highly specialised wingbeat kinematics that have been largely conserved throughout their diversification. In contrast, changes in wing morphology have evolved with the aerodynamic constraints associated with evolutionary changes in size. The existence of selective pressures maintaining stable wingbeat kinematics despite variation in body mass implies that high hovering flight performance can only be achieved over a limited range of wingbeat kinematic parameters, which remains to be ascertained.

Our study moreover highlights that unlike many other Diptera, hoverflies make use of a highly stereotyped inclined stroke-plane wingbeat kinematics for hovering, where during the forward dorsal-ventral wing stroke the wing moves not only from back to front but also downwards (Mou et al., 2011). This wingbeat kinematics pattern may contribute to the specialised hovering abilities of hoverflies, as it is consistently observed among the eight species studied here, with their stroke-plane pitch angle during hovering being approximately orientated at −30°.

Furthermore, despite significant changes in wing morphology and wingbeat frequency among the studied species, the temporal dynamics of aerodynamic force production is strikingly similar between them (Figures 6C and S6). This suggests that not only the wingbeat kinematics is conserved between species, but also the underlying aerodynamics including the relative use of different unsteady aerodynamic mechanisms (Sane, 2003). Thus, the apparent highly specialised flight style of hoverflies is maintained among several species with a large range of body masses, and robust against variations in wing shape and wingbeat frequency.